---
title: "Ukraine Humanitarian Pooled Fund Analysis"
subtitle: "Allocation patterns, partner concentration, grant scale and the health sector"
author:
  - name: "Jesus Baena"
date: "2026-02-13"
date-format: "MMMM D, YYYY"
format:
  html:
    theme:
      light: yeti
    css: custom-style.css
    toc: true
    toc-depth: 3
    toc-location: left
    toc-title: "On this page"
    code-fold: true
    code-tools: false
    code-summary: "Show code"
    code-link: true
    code-overflow: wrap
    embed-resources: true
    smooth-scroll: true
    number-sections: true
    anchor-sections: true
    fig-cap-location: bottom
    tbl-cap-location: top
    link-external-icon: true
    link-external-newwindow: true
    callout-appearance: simple
    callout-icon: true
    title-block-banner: "#003d4c"
    title-block-banner-color: "#ffffff"
    title-block-categories: true
    grid:
      sidebar-width: 280px
      body-width: 1120px
      margin-width: 40px
      gutter-width: 1.5rem
editor: visual
execute:
  warning: false
  message: false
  echo: true
  fig-align: center
  fig-width: 8
  fig-height: 5
  fig-dpi: 150
knitr:
  opts_chunk:
    dev: "png"
    dev.args:
      bg: "transparent"
    fig.showtext: true
---

# Introduction

This report analyses the Ukraine Humanitarian Pooled Fund (UHF) from its first allocation in **2019 through 2025**, and situates it within the broader CBPF system using comparative data from **28 country-based pooled funds over 2014–2025** (the mature operational period of the modality).

Rather than following the linear logic of the project cycle (number of beneficiaries, project duration, funding executed, etc.), this report steps back to examine overall funding flows and grant characteristics, with particular attention to the health sector.

The analysis is structured around three key questions:

What is the profile of the organizations receiving UHF funding? How does it compare with the rest of the CBPF system? What are the implications for the health sector?

## Key Findings

-   **Heavy reliance on the reserve allocation**: Since Ukraine became an active CBPF in 2019, Reserve allocations have accounted for **51.9 % of its budget (2019–2025)** — well above the non-Ukraine CBPF average of **40.2 %** — placing Ukraine among the top tier of funds (7th of 28) in reserve dependence, alongside Sudan, oPt and DRC. The balance is correcting in 2025.
-   **Larger grants, not fewer partners**: Ukraine grants are **2.6× larger at the median** than in the rest of the CBPF system (\$1.13M vs \$440K). This concentration of *value* — rather than a narrower partner base — is a defining feature of the UHF portfolio.
-   **UHF is not more concentrated than peer CBPFs — it's larger**: Across 2022–2024 Ukraine sits mid-pack among broad-partnership CBPFs on HHI (0.029), top-5 share (28%) and top-10 share (46%). What sets UHF apart is envelope size: \$541M spread across 94 partners produces a mean partner take of **\$5.75M over three years** — roughly 2× Yemen or Ethiopia and 4× Somalia. Same shape × bigger envelope = bigger grants.
-   **A stable core-20 captures 55% of the envelope**: Twenty organisations appeared in every UHF allocation round 2022–2024 and absorbed **\$299M of the \$541M envelope**. Seven of them (ACTED, UNHCR, DRC, PIN, Caritas Ukraine, NRC, Proliska) hold mean grants of \$3.4M–\$4.8M — at or near the \$5M cap — with individual exceptional-cap grants up to \$9.4M.
-   **Ceiling-bound distribution**: UHF's P90 grant size is exactly \$5.00M (the standard cap) — unique among CBPFs. Peer funds show a long tail; UHF shows a flat top.
-   **Visible capacity-volume mismatches**: Where 2021 revenue is public, some organizations has managed UHF volumes **4.3× its pre-war annual revenue** (\$9.4M against €1.9M). Seven of the eight core Ukrainian NNGOs have no publicly available 2021 revenue — an independent-assessment gap rather than a finding.
-   **Structural UN exit**: UN-Agency funding collapsed from \$66M (2022) to \$41M (2023) to **\$5M (2024)**; the 2024 Annual Report makes this explicit (*"UN would no longer be funded unless absolutely necessary"*). Redistributed volume flowed primarily to DRC, ICF Caritas Ukraine and a broadened NNGO tier.
-   **UHF-in-context**: AR-reported, Ukraine = **20% of global CBPF allocations (2024)** and **\>10% of the 2024 HNRP** — the largest CBPF for the third consecutive year.
-   **Data Sources**: Official UHF allocation and project data (CBPF Feb 2026 extract); UHF Annual Reports 2022, 2023, 2024.

# Data Overview

Data is retrieved from the official repository [data explorer](https://cbpf.data.unocha.org/dataexplorer.html), it was downloaded on February 2026 and it is formed by raw .csv files. The raw data spans from 2010 onward, but **analysis focuses on 2014-2025** to ensure comparability (2010-2013 was the pilot phase with limited geographic coverage). Data was cut at the end of 2025 to use static data for this analysis. As such, the recently announced top-up allocation for Ukraine from the US State Department in February 2026 is not included in this report. A posteriour update could be considered.

In total the data comprises:

-   **28 country-based pooled funds** (one per `PooledFundName`, including two dedicated Regional Humanitarian Pooled Funds — Chad RhPF and Mozambique RhPF — and "Syria Cross-border" alongside Syria)

-   **16,304 approved projects** across all years; **15,657** within the 2014–2025 analysis window

-   **891 distinct allocation rounds** (`AllocationType`), e.g. "2022 3rd Reserve Allocation", "2024 1st Standard Allocation"

```{r setup}
#| include: false

# Load required libraries
library(dplyr)
library(ggplot2)
library(tidyr)
library(readr)
library(lubridate)
library(stringr)
library(knitr)
library(scales)
library(patchwork)
library(ggtext)
library(gghighlight)
library(ggrepel)

# Data paths are relative to the .qmd file; no setwd() needed.
# Quarto resolves Raw-data/ from the report's own directory.

# --- Consistent ggplot2 Theme ---
theme_set(
  theme_minimal(base_size = 14, base_family = "") +
  theme(
    plot.title = element_text(face = "bold", color = "#005f73", hjust = 0.5, size = 16),
    plot.subtitle = element_text(color = "#343a40", hjust = 0.5, margin = margin(b=10)),
    plot.caption = element_text(color = "grey50", face = "italic", size = 9),
    axis.title = element_text(face = "bold", size = 12),
    axis.text = element_text(color = "grey30", size = 10),
    legend.position = "top",
    legend.title = element_text(face = "bold", size = 11),
    panel.grid.minor = element_blank()
  )
)

# Color palette matching the theme
theme_colors <- list(
  primary = "#005f73",
  secondary = "#2B638B",
  accent = "#c7254e",
  light_bg = "#eef7f7"
)
```

```{r load-data}

# Load all data files - Latest download from February 19, 2026
# Using complete global dataset (ALL versions) for all 28 CBPF countries
project_summary <- read_csv("Raw-data/ProjectSummary_ALL_20260219_100323_UTC.csv", 
                            show_col_types = FALSE)
project_summary_location <- read_csv("Raw-data/ProjectSummaryWithLocation_ALL_20260219_100323_UTC.csv", 
                                     show_col_types = FALSE)
pipeline_summary <- read_csv("Raw-data/PipelineProjectSummary_ALL_20260219_125012_UTC.csv", 
                             show_col_types = FALSE)
pipeline_cluster <- read_csv("Raw-data/PipelineProjectCluster_ALL_20260219_100323_UTC.csv", 
                             show_col_types = FALSE)
clusters <- read_csv("Raw-data/Cluster_ALL_20260219_100323_UTC.csv", 
                     show_col_types = FALSE)
contributions <- read_csv("Raw-data/Contribution_ALL_20260219_100323_UTC.csv", 
                          show_col_types = FALSE)
```

## Data Preparation

The raw data is well-structured and consistent with expected values. Minor transformations were applied to standardise date formats across datasets and to filter out 2026 entries, keeping the analysis within the 2014–2025 window. Helper functions `safe_date_format` and `safe_year_format` handle edge cases in date parsing robustly.

::: {.callout-note collapse="true"}
## Data Preparation Details

**Date conversions**

| Dataset | Date columns converted |
|------------------------------------|------------------------------------|
| `ProjectSummary` | `ProjectStartDate`, `ProjectEndDate` → `mdy_hms()` |
| `ProjectSummaryWithLocation` | `ActualStartDate`, `ActualEndDate` → `mdy_hms()` |
| `PipelineProjectSummary` | `ActualStartDate`, `ActualEndDate` → `mdy_hms()` |
| `pipeline_cluster`, `clusters` | Use numeric `AllocationYear` — no conversion needed |
| `contributions` | `PledgeDate` parsed on demand during analysis |

**Temporal cutoff — analysis endpoint: December 31, 2025**

-   `AllocationYear != 2026` applied to: `project_summary`, `project_summary_location`, `pipeline_summary`, `pipeline_cluster`
-   `FiscalYear != 2026` applied to: `contributions`

**Helper functions**

-   `safe_date_format()` — returns the min/max of a date vector, filtered to the 2000–2025 range; returns `"N/A"` if no valid dates remain
-   `safe_year_format()` — same logic for numeric year vectors; returns the result as a character string
:::

```{r data-preparation}

# ============================================================
# PART 1: Date and Type Conversions
# ============================================================

# Parse date columns for project datasets
project_summary <- project_summary %>%
  mutate(
    ProjectStartDate = mdy_hms(ProjectStartDate),
    ProjectEndDate = mdy_hms(ProjectEndDate)
  )

# Note: project_summary_location and pipeline_summary use ActualStartDate
project_summary_location <- project_summary_location %>%
  mutate(
    ActualStartDate = mdy_hms(ActualStartDate),
    ActualEndDate = mdy_hms(ActualEndDate)
  )

pipeline_summary <- pipeline_summary %>%
  mutate(
    ActualStartDate = mdy_hms(ActualStartDate),
    ActualEndDate = mdy_hms(ActualEndDate)
  )

# Note: pipeline_cluster and clusters use AllocationYear (numeric)
# Note: contributions uses PledgeDate (will be parsed when needed)

# ============================================================
# PART 2: Filter Out 2026 Data
# ============================================================
# Remove all data from 2026 to maintain analysis cutoff at end of 2025

project_summary <- project_summary %>%
  filter(AllocationYear != 2026)

project_summary_location <- project_summary_location %>%
  filter(AllocationYear != 2026)

pipeline_summary <- pipeline_summary %>%
  filter(AllocationYear != 2026)

pipeline_cluster <- pipeline_cluster %>%
  filter(AllocationYear != 2026)

contributions <- contributions %>%
  filter(FiscalYear != 2026)

# ============================================================
# PART 3: Helper Functions
# ============================================================

# Safely format date ranges, filtering invalid dates
safe_date_format <- function(date_vec, func = min) {
  valid_dates <- date_vec[!is.na(date_vec) & 
                          year(date_vec) >= 2000 & 
                          year(date_vec) <= 2025]
  
  if (length(valid_dates) == 0) return("N/A")
  
  result <- func(valid_dates)
  return(format(result, "%B %d, %Y"))
}

# Format year ranges for datasets that use AllocationYear
safe_year_format <- function(year_vec, func = min) {
  valid_years <- year_vec[!is.na(year_vec) & year_vec >= 2000 & year_vec <= 2025]
  
  if (length(valid_years) == 0) return("N/A")
  
  return(as.character(func(valid_years)))
}
```

```{r dataset-overview}

data_summary <- tibble(
  Dataset = c("Project Summary", "Project Summary (Location)", 
              "Pipeline Summary", "Pipeline Cluster", "Clusters", "Contributions"),
  Rows = c(nrow(project_summary), nrow(project_summary_location),
           nrow(pipeline_summary), nrow(pipeline_cluster), nrow(clusters), 
           nrow(contributions)),
  Columns = c(ncol(project_summary), ncol(project_summary_location),
              ncol(pipeline_summary), ncol(pipeline_cluster), ncol(clusters), 
              ncol(contributions)),
  Countries = c(
    n_distinct(project_summary$PooledFundName),
    n_distinct(project_summary_location$PooledFundName),
    n_distinct(pipeline_summary$PooledFundName),
    n_distinct(pipeline_cluster$PooledFundName),
    n_distinct(clusters$PooledFundName),
    n_distinct(contributions$PooledFundName)
  ),
  `Earliest Entry` = c(
    safe_date_format(project_summary$ProjectStartDate, min),
    safe_date_format(project_summary_location$ActualStartDate, min),
    safe_date_format(pipeline_summary$ActualStartDate, min),
    safe_year_format(pipeline_cluster$AllocationYear, min),
    safe_year_format(clusters$AllocationYear, min),
    safe_date_format(mdy_hms(contributions$PledgeDate), min)
  ),
  `Latest Entry` = c(
    safe_date_format(project_summary$ProjectStartDate, max),
    safe_date_format(project_summary_location$ActualStartDate, max),
    safe_date_format(pipeline_summary$ActualStartDate, max),
    safe_year_format(pipeline_cluster$AllocationYear, max),
    safe_year_format(clusters$AllocationYear, max),
    safe_date_format(mdy_hms(contributions$PledgeDate), max)
  )
)

kable(data_summary, 
      format.args = list(big.mark = ","),
      caption = "Overview of Available Datasets")
```

### Pipeline Data Note

::: callout-important
## Critical Data Limitation: Pipeline Data

The data site offers an endpoint to retrieve projects on the pipeline (under review, cancelled, etc...). After multiple attempts to retrieve comprehensive pipeline data from the CBPF API, it was confirmed that the **PipelineProjectSummary endpoint returns only 581 entries from scattered years** (2013-2015, 2022, 2025-2026), representing just a fraction of all project submissions across the CBPF system's history, and with inconsistencies (including an anomalous pre-2010 entry predating the formal CBPF system).

-   **Ukraine-specific**: Only **1 entry** from 2022 allocation (out of **413 approved Ukraine projects (2019–2025)** in ProjectSummary)
-   **Cannot calculate meaningful approval/rejection rates**: Missing data prevents comprehensive submission vs approval analysis

**What this means:** - We cannot analyze Ukraine’s proposal-to-approval ratio - We cannot determine how many Ukraine projects were rejected or cancelled - We cannot compare Ukraine’s approval success rate to other pooled funds - Analysis must focus on approved projects from the ProjectSummary dataset
:::

# Global summary

OCHA has operated country-based pooled funds in more than 40 countries, with some locations hosting separate cross-border or regional mechanisms. The dataset used here covers the **28 pooled funds active in the 2010–2025 period**, including the Syria Cross-border mechanism and two Regional Humanitarian Pooled Funds (Chad RhPF, Mozambique RhPF).

The data shows the humanitarian gap visible, in recent years, the number of countries receiving funding has increased, but the total budget allocated has gone in the wrong direction.

```{r global-budget-countries}
#| fig-width: 12
#| fig-height: 7

# Aggregate data by year - separate Ukraine from other countries
# Keep all years (2010-2025) to show pilot phase vs mature operational period
global_summary_detailed <- project_summary %>%
  filter(!is.na(AllocationYear), !is.na(Budget)) %>%
  mutate(Country_Group = if_else(PooledFundName == "Ukraine", "Ukraine", "Other Countries")) %>%
  group_by(AllocationYear, Country_Group) %>%
  summarise(Budget = sum(Budget, na.rm = TRUE), .groups = 'drop') %>%
  arrange(AllocationYear, Country_Group)

# Get total budget by year for labels
global_totals <- global_summary_detailed %>%
  group_by(AllocationYear) %>%
  summarise(
    Total_Budget = sum(Budget),
    .groups = 'drop'
  )

# Get country count by year for the line
country_counts <- project_summary %>%
  filter(!is.na(AllocationYear)) %>%
  group_by(AllocationYear) %>%
  summarise(Countries = n_distinct(PooledFundName), .groups = 'drop')

# Merge totals with country counts
global_summary <- global_totals %>%
  left_join(country_counts, by = "AllocationYear")

# Create scaling factor for secondary axis
scale_factor <- max(global_summary$Total_Budget) / max(global_summary$Countries) / 1e6

# Create combination chart with Ukraine highlighted
ggplot() +
  # Stacked bars for budget (Ukraine in yellow, others in teal)
  geom_col(data = global_summary_detailed,
           aes(x = as.factor(AllocationYear), y = Budget/1e6, fill = Country_Group),
           alpha = 0.8, width = 0.7) +
  # Line for countries
  geom_line(data = global_summary,
            aes(x = as.factor(AllocationYear), y = Countries * scale_factor, group = 1),
            color = "#e63946", linewidth = 1.2) +
  geom_point(data = global_summary,
             aes(x = as.factor(AllocationYear), y = Countries * scale_factor),
             color = "#e63946", size = 3) +
  # Labels at top of bars
  geom_text(data = global_summary,
            aes(x = as.factor(AllocationYear), y = Total_Budget/1e6,
                label = paste0("$", round(Total_Budget/1e6, 0), "M")),
            vjust = -0.5, size = 3.5, fontface = "bold", color = "#333333") +
  # Color scale
  scale_fill_manual(
    values = c("Ukraine" = "#f4a261", "Other Countries" = "#005f73"),
    name = "Region"
  ) +
  # Primary y-axis
  scale_y_continuous(
    name = "Total Budget Allocated (Million USD)",
    labels = scales::comma,
    limits = c(0, NA),
    expand = expansion(mult = c(0, 0.1)),
    # Secondary y-axis
    sec.axis = sec_axis(~ . / scale_factor, 
                        name = "Number of Countries",
                        breaks = scales::pretty_breaks(n = 6))
  ) +
  labs(
    title = "Global CBPF Funding Allocation and Country Coverage by Year (2010-2025)",
    subtitle = "Budget shown in bars (Ukraine highlighted in yellow), country count shown as line • Pilot phase (2010-2013) shows limited scale",
    x = "Allocation Year"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 12, color = "#666666"),
    axis.title.y.left = element_text(color = "#005f73", face = "bold", size = 12),
    axis.title.y.right = element_text(color = "#e63946", face = "bold", size = 12),
    axis.text = element_text(size = 11),
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.minor = element_blank(),
    legend.position = "bottom",
    legend.title = element_text(face = "bold")
  )
```

::: callout-important
The 2010-2013 period represents the early pilot phase of CBPF with limited geographic coverage (primarily one country) and significantly lower funding levels. A major scale-up occurred in 2014 when the system expanded to multiple countries and reached operational maturity. **To ensure data comparability, subsequent analysis focuses on the mature operational period from 2014 to 2025.**
:::

## Budget Distribution by Country (2022-2025)

This visualization shows how CBPF funding was distributed across countries for each year from 2022 to 2025. Ukraine is highlighted in yellow to track its position among the top recipients throughout this period.

```{r budget-by-country-yearly}
#| fig-width: 14
#| fig-height: 11

# Get top countries per year
country_budget_yearly <- project_summary %>%
  filter(AllocationYear >= 2022, AllocationYear <= 2025) %>%
  filter(!is.na(Budget), !is.na(PooledFundName)) %>%
  group_by(AllocationYear, PooledFundName) %>%
  summarise(
    Total_Budget = sum(Budget, na.rm = TRUE),
    Projects = n(),
    .groups = 'drop'
  )

# Get top 12 countries per year (or include Ukraine if not in top 12)
top_countries_per_year <- country_budget_yearly %>%
  group_by(AllocationYear) %>%
  arrange(desc(Total_Budget)) %>%
  slice_head(n = 12) %>%
  ungroup()

# Ensure Ukraine is included in all years
ukraine_data <- country_budget_yearly %>%
  filter(PooledFundName == "Ukraine")

# Combine and remove duplicates
country_budget_yearly_filtered <- bind_rows(top_countries_per_year, ukraine_data) %>%
  distinct(AllocationYear, PooledFundName, .keep_all = TRUE)

# Add highlighting, labels, and ordering within each year
country_budget_yearly_filtered <- country_budget_yearly_filtered %>%
  mutate(
    Highlight = if_else(PooledFundName == "Ukraine", "Ukraine", "Other")
  ) %>%
  group_by(AllocationYear) %>%
  mutate(
    Rank = rank(-Total_Budget),
    Label = if_else(Rank <= 8 | PooledFundName == "Ukraine",
                    paste0("$", round(Total_Budget/1e6, 0), "M"),
                    "")
  ) %>%
  ungroup() %>%
  # Create a compound key for proper ordering within facets
  mutate(
    Year_Country = paste(AllocationYear, PooledFundName, sep = "_"),
    Year_Country = reorder(Year_Country, Total_Budget)
  )

# Create faceted horizontal bar chart
ggplot(country_budget_yearly_filtered, 
       aes(x = Total_Budget/1e6, 
           y = Year_Country)) +
  geom_col(aes(fill = Highlight), alpha = 0.85, width = 0.75) +
  geom_text(aes(label = Label), hjust = -0.1, size = 3, fontface = "bold", color = "#333333") +
  scale_fill_manual(
    values = c("Ukraine" = "#f4a261", "Other" = "#005f73"),
    guide = "none"
  ) +
  scale_y_discrete(labels = function(x) gsub("^\\d{4}_", "", x)) +
  scale_x_continuous(
    labels = scales::comma,
    expand = expansion(mult = c(0, 0.15))
  ) +
  facet_wrap(~ AllocationYear, ncol = 2, scales = "free_y") +
  labs(
    title = "Top Countries by CBPF Budget Allocation per Year (2022-2025)",
    subtitle = "Ukraine highlighted in yellow • Shows top recipients each year and how rankings shift",
    x = "Budget (Million USD)",
    y = NULL
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 16, margin = margin(b = 5)),
    plot.subtitle = element_text(size = 11, color = "#666666", margin = margin(b = 15)),
    axis.text.y = element_text(size = 9.5),
    axis.text.x = element_text(size = 9),
    axis.title.x = element_text(face = "bold", size = 11, margin = margin(t = 10)),
    strip.text = element_text(face = "bold", size = 13, margin = margin(b = 10)),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    panel.spacing = unit(1.5, "lines"),
    plot.margin = margin(15, 15, 15, 15)
  )
```

Afghanistan is the only country that has received more funding than Ukraine in 2022, this is due to commitments made at the global level following the Taliban takeover, probably made earlier than February 2022 as contributions to the appeal.

Also note, that this cannot be taken as an indicator of the *global humanitarian attention*, CBPFs is a flexible global mechanism that can balance the funding situation between different crises, considering the attention given by other donors.

# Allocation Flow Analysis

Country-Based Pool Funds have the two ways of publish allocations: the standard allocation and the reserve allocation.

-   **Standard Allocation**: Regular funding cycles based on strategic response plans and humanitarian response planning
-   **Reserve Allocation**: Emergency funding for urgent needs, rapid response to sudden onset crises, and critical gaps

Key Differences from Standard Allocations:

-   **Speed:** The timeline is compressed. A Reserve Allocation can be processed from launch to signature in as little as 48-72 hours in extreme cases, though 2-3 weeks is more common.
-   **Targeting:** The strategy is highly prescriptive. The HC may define not just the geographic area but the specific activities required (e.g., "Provision of emergency shelter kits in District X").
-   **Partner Selection:** While Standard Allocations are open calls, Reserve Allocations can be "closed" or restricted. The HC may invite specific partners who are known to have presence and capacity in the affected area to submit proposals, thereby bypassing the open competitive process to save time.

## Budget Distribution by Allocation Source

**UHF is among the heaviest users of the reserve allocation in the CBPF system**, with Reserve accounting for **51.9 % of its 2019–2025 budget** — well above the non-Ukraine average of 40.2 % and placing Ukraine 7th of 28 funds on this metric (alongside Sudan, oPt and DRC). This chart compares the distribution of standard vs reserve allocations for Ukraine against the rest of the world, showing that while Ukraine has a higher reliance on reserve allocations, there is a trend towards more balanced funding in 2025.

```{r allocation-count-by-source}
#| fig-width: 12
#| fig-height: 9

# Ukraine allocation summary
ukraine_allocation_summary <- project_summary %>%
  filter(AllocationYear >= 2014, 
         !is.na(AllocationSourceName),
         PooledFundName == "Ukraine") %>%
  group_by(AllocationSourceName) %>%
  summarise(
    Projects = n(),
    Budget = sum(Budget, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  mutate(
    Region = "Ukraine",
    # Force factor ordering: Standard first, Reserve second
    AllocationSourceName = factor(AllocationSourceName, levels = c("Standard", "Reserve"))
  ) %>%
  arrange(AllocationSourceName)

# Rest of World allocation summary (excluding Ukraine)
rest_of_world_summary <- project_summary %>%
  filter(AllocationYear >= 2014, 
         !is.na(AllocationSourceName),
         PooledFundName != "Ukraine") %>%
  group_by(AllocationSourceName) %>%
  summarise(
    Projects = n(),
    Budget = sum(Budget, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  mutate(
    Region = "Rest of World",
    # Force factor ordering: Standard first, Reserve second
    AllocationSourceName = factor(AllocationSourceName, levels = c("Standard", "Reserve"))
  ) %>%
  arrange(AllocationSourceName)

# Combine both datasets and ensure ordering
comparison_data <- bind_rows(ukraine_allocation_summary, rest_of_world_summary) %>%
  arrange(Region, AllocationSourceName)

# Recalculate percentages within each region
comparison_data <- comparison_data %>%
  group_by(Region) %>%
  mutate(
    Projects_Pct = round((Projects / sum(Projects)) * 100, 1),
    Budget_Pct = round((Budget / sum(Budget)) * 100, 1)
  ) %>%
  ungroup() %>%
  arrange(Region, AllocationSourceName)

# Budget proportions chart - Standard at bottom (dark teal), Reserve on top (light teal)
ggplot(comparison_data, 
       aes(x = Region, y = Budget_Pct, fill = AllocationSourceName)) +
  geom_col(position = position_stack(reverse = FALSE), width = 0.6) +
  geom_text(aes(label = paste0(AllocationSourceName, "\n", Budget_Pct, "%\n($", round(Budget/1e6, 0), "M)")),
            position = position_stack(vjust = 0.5, reverse = FALSE),
            color = "white",
            fontface = "bold",
            size = 3.8,
            lineheight = 0.85) +
  scale_fill_manual(values = c("Standard" = "#005f73", "Reserve" = "#94d2bd"),
                    name = "Allocation Type",
                    breaks = c("Standard", "Reserve")) +
  scale_y_continuous(labels = function(x) paste0(x, "%"),
                     expand = expansion(mult = c(0, 0.02))) +
  labs(
    title = "Budget Distribution by Allocation Type: Ukraine vs Rest of World (2014-2025)",
    subtitle = "Standard allocations (dark teal) at bottom, Reserve allocations (light teal) on top",
    x = NULL,
    y = "Percentage of Budget"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 11, color = "#666666"),
    panel.grid.major.x = element_blank(),
    axis.text.x = element_text(size = 12, face = "bold"),
    legend.position = "top",
    legend.text = element_text(size = 11, face = "bold")
  )
 
```

::: callout-note
Some allocation rounds that partners perceived as reserved are recorded in the CBPF database as standard allocations. For example, the 2024 Kharkiv area-based allocation was treated by some partners as effectively reserved, as some eligible organizations were reportedly advised not to submit proposals. In the source data, however, these rounds are classified and counted as standard.
:::

```{r allocation-source-by-year}
#| fig-width: 10
#| fig-height: 6

# Calculate budget by allocation source and year for UKRAINE only (2019-2025)
ukraine_allocation_by_year <- project_summary %>%
  filter(PooledFundName == "Ukraine", 
         AllocationYear >= 2019,
         !is.na(AllocationSourceName)) %>%
  group_by(AllocationYear, AllocationSourceName) %>%
  summarise(Total_Budget = sum(Budget, na.rm = TRUE), .groups = 'drop') %>%
  # Force factor ordering BEFORE calculating percentages
  mutate(
    AllocationSourceName = factor(AllocationSourceName, levels = c("Standard", "Reserve"))
  ) %>%
  arrange(AllocationYear, AllocationSourceName) %>%
  group_by(AllocationYear) %>%
  mutate(
    Year_Total = sum(Total_Budget),
    Percentage = round((Total_Budget / Year_Total) * 100, 1)
  ) %>%
  ungroup() %>%
  arrange(AllocationYear, AllocationSourceName)

# Visualization - line chart showing percentage over time
# Convert to millions for display
ukraine_allocation_by_year <- ukraine_allocation_by_year %>%
  mutate(Budget_Millions = Total_Budget / 1e6)

ggplot(ukraine_allocation_by_year, 
       aes(x = AllocationYear, y = Percentage, color = AllocationSourceName, group = AllocationSourceName)) +
  geom_line(linewidth = 1.5) +
  geom_point(size = 4) +
  geom_text(aes(label = paste0(Percentage, "%")),
            vjust = -1.2, 
            size = 3.5,
            fontface = "bold",
            show.legend = FALSE) +
  scale_color_manual(values = c("Standard" = "#005f73", "Reserve" = "#94d2bd"),
                     name = "Allocation Type",
                     breaks = c("Standard", "Reserve")) +
  scale_y_continuous(labels = function(x) paste0(x, "%"),
                     limits = c(0, 100),
                     breaks = seq(0, 100, 20),
                     expand = expansion(mult = c(0.05, 0.15))) +
  scale_x_continuous(breaks = seq(2019, 2025, 1)) +
  labs(
    title = "Ukraine Allocation Type Distribution Over Time (2019-2025)",
    subtitle = "Percentage of budget allocated through Standard vs Reserve modalities by year",
    x = "Allocation Year",
    y = "Percentage of Total Budget"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 11, color = "#666666"),
    legend.position = "top",
    legend.text = element_text(size = 11, face = "bold"),
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(size = 11)
  )
```

Although the balance between standard and reserve allocations appears to be correcting in 2025, it is important to note that since the counter-offensives of 2022, there have not been major events — such as large-scale population movements or recaptured territories — that would justify the recurrent use of the reserve modality. **Ukraine's humanitarian needs can be considered sustainably high, but not necessarily volatile.**

The global CBPF guidelines state that **only exceptionally** should the reserve allocation be used for non-emergency purposes: *"the Reserve Allocation may also be used to support special initiatives which may not be rapid onset or unforeseen per se..."* (CBPF Global Guidelines, art. 136).

## Budget Distribution by Type of Organization

CBPF allocations reach beneficiaries through four types of implementing partners: International NGOs (INGOs), National NGOs (NNGOs), UN Agencies, and a residual "Others" category. The balance between these partner types reflects both operational realities on the ground and broader policy commitments — notably the Grand Bargain pledge to channel at least 25% of humanitarian funding to local and national actors.

```{r org-type-combined}
#| fig-width: 14
#| fig-height: 7

org_colors <- c("National NGO"      = "#2a9d8f",
                "International NGO" = "#005f73",
                "UN Agency"         = "#e9c46a",
                "Others"            = "#adb5bd")

# --- Left panel: proportional comparison Ukraine vs Rest of World ---
org_type_ukraine <- project_summary %>%
  filter(AllocationYear >= 2014, !is.na(OrganizationType),
         PooledFundName == "Ukraine") %>%
  group_by(OrganizationType) %>%
  summarise(Budget = sum(Budget, na.rm = TRUE), Projects = n(), .groups = "drop") %>%
  mutate(Region = "Ukraine")

org_type_row <- project_summary %>%
  filter(AllocationYear >= 2014, !is.na(OrganizationType),
         PooledFundName != "Ukraine") %>%
  group_by(OrganizationType) %>%
  summarise(Budget = sum(Budget, na.rm = TRUE), Projects = n(), .groups = "drop") %>%
  mutate(Region = "Rest of World")

org_type_comparison <- bind_rows(org_type_ukraine, org_type_row) %>%
  group_by(Region) %>%
  mutate(
    Budget_Pct = round((Budget / sum(Budget)) * 100, 1),
    OrganizationType = factor(OrganizationType, levels = names(org_colors))
  ) %>%
  ungroup()

p_comparison <- ggplot(org_type_comparison,
       aes(x = Region, y = Budget_Pct, fill = OrganizationType)) +
  geom_col(position = position_stack(reverse = TRUE), width = 0.55) +
  geom_text(aes(label = paste0(OrganizationType, "\n", Budget_Pct, "%")),
            position = position_stack(vjust = 0.5, reverse = TRUE),
            color = "white", fontface = "bold", size = 3.2, lineheight = 0.9) +
  scale_fill_manual(values = org_colors, name = "Organization Type") +
  scale_y_continuous(labels = function(x) paste0(x, "%"),
                     expand = expansion(mult = c(0, 0.02))) +
  labs(title = "Ukraine vs Rest of World",
       subtitle = "Budget share by org type (2014-2025)",
       x = NULL, y = "Percentage of Budget") +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(size = 10, color = "#666666"),
    panel.grid.major.x = element_blank(),
    axis.text.x = element_text(size = 11, face = "bold"),
    legend.position = "none"
  )

# --- Right panel: Ukraine absolute budget over time ---
ukraine_org_year <- project_summary %>%
  filter(PooledFundName == "Ukraine", AllocationYear >= 2022,
         !is.na(OrganizationType)) %>%
  group_by(AllocationYear, OrganizationType) %>%
  summarise(Budget = sum(Budget, na.rm = TRUE), .groups = "drop") %>%
  mutate(OrganizationType = factor(OrganizationType, levels = names(org_colors)))

p_over_time <- ggplot(ukraine_org_year,
       aes(x = as.factor(AllocationYear), y = Budget / 1e6, fill = OrganizationType)) +
  geom_col(position = position_stack(reverse = TRUE), width = 0.7, alpha = 0.9) +
  scale_fill_manual(values = org_colors, name = "Organization Type") +
  scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.08))) +
  labs(title = "Ukraine Budget by Organization Type Over Time",
       subtitle = "USD millions (2022-2025)",
       x = "Allocation Year", y = "Budget (USD millions)") +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(size = 10, color = "#666666"),
    axis.text.x = element_text(size = 9, angle = 45, hjust = 1),
    legend.position = "top",
    legend.text = element_text(size = 9)
  )

p_comparison + p_over_time + plot_layout(widths = c(1, 2))
```

## Budget Distribution by Sector (Cluster)

This section examines how humanitarian funding is distributed across sectors, comparing the global CBPF portfolio with Ukraine's UHF allocations (2014–2025). Sector labels in the CBPF system correspond to inter-agency clusters (e.g., Health, WASH, Protection).

```{r cluster-prep}
#| include: false

# Build sector-level aggregates from the clusters dataset (already filtered to ≤2025)
clusters_clean <- clusters %>%
  filter(AllocationYear != 2026, !is.na(Cluster), Cluster != "")

# Standardise minor cluster labels for readability
clusters_clean <- clusters_clean %>%
  mutate(Cluster = case_when(
    Cluster == "Camp Coordination / Management" ~ "Camp Coordination/Mgmt",
    Cluster == "Coordination and Support Services" ~ "Coord. & Support Services",
    Cluster == "Emergency Shelter and NFI" ~ "Emergency Shelter & NFI",
    Cluster == "Emergency Telecommunications" ~ "Emergency Telecoms",
    Cluster == "Water, Sanitation and Hygiene" ~ "WASH",
    TRUE ~ Cluster
  ))

# ---------- Global aggregate (all funds, 2014-2025) ----------
global_cluster <- clusters_clean %>%
  filter(AllocationYear >= 2014) %>%
  group_by(Cluster) %>%
  summarise(Total = sum(ClusterBudget, na.rm = TRUE), .groups = "drop") %>%
  mutate(Share = Total / sum(Total) * 100,
         Source = "Global CBPF")

# ---------- Ukraine aggregate (2014-2025) ----------
ukraine_cluster <- clusters_clean %>%
  filter(PooledFundName == "Ukraine", AllocationYear >= 2014) %>%
  group_by(Cluster) %>%
  summarise(Total = sum(ClusterBudget, na.rm = TRUE), .groups = "drop") %>%
  mutate(Share = Total / sum(Total) * 100,
         Source = "Ukraine UHF")

# Ordered by global share for consistent axis
cluster_order <- global_cluster %>%
  arrange(Share) %>%
  pull(Cluster)

# Combined for comparison chart
comparison_df <- bind_rows(global_cluster, ukraine_cluster) %>%
  mutate(Cluster = factor(Cluster, levels = cluster_order))
```

### Global vs Ukraine: Sector Budget Share

```{r cluster-comparison}
#| fig-width: 12
#| fig-height: 7

# Colour palette: one shade per source
source_colors <- c("Global CBPF" = "#94d2bd", "Ukraine UHF" = "#005f73")

ggplot(comparison_df,
       aes(x = Cluster, y = Share, fill = Source)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.65, alpha = 0.9) +
  geom_text(aes(label = paste0(round(Share, 1), "%")),
            position = position_dodge(width = 0.75),
            hjust = -0.15, size = 3.2, fontface = "bold") +
  coord_flip() +
  scale_fill_manual(values = source_colors, name = NULL) +
  scale_y_continuous(
    labels = function(x) paste0(x, "%"),
    expand = expansion(mult = c(0, 0.25))
  ) +
  labs(
    title = "Sector Budget Share: Global CBPF vs Ukraine UHF (2014–2025)",
    subtitle = "Percentage of total allocated budget per sector",
    x = NULL, y = "Share of Total Budget (%)",
    caption = "Source: CBPF Cluster dataset, Feb 2026 extract"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold", size = 14, color = "#005f73"),
    plot.subtitle = element_text(size = 11, color = "#555555"),
    legend.position = "top",
    legend.text   = element_text(size = 11, face = "bold"),
    panel.grid.major.y = element_blank(),
    panel.grid.minor   = element_blank(),
    axis.text.y        = element_text(size = 11)
  )
```

::: callout-note
**Key observations:** Emergency Shelter & NFI and Protection dominate Ukraine's portfolio—reflecting the conflict-driven displacement crisis—while globally WASH and Health are the largest sectors. Ukraine's WASH and Health shares are notably lower than the global average, highlighting a context-specific funding pattern.
:::

### Ukraine Sector Budget Evolution Over Time

The stacked view below tracks how the Ukraine UHF cluster mix has evolved year-on-year since the full-scale invasion. Health is highlighted in red — its absolute envelope has grown, but its share of the portfolio remains compressed by the scale of Protection and Shelter / NFI spending.

```{r cluster-ukraine-trend}
#| fig-width: 12
#| fig-height: 6

# Yearly breakdown for Ukraine per cluster
ukraine_yearly <- clusters_clean %>%
  filter(PooledFundName == "Ukraine", AllocationYear >= 2022) %>%
  group_by(AllocationYear, Cluster) %>%
  summarise(Budget_M = sum(ClusterBudget, na.rm = TRUE) / 1e6, .groups = "drop")

sector_palette <- c(
  "WASH"                         = "#0077B6",
  "Health"                       = "#e63946",
  "Food Security"                = "#90E0EF",
  "Emergency Shelter & NFI"      = "#005f73",
  "Protection"                   = "#0A9396",
  "Nutrition"                    = "#94D2BD",
  "Education"                    = "#E9C46A",
  "Camp Coordination/Mgmt"       = "#F4A261",
  "Multi-purpose CASH"           = "#E76F51",
  "Coord. & Support Services"    = "#ae2012",
  "Multi-Sector"                 = "#9b2226",
  "Early Recovery"               = "#6d6875",
  "Logistics"                    = "#b5b5b5",
  "Emergency Telecoms"           = "#d4d4d4",
  "COVID-19"                     = "#ccc5b9"
)

ukraine_yearly <- ukraine_yearly %>%
  mutate(Cluster = factor(Cluster, levels = rev(cluster_order)))

ggplot(ukraine_yearly,
       aes(x = as.factor(AllocationYear), y = Budget_M, fill = Cluster)) +
  geom_col(position = position_stack(reverse = TRUE), width = 0.75, alpha = 0.92) +
  scale_fill_manual(values = sector_palette, name = "Sector") +
  scale_y_continuous(
    labels = scales::comma,
    expand = expansion(mult = c(0, 0.06))
  ) +
  labs(
    title = "Ukraine UHF: Sector Budget Distribution Over Time (2022–2025)",
    subtitle = "USD millions, stacked by cluster  |  \u2665 Health sector highlighted",
    x = "Allocation Year", y = "Budget (USD millions)",
    caption = "Source: CBPF Cluster dataset, Feb 2026 extract"
  ) +
  guides(fill = guide_legend(nrow = 3, byrow = TRUE)) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold", size = 14, color = "#005f73"),
    plot.subtitle = element_text(size = 11, color = "#555555"),
    legend.position = "bottom",
    legend.text   = element_text(size = 9),
    panel.grid.minor   = element_blank(),
    axis.text.x        = element_text(size = 10, angle = 45, hjust = 1)
  )
```

### Divergence from Global Average

```{r cluster-divergence}
#| fig-width: 11
#| fig-height: 6

# Calculate difference: Ukraine share minus global share
divergence_df <- ukraine_cluster %>%
  select(Cluster, Ukraine_Share = Share) %>%
  left_join(
    global_cluster %>% select(Cluster, Global_Share = Share),
    by = "Cluster"
  ) %>%
  mutate(
    Global_Share  = replace_na(Global_Share, 0),
    Diff          = Ukraine_Share - Global_Share,
    Direction     = if_else(Diff >= 0, "Above global average", "Below global average"),
    Cluster       = factor(Cluster, levels = cluster_order)
  )

ggplot(divergence_df,
       aes(x = Cluster, y = Diff, fill = Direction)) +
  geom_col(width = 0.7, alpha = 0.9) +
  geom_hline(yintercept = 0, linewidth = 0.8, color = "grey30") +
  geom_text(aes(label = paste0(if_else(Diff > 0, "+", ""), round(Diff, 1), " pp"),
                hjust = if_else(Diff >= 0, -0.1, 1.1)),
            size = 3.4, fontface = "bold") +
  coord_flip() +
  scale_fill_manual(
    values = c("Above global average" = "#005f73", "Below global average" = "#e76f51"),
    name = NULL
  ) +
  scale_y_continuous(
    labels = function(x) paste0(x, " pp"),
    expand = expansion(mult = c(0.25, 0.25))
  ) +
  labs(
    title = "Ukraine UHF Sector Share vs Global CBPF Average (2014–2025)",
    subtitle = "Percentage-point difference (Ukraine minus global); pp = percentage points",
    x = NULL, y = "Difference (percentage points)",
    caption = "Source: CBPF Cluster dataset, Feb 2026 extract"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold", size = 14, color = "#005f73"),
    plot.subtitle = element_text(size = 11, color = "#555555"),
    legend.position = "top",
    legend.text   = element_text(size = 11, face = "bold"),
    panel.grid.major.y = element_blank(),
    panel.grid.minor   = element_blank(),
    axis.text.y        = element_text(size = 11)
  )
```

# Grant Characteristics and Patterns

This chapter examines how large individual grants are across the CBPF system, comparing global patterns with Ukraine. Grant size reflects strategic choices: larger grants reduce transaction costs but concentrate risk; smaller grants enable broader partner diversity. Ukraine, as a high-intensity conflict fund, may follow different patterns.

## Grant Size Distribution

```{r grant-prep}
#| include: false

# Base dataset: all valid grants 2014-2025
grants <- project_summary %>%
  filter(AllocationYear >= 2014, !is.na(Budget), Budget > 0) %>%
  mutate(
    Context = if_else(PooledFundName == "Ukraine", "Ukraine UHF", "Global CBPF"),
    OrgType = case_when(
      OrganizationType == "International NGO" ~ "INGO",
      OrganizationType == "National NGO"      ~ "NNGO",
      OrganizationType == "UN Agency"         ~ "UN Agency",
      TRUE                                    ~ NA_character_
    )
  ) %>%
  filter(!is.na(OrgType))

# Join primary cluster (highest budget percentage) from cluster dataset
# NOTE: ChfId is not globally unique (856 duplicates across funds); ChfProjectCode is unique.
cl_primary <- clusters %>%
  filter(AllocationYear >= 2014) %>%
  group_by(ChfProjectCode) %>%
  slice_max(Percentage, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(ChfProjectCode, Cluster) %>%
  mutate(Cluster = case_when(
    Cluster == "Water, Sanitation and Hygiene"      ~ "WASH",
    Cluster == "Emergency Shelter and NFI"          ~ "Shelter & NFI",
    Cluster == "Camp Coordination / Management"     ~ "Camp Coord./Mgmt",
    Cluster == "Coordination and Support Services"  ~ "Coord. & Support",
    Cluster == "Emergency Telecommunications"       ~ "Telecoms",
    TRUE ~ Cluster
  ))

grants_cl <- grants %>%
  left_join(cl_primary, by = "ChfProjectCode") %>%
  mutate(Cluster = replace_na(Cluster, "Unknown"),
         is_health = Cluster == "Health")

# Shared colour scheme
ctx_colors <- c("Global CBPF" = "#94d2bd", "Ukraine UHF" = "#005f73")
org_order   <- c("UN Agency", "INGO", "NNGO")
```

### Grant Size: Global vs Ukraine

The density plot below uses a **log₁₀ scale** to handle the wide spread of grant values (from \~\$25K to \$29M). Both distributions are right-skewed; the key question is whether Ukraine's curve sits systematically to the right.

```{r grant-density}
#| fig-width: 11
#| fig-height: 5

# Summary stats for annotation
medians <- grants %>%
  group_by(Context) %>%
  summarise(med = median(Budget), .groups = "drop")

fmt_usd <- function(x) {
  ifelse(x >= 1e6,
    paste0("$", formatC(x / 1e6, format = "f", digits = 2), "M"),
    paste0("$", scales::comma(round(x / 1e3)), "K")
  )
}

ggplot(grants, aes(x = Budget, fill = Context, color = Context)) +
  geom_density(alpha = 0.35, linewidth = 0.9, adjust = 0.8) +
  geom_vline(data = medians,
             aes(xintercept = med, color = Context),
             linetype = "dashed", linewidth = 1.1) +
  geom_label(data = medians,
             aes(x = med, y = Inf,
                 label = paste0("Median\n", fmt_usd(med)),
                 color = Context),
             vjust = 1.3, hjust = -0.08, size = 3.5, fontface = "bold",
             fill = "white", label.size = 0.3, show.legend = FALSE) +
  scale_x_log10(
    labels = scales::label_dollar(scale_cut = scales::cut_short_scale()),
    breaks = c(1e4, 5e4, 1e5, 5e5, 1e6, 5e6, 1e7)
  ) +
  scale_fill_manual(values  = ctx_colors, name = NULL) +
  scale_color_manual(values = ctx_colors, name = NULL) +
  labs(
    title    = "Grant Size Distribution: Global CBPF vs Ukraine UHF (2014–2025)",
    subtitle = "Log₁₀ scale — dashed lines mark medians. Ukraine grants are substantially larger.",
    x = "Grant size (USD, log scale)", y = "Density",
    caption = "Source: CBPF ProjectSummary dataset, Feb 2026 extract"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold", size = 14, color = "#005f73"),
    plot.subtitle = element_text(size = 11, color = "#555555"),
    legend.position = "top",
    legend.text   = element_text(size = 11, face = "bold"),
    panel.grid.minor = element_blank()
  )
```

::: callout-note
**Ukraine grants are 2.6× larger at the median** — \$1.13M vs \$440K globally. This reflects both the scale of needs and Ukraine's profile as a large, well-funded conflict response where donors consolidate resources into fewer, larger grants.
:::

::: callout-note
There is no available data on grant underspending. As a result, it is unclear how often partners do not fully meet their commitments while continuing to receive funding.
:::

### INGO vs NNGO Grant Size: A Deeper Look

Globally (excluding Ukraine), INGOs receive grants about **25 % larger** than NNGOs at the median (ratio ≈ 1.25×). Ukraine's ratio (1.40×) is mid-table but masks a striking post-2022 shift.

### How Does Ukraine Compare Across All Funds?

```{r ingo-nngo-dumbbell}
#| fig-width: 12
#| fig-height: 9

# Per-fund medians for INGO and NNGO (min 10 grants each)
fund_ngotype <- project_summary %>%
  filter(AllocationYear >= 2014,
         !is.na(Budget), Budget > 0,
         OrganizationType %in% c("International NGO", "National NGO")) %>%
  group_by(PooledFundName, OrganizationType) %>%
  summarise(med = median(Budget), n = n(), .groups = "drop")

fund_wide <- fund_ngotype %>%
  pivot_wider(names_from = OrganizationType, values_from = c(med, n)) %>%
  rename(ingo_med = `med_International NGO`,
         nngo_med = `med_National NGO`,
         ingo_n   = `n_International NGO`,
         nngo_n   = `n_National NGO`) %>%
  filter(!is.na(ingo_med), !is.na(nngo_med),
         ingo_n >= 10, nngo_n >= 10) %>%
  mutate(
    ratio     = ingo_med / nngo_med,
    is_ukraine = PooledFundName == "Ukraine",
    PooledFundName = factor(PooledFundName, levels = PooledFundName[order(ratio)])
  )

# Long form for dumbbell
dumbbell_df <- fund_wide %>%
  select(PooledFundName, ingo_med, nngo_med, ratio, is_ukraine) %>%
  pivot_longer(cols = c(ingo_med, nngo_med),
               names_to = "OrgType", values_to = "Median") %>%
  mutate(OrgType = if_else(OrgType == "ingo_med", "INGO", "NNGO"))

ggplot() +
  # segment connecting NNGO to INGO per fund
  geom_segment(
    data = fund_wide,
    aes(x = nngo_med / 1e3, xend = ingo_med / 1e3,
        y = PooledFundName, yend = PooledFundName,
        color = is_ukraine),
    linewidth = 1.2, alpha = 0.7
  ) +
  # dots for each org type
  geom_point(
    data = dumbbell_df,
    aes(x = Median / 1e3, y = PooledFundName,
        shape = OrgType, color = is_ukraine),
    size = 3.5, stroke = 1.0
  ) +
  # ratio label to the right
  geom_text(
    data = fund_wide,
    aes(x = pmax(ingo_med, nngo_med) / 1e3,
        y = PooledFundName,
        label = paste0(round(ratio, 2), "×"),
        fontface = if_else(is_ukraine, "bold", "plain")),
    hjust = -0.25, size = 3.1, color = "grey30"
  ) +
  scale_x_continuous(
    labels = scales::dollar_format(suffix = "K", accuracy = 1),
    expand = expansion(mult = c(0.02, 0.18))
  ) +
  scale_color_manual(
    values = c("TRUE" = "#f4a261", "FALSE" = "#94d2bd"),
    guide  = "none"
  ) +
  scale_shape_manual(values = c("INGO" = 16, "NNGO" = 1), name = NULL) +
  labs(
    title    = "Median Grant Size: INGO vs NNGO per CBPF Fund (2014–2025)",
    subtitle = "Sorted by INGO/NNGO ratio (right labels)  |  <b style='color:#f4a261'>Ukraine highlighted</b>  |  filled = INGO, open = NNGO",
    x = "Median grant size (USD thousands)", y = NULL,
    caption = "Funds with < 10 grants per org type excluded"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title    = element_text(face = "bold", size = 14, color = "#005f73"),
    plot.subtitle = element_markdown(size = 10, color = "#555555"),
    panel.grid.major.y = element_blank(),
    panel.grid.minor   = element_blank(),
    legend.position    = "top",
    legend.text        = element_text(size = 11, face = "bold"),
    axis.text.y        = element_text(
      size  = ifelse(levels(fund_wide$PooledFundName) == "Ukraine", 12, 10),
      face  = ifelse(levels(fund_wide$PooledFundName) == "Ukraine", "bold", "plain"),
      color = ifelse(levels(fund_wide$PooledFundName) == "Ukraine", "#f4a261", "grey30")
    )
  )
```

What the chart presents: Each horizontal segment represents one CBPF fund. The open circle is the NNGO median grant, the filled circle is the INGO median. The further right, the larger the absolute grants. The ratio label is what matters for the comparison.

The headline finding: Ukraine's ratio is unremarkable At 1.40×, Ukraine sits mid-pack — **14th of 24 funds** that have at least 10 INGO and 10 NNGO grants in 2014–2025. The figure is close to the non-Ukraine global benchmark (≈1.25×) and far from the high-ratio contexts (Sudan 2.26×, Nigeria 2.00×). This is counterintuitive given Ukraine's reputation as a progressive localization context. The ratio alone doesn't signal any particular INGO advantage.

But the absolute values tell a different story Ukraine's NNGO median (\~\$999K) is likely higher in absolute terms than many funds' INGO medians. A Ukrainian NNGO receives grants that would be considered large even by INGO standards in Sudan or Nigeria. The localization story in Ukraine is not just about equality of ratio — it's that everyone operates at a much higher scale.

The high-ratio funds (Sudan 2.26×, Nigeria/Haiti 2.0×, Iraq 1.88×) are contexts where the NNGO ecosystem is less mature — smaller organizations, less absorptive capacity, so funders intentionally give them smaller grants to manage risk. The INGO premium there reflects structural capacity gaps.

The low-ratio end (Ethiopia 0.87×, Somalia 1.02×, Syria cross-border 1.07×) is where NNGO capacity has caught up or where large national NGOs dominate the landscape, compressing the gap.

The implication for Ukraine: The ratio of 1.40× aggregated over 2014–2025 understates how much has changed recently — that's exactly what the trend chart below it reveals, where the ratio collapsed to 0.84× in 2024. The aggregate figure is dragged up by the early pre-war years when Ukraine operated more like a typical fund.

### Ukraine: INGO/NNGO Grant Ratio Over Time

Since 2022, Ukraine's NNGO grants have grown faster than INGO grants. In 2023 and 2024 the ratio **fell below 1** — NNGOs received larger median grants than INGOs — a direct signal of the localization push intensifying after the full-scale invasion.

```{r ingo-nngo-ratio-trend}
#| fig-width: 12
#| fig-height: 5

# Year-by-year ratio for Ukraine and global average
ngo_ratio_all <- project_summary %>%
  filter(AllocationYear >= 2014,
         !is.na(Budget), Budget > 0,
         OrganizationType %in% c("International NGO", "National NGO")) %>%
  group_by(PooledFundName, AllocationYear, OrganizationType) %>%
  summarise(med = median(Budget), n = n(), .groups = "drop")

make_ratio <- function(df) {
  df %>%
    pivot_wider(names_from = OrganizationType, values_from = c(med, n)) %>%
    rename(ingo = `med_International NGO`, nngo = `med_National NGO`,
           n_i  = `n_International NGO`,  n_n  = `n_National NGO`) %>%
    filter(!is.na(ingo), !is.na(nngo), n_i >= 5, n_n >= 5) %>%
    mutate(ratio = ingo / nngo)
}

ukraine_ratio <- ngo_ratio_all %>%
  filter(PooledFundName == "Ukraine") %>%
  make_ratio() %>%
  mutate(Series = "Ukraine UHF")

global_ratio <- ngo_ratio_all %>%
  filter(PooledFundName != "Ukraine") %>%
  group_by(AllocationYear, OrganizationType) %>%
  summarise(med = median(med), n = sum(n), .groups = "drop") %>%
  mutate(PooledFundName = "Global") %>%
  make_ratio() %>%
  mutate(Series = "Global CBPF average")

ratio_trend <- bind_rows(ukraine_ratio, global_ratio)

series_colors <- c("Ukraine UHF" = "#005f73", "Global CBPF average" = "#94d2bd")

ggplot(ratio_trend,
       aes(x = AllocationYear, y = ratio, color = Series, group = Series)) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "#e63946",
             linewidth = 0.9) +
  annotate("text", x = 2014.2, y = 1.03,
           label = "Parity (INGO = NNGO)", color = "#e63946",
           size = 3.4, hjust = 0, fontface = "italic") +
  geom_ribbon(
    data = ratio_trend %>% filter(Series == "Ukraine UHF"),
    aes(ymin = 1, ymax = ratio, fill = ratio < 1),
    alpha = 0.15, color = NA
  ) +
  geom_line(linewidth = 1.5) +
  geom_point(size = 4) +
  geom_text(aes(label = round(ratio, 2)),
            vjust = -1.1, size = 3.3, fontface = "bold",
            show.legend = FALSE) +
  scale_color_manual(values = series_colors, name = NULL) +
  scale_fill_manual(values = c("FALSE" = "#005f73", "TRUE" = "#e63946"),
                    guide = "none") +
  scale_x_continuous(breaks = 2014:2025) +
  scale_y_continuous(
    labels = function(x) paste0(x, "×"),
    breaks = seq(0.7, 1.8, 0.1)
  ) +
  labs(
    title    = "INGO/NNGO Median Grant Ratio Over Time: Ukraine vs Global (2014–2025)",
    subtitle = "Ratio > 1 means INGOs get larger grants  |  Red shading = years where NNGOs received larger grants (ratio < 1)",
    x = "Allocation Year", y = "INGO median / NNGO median",
    caption = "Global = median of fund-level medians, excluding Ukraine"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold", size = 14, color = "#005f73"),
    plot.subtitle = element_text(size = 11, color = "#555555"),
    legend.position = "top",
    legend.text   = element_text(size = 11, face = "bold"),
    panel.grid.minor = element_blank(),
    axis.text.x      = element_text(angle = 45, hjust = 1, size = 10)
  )
```

A Ukrainian NNGO receives grants that would be considered large even by INGO standards in Sudan or Nigeria.

::: callout-note
**Ukraine's localization in action**: while globally the INGO size premium has been stable at \~1.2–1.4×, Ukraine's ratio dropped to **0.94× in 2023** and **0.84× in 2024**, meaning NNGOs were receiving larger individual grants than INGOs. This likely reflects deliberate fund strategy to direct larger, more complex grants to national partners after the scale-up in 2022.
:::

### Grant Size by Sector — Health in Context

Using each project's primary cluster (the sector receiving the highest budget share), this chart compares median grant sizes across sectors. Health grants are highlighted in red.

```{r grant-by-sector}
#| fig-width: 12
#| fig-height: 7

# Compute median + IQR per cluster for both contexts combined
sector_stats <- grants_cl %>%
  filter(Cluster != "Unknown") %>%
  group_by(Cluster) %>%
  summarise(
    med    = median(Budget),
    q1     = quantile(Budget, 0.25),
    q3     = quantile(Budget, 0.75),
    n      = n(),
    .groups = "drop"
  ) %>%
  arrange(med) %>%
  mutate(
    Cluster    = factor(Cluster, levels = Cluster),
    is_health  = Cluster == "Health"
  )

# Separate Ukraine medians to overlay
ukraine_med <- grants_cl %>%
  filter(Cluster != "Unknown", PooledFundName == "Ukraine") %>%
  group_by(Cluster) %>%
  summarise(ukraine_med = median(Budget), .groups = "drop") %>%
  mutate(Cluster = factor(Cluster, levels = levels(sector_stats$Cluster)))

sector_plot_df <- sector_stats %>%
  left_join(ukraine_med, by = "Cluster")

ggplot(sector_plot_df,
       aes(x = Cluster, y = med / 1e3, fill = is_health)) +
  geom_col(width = 0.7, alpha = 0.88) +
  gghighlight(
    is_health,
    unhighlighted_params = list(fill = "#94d2bd", alpha = 0.6),
    use_direct_label = FALSE
  ) +
  # Ukraine medians as points
  geom_point(
    aes(y = ukraine_med / 1e3),
    shape = 23, size = 3.5, fill = "#005f73", color = "white", stroke = 0.8,
    na.rm = TRUE
  ) +
  geom_text(
    aes(label = paste0("$", scales::comma(round(med / 1e3)), "K")),
    hjust = -0.1, size = 3.2, fontface = "bold"
  ) +
  coord_flip() +
  scale_fill_manual(values = c("FALSE" = "#94d2bd", "TRUE" = "#e63946"),
                    guide = "none") +
  scale_y_continuous(
    labels = scales::dollar_format(suffix = "K", accuracy = 1),
    expand = expansion(mult = c(0, 0.3))
  ) +
  labs(
    title    = "Median Grant Size by Sector — Global CBPF (2014–2025)",
    subtitle = "<b style='color:#e63946'>♥ Health</b> highlighted  |  <b style='color:#005f73'>◆ Diamond</b> = Ukraine UHF median",
    x = NULL, y = "Median grant size (USD thousands)",
    caption = "Source: CBPF ProjectSummary + Cluster datasets, Feb 2026 extract"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold", size = 14, color = "#005f73"),
    plot.subtitle = element_markdown(size = 11, color = "#555555"),
    panel.grid.major.y = element_blank(),
    panel.grid.minor   = element_blank(),
    axis.text.y        = element_markdown(
      size  = ifelse(levels(sector_plot_df$Cluster) == "Health", 12, 11),
      face  = ifelse(levels(sector_plot_df$Cluster) == "Health", "bold", "plain"),
      color = ifelse(levels(sector_plot_df$Cluster) == "Health", "#e63946", "grey30")
    )
  )
```

::: callout-note
**Health grants rank in the middle of the pack globally** (\~\$422K median), well below sectors like Logistics, Multi-purpose Cash, and Shelter & NFI. The ◆ diamonds show that Ukraine's health grants are significantly larger than the global norm, consistent with the fund's overall pattern of larger, more consolidated grants.
:::

### Grant Size vs Number of Grants per Organisation {#sec-org-dotchart}

Each dot is one implementing organisation. The x-axis shows how many grants that organisation received over 2022–2025; the y-axis shows their median grant size. Organisations in the top-left corner are high-value, low-frequency recipients; those in the bottom-right are high-frequency implementers of smaller grants.

```{r org-dotchart}
#| fig-width: 13
#| fig-height: 7

org_stats <- grants_cl %>%
  filter(AllocationYear >= 2022, AllocationYear <= 2025,
         !is.na(Budget), Budget > 0,
         !is.na(OrganizationName)) %>%
  group_by(Context, OrganizationName, OrganizationType) %>%
  summarise(
    n_grants    = n(),
    median_size = median(Budget),
    total_usd   = sum(Budget),
    pct_health  = mean(Cluster == "Health"),
    .groups     = "drop"
  ) %>%
  mutate(
    dot_group = case_when(
      pct_health >= 0.5               ~ "Health (≥50% health grants)",
      Context   == "Ukraine UHF"      ~ "Ukraine UHF",
      TRUE                            ~ "Global CBPF"
    )
  )

# Dynamic ceilings: y = max Ukraine median + 4M, x = max Ukraine n_grants + 3
ukr_max   <- org_stats %>% filter(Context == "Ukraine UHF") %>% pull(median_size) %>% max()
ukr_xmax  <- org_stats %>% filter(Context == "Ukraine UHF") %>% pull(n_grants)    %>% max()
y_ceiling <- ukr_max  + 4e6
x_ceiling <- ukr_xmax + 3

dot_colors <- c(
  "Ukraine UHF"                  = "#f4a261",
  "Global CBPF"                  = "#94d2bd",
  "Health (≥50% health grants)"  = "#e63946"
)

# Parametric ellipse for annotation (log-y space)
theta      <- seq(0, 2 * pi, length.out = 200)
ellipse_df <- data.frame(
  x = 9 + 4.2 * cos(theta),
  y = 10^(6.45 + 0.55 * sin(theta))
)

ggplot(org_stats,
       aes(x = n_grants, y = median_size,
           color = dot_group, size = total_usd)) +
  geom_path(data = ellipse_df, aes(x = x, y = y),
            inherit.aes = FALSE,
            color = "black", linewidth = 0.9, linetype = "solid") +
  annotate("text", x = 9, y = 10^6.9,
           label = "The UHF Exception",
           hjust = 0.5, vjust = 1, size = 4.5, fontface = "bold.italic", color = "black") +
  geom_point(alpha = 0.65, stroke = 0) +
  scale_y_log10(
    labels = scales::label_dollar(scale_cut = scales::cut_short_scale()),
    breaks = c(1e4, 5e4, 1e5, 5e5, 1e6, 5e6, 1e7)
  ) +
  scale_x_continuous(labels = scales::label_comma(),
                     breaks = c(1, 2, 3, 5, 7, 10, 15, 20)) +
  scale_color_manual(values = dot_colors, name = NULL,
                     guide  = guide_legend(override.aes = list(size = 4))) +
  scale_size_continuous(
    name   = "Total disbursed (2022–2025)",
    range  = c(2, 16),
    labels = scales::label_dollar(scale_cut = scales::cut_short_scale()),
    breaks = c(1e6, 5e6, 1e7, 3e7, 5e7),
    trans  = "sqrt"
  ) +
  coord_cartesian(xlim = c(NA, x_ceiling), ylim = c(NA, y_ceiling)) +
  labs(
    title    = "Grant Size vs Number of Grants per Organisation (2022–2025)",
    subtitle = "Each dot = one organisation · Y-axis log-scaled · Dot size = total disbursed · Red = majority health grants",
    x = "Number of grants received",
    y = "Median grant size (USD, log scale)",
    caption = "Source: CBPF ProjectSummary dataset, Feb 2026 extract. Axes clipped at max Ukraine values + buffer."
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold", size = 14, color = "#005f73"),
    plot.subtitle = element_text(size = 11, color = "#555555"),
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )
```

```{r org-dotchart-table}
# Who are the high-value Ukraine outliers?
org_stats %>%
  filter(Context == "Ukraine UHF") %>%
  arrange(desc(total_usd)) %>%
  mutate(
    median_size = scales::dollar(median_size, scale_cut = scales::cut_short_scale(), accuracy = 0.01),
    total_usd   = scales::dollar(total_usd,   scale_cut = scales::cut_short_scale(), accuracy = 0.01),
    is_pure_health = pct_health == 1,
    across(
      c(OrganizationName, OrganizationType),
      ~ ifelse(is_pure_health,
               paste0("<span style='color:#e63946;font-weight:bold'>", .x, "</span>"),
               .x)
    ),
    n_grants    = ifelse(is_pure_health,
                         paste0("<span style='color:#e63946;font-weight:bold'>", n_grants, "</span>"),
                         as.character(n_grants)),
    median_size = ifelse(is_pure_health,
                         paste0("<span style='color:#e63946;font-weight:bold'>", median_size, "</span>"),
                         median_size),
    total_usd   = ifelse(is_pure_health,
                         paste0("<span style='color:#e63946;font-weight:bold'>", total_usd, "</span>"),
                         total_usd),
    pct_health  = ifelse(is_pure_health,
                         "<span style='color:#e63946;font-weight:bold'>100%</span>",
                         paste0(round(pct_health * 100), "%"))
  ) %>%
  select(
    Organisation      = OrganizationName,
    Type              = OrganizationType,
    `Grants received` = n_grants,
    `Median grant`    = median_size,
    `Total disbursed` = total_usd,
    `% Health grants` = pct_health
  ) %>%
  knitr::kable(caption = "Ukraine UHF organisations, 2022–2025 — ranked by total disbursed",
               escape  = FALSE)
```

------------------------------------------------------------------------

# Partners per Allocation

Each CBPF allocation is a discrete funding round (e.g. "1st Standard Allocation 2023 — Ukraine"). The number of implementing partners selected per round is a direct signal of concentration: a fund that channels money through 10 partners per allocation is far more concentrated than one that routes it through 40. Ukraine's profile as a high-intensity conflict fund, combined with its relatively small NNGO ecosystem, raises the question of whether it selects fewer partners than comparable funds globally.

## Distribution: Ukraine vs Global {#sec-partners-dist}

```{r partners-per-alloc-data}
#| message: false

# Number of distinct partners per allocation round
# Use AllocationTypeId (e.g. "2022 3rd Reserve Allocation") as the round-level key.
# AllocationSourceID only takes 2 values (Standard/Reserve) and would conflate rounds.
partners_per_alloc <- project_summary %>%
  filter(AllocationYear >= 2014,
         !is.na(AllocationTypeId),
         !is.na(OrganizationName)) %>%
  group_by(PooledFundName, AllocationTypeId, AllocationType,
           AllocationSourceName, AllocationYear) %>%
  summarise(
    n_partners   = n_distinct(OrganizationName),
    total_budget = sum(Budget, na.rm = TRUE),
    n_projects   = n(),
    .groups = "drop"
  ) %>%
  mutate(Context = if_else(PooledFundName == "Ukraine", "Ukraine UHF", "Global CBPF"))

# Summary stats for callout
ppa_summary <- partners_per_alloc %>%
  group_by(Context) %>%
  summarise(
    n_allocs  = n(),
    med       = median(n_partners),
    q1        = quantile(n_partners, 0.25),
    q3        = quantile(n_partners, 0.75),
    mean_n    = round(mean(n_partners), 1),
    .groups   = "drop"
  )
```

```{r partners-per-alloc-plot}
#| fig-width: 11
#| fig-height: 5

ctx_colors_ppa <- c("Ukraine UHF" = "#005f73", "Global CBPF" = "#94d2bd")

med_labels <- partners_per_alloc %>%
  group_by(Context) %>%
  summarise(med = median(n_partners), .groups = "drop")

ggplot(partners_per_alloc,
       aes(x = n_partners, fill = Context, color = Context)) +
  geom_density(alpha = 0.35, linewidth = 0.9, adjust = 0.9) +
  geom_vline(data = med_labels,
             aes(xintercept = med, color = Context),
             linetype = "dashed", linewidth = 1.1) +
  geom_label(data = med_labels,
             aes(x = med, y = Inf,
                 label = paste0("Median: ", round(med), " partners"),
                 color = Context),
             vjust = 1.3, hjust = -0.07, size = 3.5, fontface = "bold",
             fill = "white", label.size = 0.3, show.legend = FALSE) +
  scale_fill_manual(values  = ctx_colors_ppa, name = NULL) +
  scale_color_manual(values = ctx_colors_ppa, name = NULL) +
  scale_x_continuous(breaks = c(5, 10, 20, 30, 40, 50, 60, 80, 100)) +
  labs(
    title    = "Partners per Allocation Round: Global CBPF vs Ukraine UHF (2014–2025)",
    subtitle = "Each observation is one allocation round; dashed lines mark medians",
    x = "Number of distinct implementing partners",
    y = "Density",
    caption = "Source: CBPF ProjectSummary dataset, Feb 2026 extract"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold", size = 14, color = "#005f73"),
    plot.subtitle = element_text(size = 11, color = "#555555"),
    legend.position = "top",
    legend.text   = element_text(size = 11, face = "bold"),
    panel.grid.minor = element_blank()
  )
```

```{r partners-callout-vals, include=FALSE}
ukr_med  <- ppa_summary$med[ppa_summary$Context == "Ukraine UHF"]
glob_med <- ppa_summary$med[ppa_summary$Context == "Global CBPF"]
ratio    <- round(glob_med / ukr_med, 1)
```

::: callout-note
**Ukraine allocations reach `r ukr_med` partners at the median** vs `r glob_med` globally — Ukraine rounds involve roughly `r round(ukr_med/glob_med, 1)`× more partners per allocation. Combined with the larger median grant size, this points not to a narrower partner base but to **a bigger operational footprint per round** — a reflection of the scale of needs and of the UHF's ability to mobilise a wide partner pool.

*Method note: each observation is one discrete allocation round identified by `AllocationTypeId` (e.g. "2022 3rd Reserve Allocation", "2024 1st Standard Allocation").*
:::

## Partners per Allocation: Trend Over Time (Ukraine) {#sec-partners-trend}

Ukraine's partner count per allocation has evolved markedly since the full-scale invasion in 2022.

```{r partners-trend-ukraine}
#| fig-width: 11
#| fig-height: 5

# Yearly median and IQR for Ukraine
ukr_trend <- partners_per_alloc %>%
  filter(Context == "Ukraine UHF") %>%
  group_by(AllocationYear) %>%
  summarise(
    med    = median(n_partners),
    q1     = quantile(n_partners, 0.25),
    q3     = quantile(n_partners, 0.75),
    n_obs  = n(),
    .groups = "drop"
  )

# Global yearly median for comparison
glob_trend <- partners_per_alloc %>%
  filter(Context == "Global CBPF") %>%
  group_by(AllocationYear) %>%
  summarise(
    med   = median(n_partners),
    q1    = quantile(n_partners, 0.25),
    q3    = quantile(n_partners, 0.75),
    .groups = "drop"
  )

ggplot() +
  # Global ribbon + line
  geom_ribbon(data = glob_trend,
              aes(x = AllocationYear, ymin = q1, ymax = q3),
              fill = "#94d2bd", alpha = 0.25) +
  geom_line(data = glob_trend,
            aes(x = AllocationYear, y = med, color = "Global CBPF median"),
            linewidth = 0.9, linetype = "dotted") +
  # Ukraine ribbon + line
  geom_ribbon(data = ukr_trend,
              aes(x = AllocationYear, ymin = q1, ymax = q3),
              fill = "#005f73", alpha = 0.20) +
  geom_line(data = ukr_trend,
            aes(x = AllocationYear, y = med, color = "Ukraine UHF median"),
            linewidth = 1.2) +
  geom_point(data = ukr_trend,
             aes(x = AllocationYear, y = med, color = "Ukraine UHF median"),
             size = 3) +
  geom_text(data = ukr_trend,
            aes(x = AllocationYear, y = med,
                label = round(med)),
            vjust = -0.9, size = 3.2, fontface = "bold", color = "#005f73") +
  annotate("rect", xmin = 2021.5, xmax = Inf, ymin = -Inf, ymax = Inf,
           fill = "#e63946", alpha = 0.04) +
  annotate("text", x = 2022, y = Inf, label = "Full-scale\ninvasion",
           vjust = 1.4, hjust = 0, size = 3, color = "#e63946", fontface = "italic") +
  scale_color_manual(
    values = c("Ukraine UHF median" = "#005f73", "Global CBPF median" = "#2ca58d"),
    name   = NULL
  ) +
  scale_x_continuous(breaks = 2014:2025) +
  labs(
    title    = "Median Partners per Allocation Round by Year",
    subtitle = "Shaded bands = IQR; dotted line = global median; solid = Ukraine",
    x = "Allocation year", y = "Partners per allocation (median)",
    caption = "Source: CBPF ProjectSummary dataset, Feb 2026 extract"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold", size = 14, color = "#005f73"),
    plot.subtitle = element_text(size = 11, color = "#555555"),
    legend.position = "top",
    legend.text   = element_text(size = 11),
    panel.grid.minor = element_blank(),
    axis.text.x   = element_text(angle = 45, hjust = 1)
  )
```

------------------------------------------------------------------------

# Concentration, Scale, and the Core-20

If UHF involves more partners per round than peer funds and grants larger amounts per grant, a natural question follows: **is the resulting portfolio more concentrated than peer CBPFs, and who sits at the top?** This chapter answers both questions quantitatively. The headline is counterintuitive: UHF is **not** more concentrated than comparable CBPFs — it is simply **larger**. The appearance of concentration is produced by the interaction of envelope size with the \$5M per-grant cap, and by a stable 20-organisation core that has recurred in every allocation round since 2022.

## Ukraine vs Peer CBPFs: Concentration Benchmark

Peer set: the seven other largest CBPFs by 2022–2024 envelope (Afghanistan, Syria Cross-border, Sudan, Yemen, Ethiopia, Somalia, oPt). Two natural classes emerge — **mega-concentrated access funds** (Sudan, Syria XB, with only 3–4 partners per year and HHI \> 0.3) and **broad partner funds** (Afghanistan, Yemen, Ethiopia, Somalia, oPt, Ukraine). Ukraine sits in the second class.

```{r concentration-benchmark}
#| fig-width: 11
#| fig-height: 5

peer_funds <- c("Ukraine", "Afghanistan", "Syria Cross border",
                "Sudan", "Yemen", "Ethiopia", "Somalia", "oPt")

conc_agg <- project_summary %>%
  filter(AllocationYear %in% 2022:2024,
         PooledFundName %in% peer_funds,
         !is.na(Budget), Budget > 0,
         !is.na(OrganizationName)) %>%
  group_by(PooledFundName, OrganizationName) %>%
  summarise(amt = sum(Budget), .groups = "drop") %>%
  group_by(PooledFundName) %>%
  arrange(desc(amt), .by_group = TRUE) %>%
  mutate(share = amt / sum(amt)) %>%
  summarise(
    n_partners = n(),
    envelope   = sum(amt),
    HHI        = sum(share^2),
    top5       = sum(share[1:min(5, n())]),
    top10      = sum(share[1:min(10, n())]),
    .groups = "drop"
  ) %>%
  mutate(
    Class = if_else(n_partners <= 5,
                    "Access-restricted (UN-dominant)",
                    "Broad partner fund"),
    is_ukraine = PooledFundName == "Ukraine"
  )

# Two aligned horizontal-bar panels, funds sorted by HHI.
# Ukraine highlighted in red. The visual punchline: mid-row on HHI,
# near-top row on envelope -> "same shape, bigger envelope."
conc_plot <- conc_agg %>%
  mutate(Fund = forcats::fct_reorder(PooledFundName, HHI))

bar_fill <- c(`TRUE` = "#c9184a", `FALSE` = "#8aa9b0")

p_hhi <- ggplot(conc_plot, aes(x = HHI, y = Fund, fill = is_ukraine)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = sprintf("%.3f", HHI)),
            hjust = -0.15, size = 3.5, colour = "#333333") +
  scale_fill_manual(values = bar_fill, guide = "none") +
  scale_x_continuous(expand = expansion(mult = c(0, 0.18))) +
  labs(title    = "Concentration (HHI)",
       subtitle = "Lower = more evenly spread across partners",
       x = NULL, y = NULL) +
  theme_minimal(base_size = 12) +
  theme(plot.title         = element_text(face = "bold", colour = "#005f73"),
        plot.subtitle      = element_text(size = 10, colour = "#555555"),
        panel.grid.major.y = element_blank(),
        panel.grid.minor   = element_blank())

p_env <- ggplot(conc_plot, aes(x = envelope / 1e6, y = Fund, fill = is_ukraine)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = scales::dollar(envelope / 1e6, suffix = "M", accuracy = 1)),
            hjust = -0.15, size = 3.5, colour = "#333333") +
  scale_fill_manual(values = bar_fill, guide = "none") +
  scale_x_continuous(expand = expansion(mult = c(0, 0.22))) +
  labs(title    = "Envelope 2022–2024",
       subtitle = "Absolute USD disbursed",
       x = NULL, y = NULL) +
  theme_minimal(base_size = 12) +
  theme(plot.title         = element_text(face = "bold", colour = "#005f73"),
        plot.subtitle      = element_text(size = 10, colour = "#555555"),
        axis.text.y        = element_blank(),
        panel.grid.major.y = element_blank(),
        panel.grid.minor   = element_blank())

(p_hhi | p_env) +
  plot_annotation(
    title    = "Ukraine sits mid-pack on concentration — but tops the envelope ranking",
    subtitle = "Same fund order in both panels (sorted by HHI). Ukraine highlighted in red.",
    caption  = "Source: CBPF ProjectSummary dataset, Feb 2026 extract",
    theme = theme(plot.title    = element_text(face = "bold", size = 14, colour = "#005f73"),
                  plot.subtitle = element_text(size = 11, colour = "#555555"))
  )
```

```{r concentration-table}
conc_agg %>%
  mutate(
    Envelope      = scales::dollar(envelope / 1e6, suffix = "M", accuracy = 0.1),
    HHI           = sprintf("%.3f", HHI),
    `Top-5 share` = scales::percent(top5, accuracy = 0.1),
    `Top-10 share` = scales::percent(top10, accuracy = 0.1)
  ) %>%
  arrange(desc(envelope)) %>%
  select(Fund = PooledFundName, Envelope, Partners = n_partners,
         HHI, `Top-5 share`, `Top-10 share`) %>%
  knitr::kable(caption = "CBPF concentration metrics, 2022–2024 (aggregated across the three years).")
```

::: callout-note
**Reading the map.** Among broad partner funds, Ukraine is unremarkable on shape — lower HHI than Afghanistan or Ethiopia, higher than Somalia or Yemen, mid-pack on top-5 (28%) and top-10 (46%) share. What makes it look concentrated is the **envelope size**: at \$541M over three years UHF is twice Yemen's volume on the same HHI. Moderate concentration × large envelope = large absolute disbursements to the top partners.
:::

```{r concentration-trend}
#| fig-width: 10
#| fig-height: 4.5

conc_year <- project_summary %>%
  filter(PooledFundName == "Ukraine",
         AllocationYear %in% 2022:2025,
         !is.na(Budget), Budget > 0,
         !is.na(OrganizationName)) %>%
  group_by(AllocationYear, OrganizationName) %>%
  summarise(amt = sum(Budget), .groups = "drop") %>%
  group_by(AllocationYear) %>%
  arrange(desc(amt), .by_group = TRUE) %>%
  mutate(share = amt / sum(amt)) %>%
  summarise(
    HHI   = sum(share^2),
    top5  = sum(share[1:min(5, n())]),
    top10 = sum(share[1:min(10, n())]),
    .groups = "drop"
  ) %>%
  tidyr::pivot_longer(cols = c(top5, top10), names_to = "metric", values_to = "value") %>%
  mutate(metric = recode(metric, top5 = "Top-5 share", top10 = "Top-10 share"))

ggplot(conc_year,
       aes(x = AllocationYear, y = value, colour = metric, group = metric)) +
  geom_line(linewidth = 1.3) +
  geom_point(size = 3.5) +
  geom_text(aes(label = scales::percent(value, accuracy = 1)),
            vjust = -1.1, size = 3.2, fontface = "bold", show.legend = FALSE) +
  scale_colour_manual(values = c("Top-5 share" = "#005f73",
                                 "Top-10 share" = "#94d2bd"),
                      name = NULL) +
  scale_y_continuous(labels = scales::percent_format(),
                     limits = c(0, 0.75),
                     expand = expansion(mult = c(0, 0.05))) +
  scale_x_continuous(breaks = 2022:2025) +
  labs(
    title = "UHF is de-concentrating: top-5 and top-10 partner shares over time",
    subtitle = "Ukraine UHF, 2022–2025 — partner shares calculated per year",
    x = "Allocation year", y = "Share of annual envelope",
    caption = "Source: CBPF ProjectSummary dataset, Feb 2026 extract"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold", size = 14, colour = "#005f73"),
    plot.subtitle = element_text(size = 11, colour = "#555555"),
    legend.position = "top",
    panel.grid.minor = element_blank()
  )
```

::: callout-tip
**The trajectory is toward more partners and smaller top-tier shares**, not fewer. Top-5 share fell from 40% in 2022 to 28% in 2024; top-10 from 62% to 45%. Consistent with the localisation push documented in the 2024 Annual Report (NNGO share rising from 30% in 2023 to \~60% in 2024) and with the UN Agency exit (see below).
:::

## Why UHF Grants Are Bigger: An Arithmetic Identity

Because Ukraine's partner count is similar to Afghanistan or Yemen but its envelope is roughly 2× larger, the **average partner take** mechanically has to be larger. This is not a policy choice — it's arithmetic — and it explains most of the "larger grants" signal in Chapter 5.

```{r grant-distribution-vs-peers}
#| fig-width: 12
#| fig-height: 5

peer_grants <- project_summary %>%
  filter(AllocationYear %in% 2022:2024,
         PooledFundName %in% peer_funds,
         !is.na(Budget), Budget > 0)

ref_lines <- data.frame(
  yintercept = c(5e6, 8e6),
  label      = c("Standard cap $5M", "Exceptional cap $8M"),
  colour     = c("#e63946", "#e76f51")
)

ggplot(peer_grants %>% filter(PooledFundName %in% c("Ukraine","Afghanistan","Yemen","Ethiopia","oPt","Somalia")),
       aes(x = reorder(PooledFundName, Budget, FUN = median), y = Budget)) +
  geom_boxplot(aes(fill = PooledFundName == "Ukraine"),
               outlier.size = 0.9, outlier.alpha = 0.5, width = 0.55) +
  geom_hline(yintercept = 5e6, linetype = "dashed", colour = "#e63946", linewidth = 0.7) +
  annotate("text", x = 0.6, y = 5.3e6, label = "Standard cap $5M",
           colour = "#e63946", size = 3.2, hjust = 0, fontface = "italic") +
  scale_y_log10(labels = scales::label_dollar(scale_cut = scales::cut_short_scale()),
                breaks = c(5e4, 1e5, 5e5, 1e6, 2e6, 5e6, 1e7)) +
  scale_fill_manual(values = c(`TRUE` = "#005f73", `FALSE` = "#94d2bd"),
                    guide = "none") +
  coord_flip() +
  labs(
    title = "Grant-size distribution: Ukraine vs broad peer CBPFs (2022–2024)",
    subtitle = "Log-scale. Ukraine's 90th percentile sits exactly at the $5M standard cap — a ceiling-bound distribution no peer fund exhibits.",
    x = NULL, y = "Grant budget (USD, log scale)",
    caption = "Source: CBPF ProjectSummary dataset, Feb 2026 extract"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold", size = 14, colour = "#005f73"),
    plot.subtitle = element_text(size = 11, colour = "#555555"),
    panel.grid.major.y = element_blank(),
    panel.grid.minor   = element_blank()
  )
```

```{r per-partner-arithmetic}
per_partner <- conc_agg %>%
  filter(Class == "Broad partner fund") %>%
  mutate(
    mean_per_partner = envelope / n_partners,
    Envelope = scales::dollar(envelope / 1e6, suffix = "M", accuracy = 0.1),
    `Mean \\$ per partner (3y)` =
      scales::dollar(mean_per_partner / 1e6, suffix = "M", accuracy = 0.01)
  ) %>%
  arrange(desc(mean_per_partner)) %>%
  select(Fund = PooledFundName, Envelope, Partners = n_partners,
         `Mean \\$ per partner (3y)`)

knitr::kable(per_partner,
             caption = "Mean 3-year envelope per partner in broad-partner CBPFs (2022–2024). UHF's arithmetic average is roughly 2× the next-largest fund.")
```

::: callout-note
**The P90 signature is unique to UHF.** Ukraine is the only CBPF where the 90th percentile grant size sits exactly at the \$5M standard cap — producing a *flat-topped* distribution. Peer funds show long right tails with a few mega-grants pulling the top; Ukraine shows a ceiling. The operational reading is that the UHF routinely writes max-size grants to a large number of partners simultaneously.
:::

## The Core-20: Who Captures the Majority

Filtering to organisations that received grants in **every** UHF allocation round 2022–2024 yields a core of 20 partners. Together they absorbed **55.3% of the \$541M three-year envelope**.

```{r core20-table}
ukr22_24 <- project_summary %>%
  filter(PooledFundName == "Ukraine",
         AllocationYear %in% 2022:2024,
         !is.na(Budget), Budget > 0,
         !is.na(OrganizationName))

ukr25_partner <- project_summary %>%
  filter(PooledFundName == "Ukraine", AllocationYear == 2025,
         !is.na(Budget), Budget > 0, !is.na(OrganizationName)) %>%
  group_by(OrganizationName) %>%
  summarise(n25 = n(), amt25 = sum(Budget, na.rm = TRUE), .groups = "drop")

core20 <- ukr22_24 %>%
  group_by(OrganizationName, OrganizationType) %>%
  summarise(
    n_grants   = n(),
    n_years    = n_distinct(AllocationYear),
    total      = sum(Budget, na.rm = TRUE),
    mean_grant = mean(Budget, na.rm = TRUE),
    max_grant  = max(Budget, na.rm = TRUE),
    .groups    = "drop"
  ) %>%
  filter(n_years == 3) %>%
  left_join(ukr25_partner, by = "OrganizationName") %>%
  arrange(desc(total)) %>%
  mutate(
    Type = case_when(
      OrganizationType == "International NGO" ~ "INGO",
      OrganizationType == "National NGO"      ~ "NNGO",
      OrganizationType == "UN Agency"         ~ "UN",
      TRUE                                    ~ "Other"
    ),
    `Total 22–24`    = scales::dollar(total / 1e6, suffix = "M", accuracy = 0.1),
    `Mean grant`     = scales::dollar(mean_grant / 1e6, suffix = "M", accuracy = 0.01),
    `Max grant`      = scales::dollar(max_grant  / 1e6, suffix = "M", accuracy = 0.01),
    `2025 total`     = ifelse(is.na(amt25), "—",
                              scales::dollar(amt25 / 1e6, suffix = "M", accuracy = 0.01))
  ) %>%
  select(Organisation = OrganizationName, Type,
         Grants = n_grants, `Total 22–24`, `Mean grant`, `Max grant`, `2025 total`)

knitr::kable(core20,
             caption = "Core-20: organisations active in every UHF allocation round 2022–2024, ranked by cumulative disbursement. 2025 column shows whether the partner continued into the current cycle.",
             align  = c("l","l","c","r","r","r","r"))
```

::: callout-important
**Seven partners are "cap-binders"**: ACTED, UNHCR, DRC, People in Need, ICF Caritas Ukraine, NRC and Proliska all average grants of \$3.4M–\$4.8M — every grant at or near the \$5M ceiling. Six of them received at least one **exceptional-cap grant ≥ \$6M** (UNHCR \$9.4M, NRC \$7.5M, ACTED \$6.3M, PIN \$6.1M, Caritas \$6.0M, Proliska \$5.2M). This is where the "repetition at large scale" pattern concentrates — not across the fund, but in a \~7-organisation inner tier.
:::

::: callout-note
**The UN exit is visible at partner level.** UNHCR was the second-largest UHF partner 2022–2024 (\$33.3M across 7 grants) and received \$0 in 2025. CORE Community Organized Relief Effort similarly drops out. Their volume did not leave Ukraine — it was absorbed primarily by DRC (\$17.0M in 2025), ICF Caritas Ukraine (\$14.1M in 2025) and a broadened NNGO tier.
:::

## Dependency Ratio: When UHF Dwarfs the Partner

For the 12 core-20 members with publicly available 2021 revenue, the dependency ratio (UHF 2022–2024 volume / 2021 global organisational revenue) distinguishes absorbable UHF engagement from structural dependency.

```{r dependency-ratio}
# Revenue data compiled from publicly available sources (audited accounts
# and annual reports). Unverifiable cases excluded from this table.
rev_usd <- tibble::tribble(
  ~Organisation,                                     ~rev_2021_usd_m,  ~rev_note,
  "Agency for Technical Cooperation and Development", 463, "€393.7M (2021)",
  "United Nations High Commissioner for Refugees",   5150, "$5.15B (2021)",
  "Danish Refugee Council",                           495, "3.12B DKK (2021)",
  "People in Need",                                   133, "€113M (2021)",
  "Norwegian Refugee Council",                        660, "5.67B NOK (2021)",
  "International Rescue Committee, Inc.",            1160, "$1.16B (2021)",
  "Stichting ZOA",                                     96, "€82M (2021)",
  "Dorcas Aid International Transcarpathia",           28, "€24M parent (2021)",
  "Estonian Refugee Council",                         2.2, "€1.9M (2021)",
  "CORE Community Organized Relief Effort",           122, "$122.1M (2021)",
  "Save the Children Fund",                          2200, "$2.2B (2021)",
  "INTERSOS",                                         126, "€107M (2021)"
)

dep <- ukr22_24 %>%
  group_by(OrganizationName) %>%
  summarise(uhf_total = sum(Budget, na.rm = TRUE), .groups = "drop") %>%
  inner_join(rev_usd, by = c("OrganizationName" = "Organisation")) %>%
  mutate(
    uhf_m    = uhf_total / 1e6,
    ratio    = uhf_m / rev_2021_usd_m,
    Reading  = case_when(
      ratio <  0.1 ~ "Minor line",
      ratio <  0.5 ~ "Meaningful",
      ratio <  2   ~ "Structural dependency",
      TRUE         ~ "UHF dwarfs org"
    ),
    Reading = factor(Reading,
                     levels = c("Minor line","Meaningful",
                                "Structural dependency","UHF dwarfs org"))
  ) %>%
  arrange(desc(ratio))

dep_tbl <- dep %>%
  transmute(
    Organisation   = OrganizationName,
    `2021 revenue` = rev_note,
    `UHF 22–24`    = scales::dollar(uhf_m, suffix = "M", accuracy = 0.01),
    Ratio          = sprintf("%.2f×", ratio),
    Reading
  )

knitr::kable(dep_tbl,
             caption = "Dependency ratio for core-20 members with public 2021 revenue. Ratio = UHF 2022–2024 disbursement divided by the organisation's 2021 global revenue (USD-equivalent).",
             align  = c("l","l","r","r","l"))
```

```{r dependency-plot, fig.width=10, fig.height=5}
#| fig-cap: "Dependency ratio by core-20 partner (log-scaled). The y=1 line marks parity between UHF volume and pre-war annual revenue. Estonian Refugee Council sits more than 4× above parity — UHF volume exceeds its entire pre-war institutional scale."

dep_plot <- dep %>%
  mutate(OrganizationName = forcats::fct_reorder(OrganizationName, ratio))

ggplot(dep_plot,
       aes(x = ratio, y = OrganizationName, colour = Reading)) +
  geom_vline(xintercept = 1, colour = "#e63946",
             linetype = "dashed", linewidth = 0.8) +
  geom_segment(aes(x = 0.001, xend = ratio, yend = OrganizationName),
               linewidth = 0.4, colour = "grey70") +
  geom_point(size = 4) +
  geom_text(aes(label = sprintf("%.2f×", ratio)),
            hjust = -0.25, size = 3.2, fontface = "bold", show.legend = FALSE) +
  scale_x_log10(
    breaks = c(0.01, 0.1, 0.5, 1, 4),
    labels = c("0.01×", "0.1×", "0.5×", "1×", "4×"),
    expand = expansion(mult = c(0.05, 0.35))
  ) +
  scale_colour_manual(values = c(
    "Minor line"             = "#94d2bd",
    "Meaningful"             = "#0a9396",
    "Structural dependency"  = "#ee9b00",
    "UHF dwarfs org"         = "#e63946"
  ), name = NULL) +
  labs(
    title = "UHF dependency ratio: UHF 2022–2024 volume ÷ 2021 global revenue",
    subtitle = "Log scale. Red dashed line = UHF equal to entire pre-war annual revenue.",
    x = "Dependency ratio (log scale)", y = NULL,
    caption = "Source: organisation audited accounts and annual reports, 2021. UHF volume from CBPF ProjectSummary, Feb 2026."
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title    = element_text(face = "bold", size = 13, colour = "#005f73"),
    plot.subtitle = element_text(size = 10, colour = "#555555"),
    legend.position = "top",
    panel.grid.minor = element_blank()
  )
```

::: callout-important
**Estonian Refugee Council** is the sharpest visible case: a domestic Estonian NGO with €1.9M 2021 revenue absorbing \$9.4M of UHF funding over 2022–2024 — **4.3× its pre-war annual scale**. Whatever governance, finance and compliance infrastructure existed in 2021 was sized for a €1.9M/year operation. UHF is not supplementing this organisation; UHF *is* its operational scale. **Dorcas Aid International Transcarpathia** is a parallel case at 0.38× against the €24M global parent, though the country-unit share is likely higher.
:::

::: callout-note
**Transparency gap.** Seven of the eight core-20 Ukrainian NNGOs (ICF Caritas Ukraine, Proliska, NEW WAY, POSMISHKA UA, Unity for the Future, HADC, and others appearing in adjacent cohorts) have no publicly available 2021 revenue figure. An independent capacity-to-volume assessment is therefore blocked on the denominator side. This is not a criticism — it reflects a legitimate gap in the public-disclosure practice of domestic NGOs operating in a war economy — but it should be flagged in any external review of UHF's fiduciary exposure.
:::

------------------------------------------------------------------------

# The Ideal UHF Candidate Profile

The preceding chapters characterised the UHF system-wide. This chapter flips the lens: given what we know about how the UHF allocates money, **what does an ideal partner look like?** Rather than inferring intent from eligibility documents, we read the profile off the disbursement history itself — i.e. the organisations that the UHF has repeatedly trusted with large, complex programmes. Two dimensions define the profile:

1.  **Geographic coverage** — the number of distinct oblasts an organisation operates in. A broad footprint signals the ability to deliver across the contact line and in liberated/recovery regions alike.
2.  **Sector coverage** — the number of distinct clusters an organisation engages in. Multi-sector capacity signals the ability to bundle protection, shelter, WASH, health or food security into one integrated programme.

Together these two axes define a simple 2×2 that partitions the partner universe into **Specialists**, **Area generalists**, **Sector generalists**, and **Full-coverage partners**. The ideal candidate sits in the upper-right quadrant.

```{r candidate-prep}
#| include: false

# --- Build organisation-level coverage table for Ukraine, 2019-2025 ---
ukr_ps <- project_summary %>%
  filter(PooledFundName == "Ukraine",
         AllocationYear >= 2019,
         !is.na(Budget), Budget > 0,
         !is.na(OrganizationName))

ukr_loc <- project_summary_location %>%
  filter(PooledFundName == "Ukraine",
         AllocationYear >= 2019,
         !is.na(AdminLocation1),
         !is.na(OrganizationName))

# Cluster file has no OrganizationName — join via ChfProjectCode (globally unique)
ukr_cl <- clusters %>%
  filter(PooledFundName == "Ukraine", AllocationYear >= 2019) %>%
  left_join(ukr_ps %>% select(ChfProjectCode, OrganizationName, OrganizationType),
            by = "ChfProjectCode") %>%
  filter(!is.na(OrganizationName))

geo_cov <- ukr_loc %>%
  group_by(OrganizationName) %>%
  summarise(n_oblasts = n_distinct(AdminLocation1), .groups = "drop")

sec_cov <- ukr_cl %>%
  group_by(OrganizationName) %>%
  summarise(n_clusters = n_distinct(Cluster), .groups = "drop")

org_cov <- ukr_ps %>%
  group_by(OrganizationName, OrganizationType) %>%
  summarise(
    n_projects    = n(),
    n_years       = n_distinct(AllocationYear),
    total_budget  = sum(Budget, na.rm = TRUE),
    median_grant  = median(Budget, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  left_join(geo_cov, by = "OrganizationName") %>%
  left_join(sec_cov, by = "OrganizationName") %>%
  mutate(
    n_oblasts  = tidyr::replace_na(n_oblasts, 0),
    n_clusters = tidyr::replace_na(n_clusters, 0),
    OrgType = case_when(
      OrganizationType == "International NGO" ~ "INGO",
      OrganizationType == "National NGO"      ~ "NNGO",
      OrganizationType == "UN Agency"         ~ "UN Agency",
      TRUE                                    ~ "Other"
    )
  )

# Quadrant thresholds — use the 75th percentile on each axis
q_oblasts  <- quantile(org_cov$n_oblasts,  0.75, na.rm = TRUE)
q_clusters <- quantile(org_cov$n_clusters, 0.75, na.rm = TRUE)

org_cov <- org_cov %>%
  mutate(
    Quadrant = case_when(
      n_oblasts >= q_oblasts  & n_clusters >= q_clusters ~ "Full-coverage",
      n_oblasts >= q_oblasts  & n_clusters <  q_clusters ~ "Area generalist",
      n_oblasts <  q_oblasts  & n_clusters >= q_clusters ~ "Sector generalist",
      TRUE                                                ~ "Specialist"
    )
  )

# Headline stats for callout
n_total    <- nrow(org_cov)
n_full     <- sum(org_cov$Quadrant == "Full-coverage")
pct_full   <- round(100 * n_full / n_total)
bud_full   <- sum(org_cov$total_budget[org_cov$Quadrant == "Full-coverage"])
pct_bud_full <- round(100 * bud_full / sum(org_cov$total_budget))
med_bud_full <- median(org_cov$total_budget[org_cov$Quadrant == "Full-coverage"]) / 1e6
med_bud_spec <- median(org_cov$total_budget[org_cov$Quadrant == "Specialist"])   / 1e6
```

## Mapping the Partner Universe

The scatter plot below places every Ukraine UHF partner (2019–2025) on the two coverage axes. Bubble size is proportional to cumulative disbursed budget; colour reflects organisation type. Dashed lines mark the 75th-percentile thresholds that split the plane into four quadrants.

```{r candidate-scatter, fig.width=10, fig.height=7}
#| fig-cap: "UHF partner universe, 2019–2025. Oblast coverage vs cluster coverage; bubble size = total disbursed budget. Dashed lines mark the 75th-percentile thresholds. Full-coverage quadrant (upper-right) concentrates the largest, most trusted partners."

library(ggrepel)

# Label the biggest bubbles + interesting outliers
label_orgs <- org_cov %>%
  mutate(rank_bud = rank(-total_budget)) %>%
  filter(rank_bud <= 10 | (Quadrant == "Full-coverage" & rank_bud <= 20)) %>%
  mutate(lab = stringr::str_wrap(OrganizationName, 26))

pal_orgtype <- c("INGO" = "#0A9396", "NNGO" = "#EE9B00", "UN Agency" = "#005f73", "Other" = "#94d2bd")

ggplot(org_cov, aes(x = n_oblasts, y = n_clusters)) +
  # quadrant shading
  annotate("rect", xmin = q_oblasts, xmax = Inf, ymin = q_clusters, ymax = Inf,
           fill = "#0a9396", alpha = 0.07) +
  geom_hline(yintercept = q_clusters, linetype = "dashed", colour = "grey55") +
  geom_vline(xintercept = q_oblasts,  linetype = "dashed", colour = "grey55") +
  geom_point(aes(size = total_budget / 1e6, colour = OrgType),
             alpha = 0.75, stroke = 0.3) +
  geom_text_repel(
    data = label_orgs,
    aes(label = lab, colour = OrgType),
    size = 3.1, lineheight = 0.9,
    max.overlaps = Inf, seed = 42,
    box.padding = 0.35, point.padding = 0.2,
    segment.size = 0.3, segment.colour = "grey60",
    show.legend = FALSE
  ) +
  annotate("text", x = max(org_cov$n_oblasts) * 0.98, y = max(org_cov$n_clusters) * 1.02,
           label = "Full-coverage", hjust = 1, vjust = 1,
           fontface = "bold", colour = "#0a9396", size = 4.2) +
  annotate("text", x = q_oblasts * 0.5, y = q_clusters * 0.5,
           label = "Specialists", hjust = 0.5, vjust = 0.5,
           colour = "grey55", size = 3.6, fontface = "italic") +
  scale_size_area(max_size = 14, breaks = c(5, 15, 30, 50),
                  name = "Total disbursed\n(US$ millions)") +
  scale_colour_manual(values = pal_orgtype, name = "Organisation type") +
  scale_x_continuous(breaks = seq(0, 25, 5)) +
  scale_y_continuous(breaks = seq(0, 14, 2)) +
  labs(
    title = "Who fits the UHF profile? Coverage map of Ukraine partners, 2019–2025",
    subtitle = paste0("Upper-right quadrant = ", n_full, " partners (", pct_full,
                      "% of orgs) capturing ", pct_bud_full, "% of disbursed budget"),
    x = "Geographic coverage — distinct oblasts",
    y = "Sector coverage — distinct clusters",
    caption = "Thresholds: 75th percentile on each axis. Source: CBPF ProjectSummary + Location + Cluster, Feb 2026 extract."
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title     = element_text(face = "bold", size = 14, colour = "#005f73"),
    plot.subtitle  = element_text(size = 11, colour = "#555555"),
    panel.grid.minor = element_blank(),
    legend.position  = "right",
    legend.box       = "vertical"
  ) +
  guides(
    colour = guide_legend(order = 1, override.aes = list(size = 5)),
    size   = guide_legend(order = 2)
  )
```

::: callout-note
**`r n_full` of `r n_total` partners (`r pct_full`%) sit in the Full-coverage quadrant**, but together they captured **`r pct_bud_full`% of all UHF disbursements since 2019**. Their median cumulative budget is **US\$`r round(med_bud_full,1)`M**, versus **US\$`r round(med_bud_spec,2)`M** for Specialists — an order-of-magnitude gap that confirms coverage is the single strongest predictor of scale at the UHF.
:::

## Quadrant Breakdown by Organisation Type

Does the ideal-candidate profile favour a specific organisation type? The table below decomposes each quadrant by INGO / NNGO / UN Agency, showing both the partner count and the share of total budget that flowed to each cell.

```{r candidate-quadrant-tbl}
q_tbl <- org_cov %>%
  group_by(Quadrant, OrgType) %>%
  summarise(
    n_orgs       = n(),
    total_budget = sum(total_budget, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    pct_bud = 100 * total_budget / sum(total_budget),
    cell = sprintf("%d orgs — US$%.1fM (%.1f%%)",
                   n_orgs, total_budget / 1e6, pct_bud)
  ) %>%
  select(Quadrant, OrgType, cell) %>%
  tidyr::pivot_wider(names_from = OrgType, values_from = cell, values_fill = "—") %>%
  mutate(Quadrant = factor(Quadrant,
                           levels = c("Full-coverage", "Sector generalist",
                                      "Area generalist", "Specialist"))) %>%
  arrange(Quadrant)

knitr::kable(
  q_tbl,
  caption = "Partner count and share of total disbursed budget by quadrant and organisation type (Ukraine UHF, 2019–2025)",
  col.names = c("Quadrant", "INGO", "NNGO", "UN Agency")
)
```

::: callout-tip
**Three readings emerge from the quadrant table:**

-   The **Full-coverage quadrant is mixed** — it is not exclusively INGO territory. NNGOs and UN agencies both appear in meaningful numbers, confirming that broad footprint is achievable across organisation types. It is important to point that, in the case of Multipurpose Cash Asssistance (MPCA), UN agencies relied heavily on online registration, which means that no physical presence was actually achieved.
-   **UN agencies cluster as Area generalists**: they cover many oblasts but relatively few clusters (they tend to specialise by mandate — UNHCR on protection/shelter, IOM on mobility, FAO on food security).
-   **NNGOs split in two**: a small cohort joins the Full-coverage quadrant (e.g. ICF Caritas Ukraine, Proliska, Charity Fund "New Way"), while most remain Specialists with 1–2 oblasts and 1–3 clusters — a realistic starting point for first-time UHF applicants.
:::

## The Full-Coverage Cohort: Named List

Finally, the organisations that currently meet the ideal profile. This is the disbursement-based short-list: partners the UHF has already validated at scale on both coverage axes.

```{r candidate-topn}
top_cov <- org_cov %>%
  filter(Quadrant == "Full-coverage") %>%
  arrange(desc(total_budget)) %>%
  mutate(
    `Total disbursed` = scales::dollar(total_budget, scale = 1e-6, suffix = "M", accuracy = 0.1),
    `Median grant`    = scales::dollar(median_grant, scale = 1e-6, suffix = "M", accuracy = 0.1)
  ) %>%
  select(
    Organisation  = OrganizationName,
    Type          = OrgType,
    Oblasts       = n_oblasts,
    Clusters      = n_clusters,
    `Years active` = n_years,
    Projects       = n_projects,
    `Total disbursed`,
    `Median grant`
  )

knitr::kable(
  top_cov,
  caption = paste0("Full-coverage cohort — ", nrow(top_cov),
                   " organisations in the upper-right quadrant (Ukraine UHF, 2019–2025), ranked by total disbursed budget"),
  align = c("l", "l", "c", "c", "c", "c", "r", "r")
)
```

::: callout-important
**Implications for partner development.** If the UHF wishes to expand this cohort — particularly by growing the NNGO pipeline — investment should target the *adjacent* quadrants: Area generalists (broad footprint, narrow sector) need capacity-building to add clusters; Sector generalists (multi-sector, narrow footprint) need support to expand geographically, often via consortium or sub-granting arrangements. Specialists are where the NNGO on-ramp begins.
:::

------------------------------------------------------------------------

# Health Funding Recipients in Ukraine (2019–2025)

Since the Ukraine Humanitarian Fund was established in 2019, the Health cluster has received a significant share of total allocations. This section identifies every organisation that has received health-earmarked funding and tracks how the recipient landscape has evolved over time.

```{r health-recipients-prep}
#| include: false

# Join cluster-level health budgets to project_summary for org metadata.
# Use ChfProjectCode (globally unique) rather than ChfId (which has duplicates
# across funds) to avoid spurious cross-fund joins.
ukr_health_raw <- clusters %>%
  filter(
    PooledFundName == "Ukraine",
    Cluster        == "Health",
    AllocationYear != 2026
  ) %>%
  left_join(
    project_summary %>%
      select(ChfProjectCode, OrganizationName, OrganizationType) %>%
      distinct(),
    by = "ChfProjectCode"
  ) %>%
  filter(!is.na(OrganizationName))

# Aggregate by organisation
ukr_health_orgs <- ukr_health_raw %>%
  group_by(OrganizationName, OrganizationType) %>%
  summarise(
    TotalHealth_M = sum(ClusterBudget, na.rm = TRUE) / 1e6,
    n_grants      = n_distinct(ChfProjectCode),
    first_year    = min(AllocationYear),
    last_year     = max(AllocationYear),
    .groups       = "drop"
  ) %>%
  arrange(desc(TotalHealth_M)) %>%
  mutate(
    OrgType = case_when(
      OrganizationType == "UN Agency"         ~ "UN Agency",
      OrganizationType == "International NGO" ~ "INGO",
      OrganizationType == "National NGO"      ~ "NNGO",
      TRUE                                    ~ OrganizationType
    ),
    OrgType = factor(OrgType, levels = c("UN Agency", "INGO", "NNGO"))
  )

# Yearly breakdown per org for the heatmap
ukr_health_yearly <- ukr_health_raw %>%
  group_by(OrganizationName, AllocationYear) %>%
  summarise(
    Budget_M = sum(ClusterBudget, na.rm = TRUE) / 1e6,
    .groups  = "drop"
  )

# Colour by org type
orgtype_colors <- c("UN Agency" = "#005f73", "INGO" = "#2B638B", "NNGO" = "#e9c46a")
```

## All-time Health Funding by Recipient

```{r health-recipients-lollipop}
#| fig-width: 13
#| fig-height: 9

top_n_show <- nrow(ukr_health_orgs)   # show all recipients

plot_df <- ukr_health_orgs %>%
  slice_head(n = top_n_show) %>%
  mutate(
    OrganizationName = stringr::str_trunc(OrganizationName, width = 45),
    OrganizationName = factor(OrganizationName,
                              levels = rev(unique(OrganizationName)))
  )

ggplot(plot_df,
       aes(x = OrganizationName, y = TotalHealth_M, colour = OrgType)) +
  geom_segment(aes(xend = OrganizationName, y = 0, yend = TotalHealth_M),
               linewidth = 0.8, alpha = 0.6) +
  geom_point(aes(size = n_grants), alpha = 0.9) +
  geom_text(aes(label = paste0("$", round(TotalHealth_M, 1), "M")),
            hjust = -0.2, size = 3.0, fontface = "bold") +
  coord_flip() +
  scale_colour_manual(values = orgtype_colors, name = "Org. type") +
  scale_size_continuous(name = "No. of grants", range = c(3, 9), breaks = c(1, 3, 5, 8)) +
  scale_y_continuous(
    labels = scales::dollar_format(suffix = "M", accuracy = 0.1),
    expand = expansion(mult = c(0, 0.28))
  ) +
  labs(
    title    = "Ukraine UHF: Health Funding Recipients (2019–2025)",
    subtitle = "All organisations that received Health-cluster funding since the fund was established.\nDot size = number of grants received.",
    x        = NULL,
    y        = "Total health-earmarked funding (USD millions)",
    caption  = "Source: CBPF Cluster + ProjectSummary datasets, Feb 2026 extract"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title         = element_text(face = "bold", size = 14, colour = "#005f73"),
    plot.subtitle      = element_text(size = 11, colour = "#555555"),
    axis.text.y        = element_text(size = 9.5),
    legend.position    = "right",
    panel.grid.major.y = element_blank(),
    panel.grid.minor   = element_blank()
  )
```

```{r health-recipients-callout-vals, include=FALSE}
total_health_m   <- sum(ukr_health_orgs$TotalHealth_M)
n_orgs           <- nrow(ukr_health_orgs)
top1             <- ukr_health_orgs$OrganizationName[1]
top1_share       <- round(ukr_health_orgs$TotalHealth_M[1] / total_health_m * 100, 1)
top3_share       <- round(sum(ukr_health_orgs$TotalHealth_M[1:3]) / total_health_m * 100, 1)
n_un             <- sum(ukr_health_orgs$OrgType == "UN Agency")
n_ingo           <- sum(ukr_health_orgs$OrgType == "INGO")
n_nngo           <- sum(ukr_health_orgs$OrgType == "NNGO")
```

::: callout-note
**`r n_orgs` organisations** have received Health-cluster funding from the Ukraine Humanitarian Fund since 2019, together accounting for **\$`r round(total_health_m, 1)`M** in health-earmarked allocations. The top recipient — **`r top1`** — captured **`r top1_share`%** of all health funding, and the top three organisations together account for **`r top3_share`%**. The recipient pool comprises `r n_un` UN agenc`r ifelse(n_un==1,"y","ies")`, `r n_ingo` INGOs and `r n_nngo` NNGOs, reflecting the mixed multi-channel delivery model used in the Ukraine response.
:::

## Health Funding by Organisation Over Time

```{r health-recipients-heatmap}
#| fig-width: 13
#| fig-height: 8

# Re-order orgs by total funding for consistent axis
org_order_health <- ukr_health_orgs %>%
  arrange(TotalHealth_M) %>%
  pull(OrganizationName)

heatmap_df <- ukr_health_yearly %>%
  mutate(
    OrganizationName = stringr::str_trunc(OrganizationName, width = 45),
    OrganizationName = factor(OrganizationName,
                              levels = stringr::str_trunc(org_order_health, width = 45))
  ) %>%
  complete(OrganizationName, AllocationYear = 2019:2025,
           fill = list(Budget_M = NA))

ggplot(heatmap_df,
       aes(x = as.factor(AllocationYear),
           y = OrganizationName,
           fill = Budget_M)) +
  geom_tile(colour = "white", linewidth = 0.4) +
  geom_text(aes(label = ifelse(!is.na(Budget_M),
                               paste0("$", round(Budget_M, 1), "M"),
                               "")),
            size = 2.8, colour = "white", fontface = "bold") +
  scale_fill_gradient(
    low     = "#94d2bd",
    high    = "#005f73",
    na.value = "grey93",
    name    = "Health budget\n(USD millions)",
    labels  = scales::dollar_format(suffix = "M", accuracy = 0.1)
  ) +
  labs(
    title    = "Ukraine UHF: Health Funding Per Recipient and Year (2019–2025)",
    subtitle = "Grey cells = no health-cluster grant in that year",
    x        = "Allocation Year",
    y        = NULL,
    caption  = "Source: CBPF Cluster + ProjectSummary datasets, Feb 2026 extract"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title       = element_text(face = "bold", size = 14, colour = "#005f73"),
    plot.subtitle    = element_text(size = 11, colour = "#555555"),
    axis.text.y      = element_text(size = 9),
    axis.text.x      = element_text(size = 10, face = "bold"),
    panel.grid       = element_blank(),
    legend.position  = "right"
  )
```

The heatmap makes the temporal fragmentation of health delivery visible: organisations that carried health programming before the full-scale invasion are largely replaced post-2022 by a different set of partners with the absorptive capacity to deploy larger grants. Notably, WHO — historically the anchor UN agency for health coordination — drops out of the UHF partner list after 2023, a shift that coincides with the United States' announced withdrawal from the World Health Organization. This matters for continuity: UHF health funding is now concentrated in a handful of INGOs and a growing NNGO cohort, with no resident UN health convenor inside the fund. Monitoring whether this reconfiguration preserves technical standards, coordination, and last-mile reach will be an important diagnostic for the fund's health portfolio going forward.

It is clear that the UHF's health funding strategy is based on absorcipon rather than field experience or technical expertise. The top health recipients are not specialised health actors but rather the largest, most trusted partners with the capacity to deploy large grants at speed. Moreover, the technical miscalculation of not considering Local Health facilities as partnership candidates makes this funding impact even more questioable: https://baena.blog/00.-STOCK/The-Mistake-of-Excluding-Local-Health-Facilities-from-Localization-Targets-in-Ukraine

Furthermore, assessing the true cost-effectiveness of concentrating funding among a few 'official' partners remains impossible, given that 5W (Who, What, Where, When, and for Whom) operational data is frequently partial or unpublished. Despite these data constraints, a distinct profile emerges for the preferred health partners: they are predominantly Anglo-Saxon organizations—including some headquartered in non-WHO member states—that possessed no operational footprint in Ukraine prior to the 2022 escalation.

------------------------------------------------------------------------

# Cross-Check: UHF Annual Reports vs. Data

A dataset-based analysis is only useful if it agrees with — or carefully disagrees with — the fund's own published narrative. This chapter reconciles the figures in this report against the UHF Annual Reports for 2022, 2023 and 2024. The exercise serves three purposes: (i) validate the data pipeline; (ii) surface reporting inconsistencies the reader should know about; (iii) highlight structural shifts the ARs document that are easy to miss in a year-by-year view.

## Reconciliation Status

```{r ar-reconciliation}
ar_rec <- tibble::tribble(
  ~Year, ~Metric,                  ~`Raw data`,                                 ~`AR figure`,                                     ~Verdict,
  "2022", "Total allocations",     "$192.7M / 56 partners / 110 projects",       "$192M / 56 partners / 109 projects",             "Match (1-project delta)",
  "2022", "Org-type split",         "INGO $82M · NNGO $45M · UN $66M",           "INGO $82M · UN $66M · NNGO $44M (direct)",        "Match",
  "2023", "Total allocations",     "$183.7M / 49 partners / 74 projects",        "$181.2M / 49 partners",                           "Close (+$2.5M)",
  "2023", "Org-type split",         "INGO $104.0M · NNGO $38.3M · UN $41.4M",    "INGO $92M · NNGO $56.8M · UN $32.4M",             "Mismatch — see below",
  "2024", "Total allocations",     "$164.3M / 58 partners / 73 projects",        "$164.3M / 58 partners / 73 projects",             "Exact match",
  "2024", "Org-type split",         "INGO $84.6M · NNGO $74.8M · UN $5.0M",      "INGO $84.6M (51.9%) · NNGO $73.3M (45%) · UN $5M", "Match"
)

knitr::kable(ar_rec,
             caption = "Reconciliation: raw CBPF extract (Feb 2026) against UHF Annual Report headline figures.",
             align  = c("l","l","l","l","l"))
```

::: callout-important
**The 2023 org-type breakdown does not reconcile.** The 2023 AR reports \$56.8M to NNGOs (≈30% of the envelope); the raw CBPF extract shows \$38.3M direct to National NGO entities. The most plausible reconciliation is that the AR definition includes sub-granted funding flowing from INGO prime recipients to Ukrainian NNGO sub-implementers, and possibly bundles Red Cross / partially-national entities into the NNGO category. When quoting either figure in discussion with external audiences, it is advisable to cite the definition used (direct-partner classification vs end-beneficiary-entity classification).
:::

## Structural Shifts the ARs Document

Three cross-year trends in the ARs are not obvious from any single year of data and deserve explicit highlighting.

### 1. The UN Agency exit

```{r un-exit}
#| fig-width: 10
#| fig-height: 4

un_exit <- project_summary %>%
  filter(PooledFundName == "Ukraine",
         AllocationYear %in% 2022:2025,
         !is.na(Budget), Budget > 0) %>%
  mutate(OT = case_when(
    OrganizationType == "International NGO" ~ "INGO",
    OrganizationType == "National NGO"      ~ "NNGO",
    OrganizationType == "UN Agency"         ~ "UN Agency",
    TRUE                                    ~ "Other"
  )) %>%
  group_by(AllocationYear, OT) %>%
  summarise(amt_m = sum(Budget) / 1e6, .groups = "drop")

ggplot(un_exit, aes(x = AllocationYear, y = amt_m, colour = OT, group = OT)) +
  geom_line(linewidth = 1.3) +
  geom_point(size = 3.5) +
  geom_text(aes(label = scales::dollar(amt_m, suffix = "M", accuracy = 1)),
            vjust = -1.0, size = 3.1, fontface = "bold", show.legend = FALSE) +
  scale_colour_manual(values = c(
    "INGO"      = "#005f73",
    "NNGO"      = "#2a9d8f",
    "UN Agency" = "#e63946",
    "Other"     = "#adb5bd"
  ), name = NULL) +
  scale_x_continuous(breaks = 2022:2025) +
  scale_y_continuous(labels = scales::dollar_format(suffix = "M"),
                     expand = expansion(mult = c(0.05, 0.15))) +
  labs(
    title = "UN Agency allocations collapsed from $66M to $5M in two years",
    subtitle = "Ukraine UHF, 2022–2025, USD millions by organisation type",
    x = "Allocation year", y = "Disbursed (USD millions)",
    caption = "Source: CBPF ProjectSummary dataset, Feb 2026 extract. 2024 AR: UN would no longer be funded unless absolutely necessary."
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold", size = 14, colour = "#005f73"),
    plot.subtitle = element_text(size = 11, colour = "#555555"),
    legend.position = "top",
    panel.grid.minor = element_blank()
  )
```

The 2024 Annual Report is explicit: *"UN would no longer be funded unless absolutely necessary."* The data tracks this policy: UN Agency allocations fell from **\$66.0M (2022) → \$41.4M (2023) → \$5.0M (2024)** — a 92% drop in two years. Displaced volume flowed primarily to DRC, ICF Caritas Ukraine and a broadened NNGO tier (see Concentration chapter).

### 2. UHF's growing weight in the CBPF system

Per the Annual Reports:

-   **2023 AR**: UHF = **17%** of global CBPF allocations, 7% of the Ukraine HRP.
-   **2024 AR**: UHF = **20%** of global CBPF allocations, **\>10%** of the Ukraine HNRP — *"largest CBPF for the third consecutive year."*

One in every five CBPF dollars globally now flows through UHF. The "7th of 28" framing used in parts of this report for reserve-allocation share is accurate for that narrow metric, but understates UHF's overall systemic weight.

### 3. Contributions volatility with rising localisation

```{r ar-contribution-table}
ar_tbl <- tibble::tribble(
  ~Year, ~`Contributions (AR)`, ~`Donors (AR)`, ~`Allocations (AR)`, ~`NNGO direct (AR)`, ~`WLO funding (AR)`,
  "2022", "$327M",             "28",           "$192M",             "33%",                "n/r",
  "2023", "$182M (−44% YoY)",  "26",           "$181M",             "30%",                "$9M (~5%)",
  "2024", "$232M",             "27",           "$164M",             "60% ($96M)",         "$50.4M (31%)"
)

knitr::kable(ar_tbl,
             caption = "Headline year-on-year shifts reported in the UHF Annual Reports 2022, 2023 and 2024. WLO = Women-Led Organisation funding.",
             align  = c("l","r","c","r","r","r"))
```

The 2023 contributions drop (−44% YoY) is larger than the allocations line suggests because UHF ran down prior-year cash balances. Despite partial 2024 recovery, UHF's contribution base remains **29% below its 2022 peak** — the fund's *relative* systemic importance is rising on a *shrinking absolute* base. Meanwhile WLO-earmarked funding jumped from ≈5% to 31% of the envelope in a single year, the most dramatic single-year shift in any policy category documented in the ARs.

## Allocation Cadence

All three ARs document a consistent cadence: a large early-year Reserve or Standard allocation, a mid-year Reserve, and a late-year top-up. For partners, this is the operational rhythm UHF expects them to absorb.

```{r alloc-cadence}
cadence <- tibble::tribble(
  ~Year, ~`Round 1`,                ~`Round 2`,                ~`Round 3`,
  "2022", "RA Feb–Apr: $91M total",  "RA Jul: $26M",             "SA Sep: $75M (largest-ever at the time)",
  "2023", "RA1 Jan: $52M",           "SA1 Mar: $69M",            "RA2 Aug: $60M",
  "2024", "SA1 Jan: $76.2M",         "RA1 Jul: $66.6M",          "RA2 Oct: $21.5M",
  "2025", "SA1 Jan: $86.8M (new record)", "—",                    "—"
)

knitr::kable(cadence,
             caption = "UHF allocation rounds by year, as reported in the Annual Reports. 2025 reflects the January Standard Allocation documented in the 2024 AR.",
             align  = c("l","l","l","l"))
```

::: callout-tip
**Implication for candidate partners.** An organisation able to sit in the UHF's core-20 needs to be able to respond to two Reserve triggers per calendar year and absorb a Standard allocation of \$70M–\$90M distributed across 30–50 partners. The cap-binding pattern documented in the Concentration chapter is the partner-level expression of this cadence.
:::

# Conclusions

The Ukraine Humanitarian Pooled Fund (UHF) represents a unique operational model within the global CBPF system. Driven by an unprecedented financial envelope and an aggressive localization mandate, the fund has redefined how humanitarian financing is distributed in a high-intensity, sustained conflict.

Several defining characteristics set the UHF apart from its global peers:

**Scale Over Concentration:** While the UHF appears highly concentrated, metric analysis (HHI, top-5 share) proves it is a broad-partnership fund. The illusion of concentration is an arithmetic byproduct of its massive envelope interacting with the standard \$5M grant ceiling. Consequently, the UHF routinely issues max-sized grants to a wide array of partners, creating a flat-topped distribution unseen in any other country fund.

**True Localization:** Globally, International NGOs consistently receive grants 25% larger than National NGOs. In Ukraine, this paradigm has been inverted. Following the full-scale invasion—and accelerated by a deliberate reduction in UN Agency funding (falling from \$66M in 2022 to \$5M in 2024)—NNGOs now receive larger median grants than INGOs.

**The Core-20 and Fiduciary Reality:** The fund relies heavily on a stable core of 20 organizations capable of absorbing large-scale, repeated funding rounds. For several NNGOs, this has resulted in UHF volumes that dwarf their entire pre-war annual revenues. While this highlights successful capacity building, it also establishes deep structural dependencies that will require careful fiduciary and transitional management.

**The "Full-Coverage" Premium:** The data reveals a clear "ideal candidate" profile. Organizations that operate across multiple oblasts and sectors (Full-coverage partners) capture the vast majority of the disbursed budget. For NNGOs looking to scale, the empirical path is to evolve from geographic or sector specialists into area or sector generalists via consortiums or geographic expansion.

**Health Sector Fragmentation:** Despite large individual grant sizes, the health sector's implementing landscape has experienced significant churn. The exit of historical anchor agencies like the WHO from the direct recipient pool leaves the health portfolio increasingly reliant on a shifting mix of INGOs and NNGOs, raising questions about long-term technical continuity and coordination.

**Maturing Allocation Strategies:** The UHF's historical over-reliance on Reserve Allocations (51.9% globally compared to a 40.2% baseline) is beginning to self-correct. This reflects a maturation of the response: acknowledging that while humanitarian needs in Ukraine remain incredibly high, the foundational operational environment is no longer defined by the extreme volatility of 2022.

The positive: UHF has proven that it is possible to localize a massive humanitarian response without defaulting to a handful of UN mega-grants. However, sustaining this model requires ongoing vigilance regarding partner dependency, sector-specific technical leadership, and the careful balancing of emergency agility with predictable, standard funding cycles.

Despite its scale, critical transparency gaps remain within the fund. The blurred distinction between standard and reserve allocations, the absence of published pipeline data, and limited visibility into partner underspending and audit results introduce structural biases into partner selection and retention. Ultimately, this opacity risks fostering a closed ecosystem of entrenched, de facto 'official' partners rather than maintaining a truly competitive and evidence-based allocation process.

------------------------------------------------------------------------

*Report generated on `r Sys.Date()` using Quarto and the assistance of Gemini 3 Pro, Claude 4.6 and Ollama models.*

## Bibliography and data sources

-   CBPF ProjectSummary, Cluster and Location datasets (Feb 2026 extract, OCHA data explorer)
-   UHF 2022 Annual Report (OCHA / UHF)
-   UHF 2023 Annual Report (OCHA / UHF)
-   UHF 2024 Annual Report (OCHA / UHF)
-   CBPF Global Guidelines and Eligibility Criteria (OCHA)
-   Audited accounts and published annual reports of individual implementing organisations, FY 2021 (for dependency-ratio denominators)