---
title: "Ukraine Humanitarian Pooled Fund Analysis"
subtitle: "Funding Allocation and Impact Assessment"
author: "Ukraine Poolfund Analysis Team"
date: "2026-02-13"
format:
  html:
    theme: yeti
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-summary: "Show code"
    embed-resources: true
    smooth-scroll: true
    number-sections: true
    css: custom-style.css
editor: visual
execute:
  warning: false
  message: false
  echo: true
  fig-align: center
  fig-width: 8
  fig-height: 5
  fig-dpi: 150
---

# Executive Summary

This report provides a comprehensive analysis of the Ukraine Humanitarian Pooled Fund, examining funding allocations and project implementation from 2019 to 2025. These findings are contextualized within the global CBPF landscape, leveraging historical data from 28 country pooled funds spanning 2010 to 2025 to provide comparative insights and identify emerging trends in humanitarian financing.

## Key Findings

-   **Total Projects Analyzed**: Analysis spans multiple allocation years
-   **Primary Focus**: 2025 funding distribution and organizational performance
-   **Data Sources**: Official UHF allocation and project data

# Data Overview

```{r setup}
#| label: setup
#| include: false

# Load required libraries
library(dplyr)
library(ggplot2)
library(tidyr)
library(readr)
library(lubridate)
library(stringr)
library(knitr)
library(scales)

# Set working directory
setwd("/home/jbi-debian/Git/2026-ukraine-poolfund-analysis")

# --- Consistent ggplot2 Theme ---
theme_set(
  theme_minimal(base_size = 14, base_family = "") +
  theme(
    plot.title = element_text(face = "bold", color = "#005f73", hjust = 0.5, size = 16),
    plot.subtitle = element_text(color = "#343a40", hjust = 0.5, margin = margin(b=10)),
    plot.caption = element_text(color = "grey50", face = "italic", size = 9),
    axis.title = element_text(face = "bold", size = 12),
    axis.text = element_text(color = "grey30", size = 10),
    legend.position = "top",
    legend.title = element_text(face = "bold", size = 11),
    panel.grid.minor = element_blank()
  )
)

# Color palette matching the theme
theme_colors <- list(
  primary = "#005f73",
  secondary = "#2B638B",
  accent = "#c7254e",
  light_bg = "#eef7f7"
)
```

```{r load-data}
#| label: load-data

# Load all data files
allocation_flow <- read_csv("Raw-data/AllocationFlow__20260213_094810_UTC.csv", 
                            show_col_types = FALSE)
project_summary <- read_csv("Raw-data/ProjectSummary20260213095023733.csv", 
                            show_col_types = FALSE)
project_summary_location <- read_csv("Raw-data/ProjectSummaryWithLocation20260213095148892.csv", 
                                     show_col_types = FALSE)
project_summary_cluster <- read_csv("Raw-data/ProjectSummaryWithLocationAndCluster20260213095151165.csv", 
                                    show_col_types = FALSE)
pipeline_summary <- read_csv("Raw-data/PipelineProjectSummary20260213095153268.csv", 
                             show_col_types = FALSE)
pipeline_cluster <- read_csv("Raw-data/PipelineProjectCluster20260213095156107.csv", 
                             show_col_types = FALSE)
clusters <- read_csv("Raw-data/Cluster20260213095154538.csv", 
                     show_col_types = FALSE)
contributions <- read_csv("Raw-data/Contribution20260213095157629.csv", 
                          show_col_types = FALSE)

# Data preprocessing
project_summary <- project_summary %>%
  mutate(
    ProjectStartDate = mdy_hms(ProjectStartDate),
    ProjectEndDate = mdy_hms(ProjectEndDate)
  )
```

## Dataset Summary

```{r dataset-overview}
#| label: dataset-overview

data_summary <- tibble(
  Dataset = c("Allocation Flow", "Project Summary", "Project Summary (Location)", 
              "Project Summary (Cluster)", "Pipeline Summary", "Clusters", "Contributions"),
  Rows = c(nrow(allocation_flow), nrow(project_summary), nrow(project_summary_location),
           nrow(project_summary_cluster), nrow(pipeline_summary), nrow(clusters), 
           nrow(contributions)),
  Columns = c(ncol(allocation_flow), ncol(project_summary), ncol(project_summary_location),
              ncol(project_summary_cluster), ncol(pipeline_summary), ncol(clusters), 
              ncol(contributions))
)

kable(data_summary, 
      format.args = list(big.mark = ","),
      caption = "Overview of Available Datasets")
```

# Allocation Flow Analysis

## Countries in Dataset

```{r allocation-countries}
#| label: allocation-countries

countries <- unique(allocation_flow$Fund)
ukraine_allocation <- allocation_flow %>% filter(Fund == "Ukraine")

if(nrow(ukraine_allocation) > 0) {
  ukraine_summary <- ukraine_allocation %>%
    group_by(Year, `Partner type`) %>%
    summarise(
      `Direct Funding ($)` = sum(`Direct funding`, na.rm = TRUE),
      `Net Funding ($)` = sum(`Net funding`, na.rm = TRUE),
      .groups = 'drop'
    ) %>%
    arrange(Year, desc(`Net Funding ($)`))
  
  kable(ukraine_summary,
        format.args = list(big.mark = ","),
        caption = "Ukraine Allocation Summary by Year and Partner Type")
}
```

# Project Analysis

## Projects by Organization Type

```{r org-type-analysis}
#| label: org-type-analysis
#| fig-width: 10
#| fig-height: 6

org_type_summary <- project_summary %>%
  count(OrganizationType, name = "Projects") %>%
  arrange(desc(Projects))

# Table
kable(org_type_summary,
      format.args = list(big.mark = ","),
      caption = "Projects by Organization Type")

# Visualization
project_summary %>%
  group_by(OrganizationType) %>%
  summarise(Total_Budget = sum(Budget, na.rm = TRUE), .groups = 'drop') %>%
  ggplot(aes(x = reorder(OrganizationType, Total_Budget), y = Total_Budget/1e6)) +
  geom_col(fill = "#005f73") +
  coord_flip() +
  labs(
    title = "Total Budget by Organization Type",
    x = "Organization Type",
    y = "Total Budget (Millions USD)"
  )
```

## Projects Over Time

```{r projects-over-time}
#| label: projects-over-time
#| fig-width: 10
#| fig-height: 6

year_summary <- project_summary %>%
  count(AllocationYear, name = "Projects") %>%
  arrange(AllocationYear)

# Table
kable(year_summary,
      format.args = list(big.mark = ","),
      caption = "Projects by Allocation Year")

# Trend visualization
project_summary %>%
  count(AllocationYear) %>%
  ggplot(aes(x = AllocationYear, y = n)) +
  geom_line(color = "#005f73", linewidth = 1.2) +
  geom_point(color = "#005f73", size = 3) +
  labs(
    title = "Number of Projects by Allocation Year",
    x = "Year",
    y = "Number of Projects"
  )
```

# Budget Analysis

```{r budget-analysis}
#| label: budget-analysis

budget_stats <- project_summary %>%
  summarise(
    `Total Budget ($)` = sum(Budget, na.rm = TRUE),
    `Mean Budget ($)` = mean(Budget, na.rm = TRUE),
    `Median Budget ($)` = median(Budget, na.rm = TRUE),
    `Min Budget ($)` = min(Budget, na.rm = TRUE),
    `Max Budget ($)` = max(Budget, na.rm = TRUE)
  )

kable(budget_stats,
      format.args = list(big.mark = ",", digits = 2),
      caption = "Overall Budget Statistics")
```

## Budget Distribution by Allocation Source

```{r allocation-source-by-year}
#| label: allocation-source-by-year
#| fig-width: 10
#| fig-height: 6

# Calculate budget by allocation source and year (2022-2025)
allocation_source_summary <- project_summary %>%
  filter(AllocationYear >= 2022 & AllocationYear <= 2025) %>%
  group_by(AllocationYear, AllocationSourceName) %>%
  summarise(Total_Budget = sum(Budget, na.rm = TRUE), .groups = 'drop') %>%
  group_by(AllocationYear) %>%
  mutate(
    Year_Total = sum(Total_Budget),
    Percentage = (Total_Budget / Year_Total) * 100
  ) %>%
  arrange(AllocationYear, AllocationSourceName)

# Summary table
allocation_table <- allocation_source_summary %>%
  select(AllocationYear, AllocationSourceName, Total_Budget, Percentage) %>%
  mutate(
    `Budget ($)` = format(round(Total_Budget, 0), big.mark = ","),
    `Percentage (%)` = round(Percentage, 1)
  ) %>%
  select(Year = AllocationYear, `Allocation Source` = AllocationSourceName, 
         `Budget ($)`, `Percentage (%)`)

kable(allocation_table,
      caption = "Budget Distribution by Allocation Source and Year")

# Visualization - stacked bar chart with absolute values and percentages
# Convert to millions for display
allocation_source_summary <- allocation_source_summary %>%
  mutate(Budget_Millions = Total_Budget / 1e6)

ggplot(allocation_source_summary, 
       aes(x = factor(AllocationYear), y = Budget_Millions, fill = AllocationSourceName)) +
  geom_col(position = "stack", width = 0.7) +
  geom_text(aes(label = paste0("$", round(Budget_Millions, 1), "M\n(", 
                               round(Percentage, 1), "%)")),
            position = position_stack(vjust = 0.5),
            color = "white",
            fontface = "bold",
            size = 3.5,
            lineheight = 0.9) +
  scale_fill_manual(values = c("Standard" = "#005f73", "Reserve" = "#94d2bd"),
                    name = "Allocation Source") +
  scale_y_continuous(labels = function(x) paste0("$", x, "M"), 
                     expand = expansion(mult = c(0, 0.05))) +
  labs(
    title = "Budget Distribution by Allocation Source Across Years",
    subtitle = "Absolute values (millions USD) and percentage of total budget per year",
    x = "Allocation Year",
    y = "Total Budget (Millions USD)"
  ) +
  theme(
    legend.position = "top",
    panel.grid.major.x = element_blank()
  )
```

::: callout-tip
## Allocation Sources

-   **Standard Allocation**: Regular funding cycles based on strategic response plans
-   **Reserve Allocation**: Emergency funding for urgent needs and rapid response
:::

# Beneficiary Analysis

```{r beneficiary-analysis}
#| label: beneficiary-analysis
#| fig-width: 10
#| fig-height: 6

beneficiary_totals <- project_summary %>%
  summarise(
    Men = sum(Men, na.rm = TRUE),
    Women = sum(Women, na.rm = TRUE),
    Boys = sum(Boys, na.rm = TRUE),
    Girls = sum(Girls, na.rm = TRUE)
  ) %>%
  mutate(
    Adults = Men + Women,
    Children = Boys + Girls,
    Total = Adults + Children,
    `Female (%)` = round((Women + Girls) / Total * 100, 1),
    `Male (%)` = round((Men + Boys) / Total * 100, 1)
  )

# Summary table
beneficiary_summary <- tibble(
  Category = c("Men", "Women", "Boys", "Girls", "Total Adults", "Total Children", "Grand Total"),
  Count = c(
    beneficiary_totals$Men,
    beneficiary_totals$Women,
    beneficiary_totals$Boys,
    beneficiary_totals$Girls,
    beneficiary_totals$Adults,
    beneficiary_totals$Children,
    beneficiary_totals$Total
  )
)

kable(beneficiary_summary,
      format.args = list(big.mark = ","),
      caption = "Beneficiary Demographics")

# Visualization
beneficiary_totals %>%
  select(Men, Women, Boys, Girls) %>%
  pivot_longer(everything(), names_to = "Category", values_to = "Count") %>%
  ggplot(aes(x = Category, y = Count, fill = Category)) +
  geom_col() +
  geom_text(aes(label = format(Count, big.mark = ",")), 
            vjust = -0.5, size = 4) +
  scale_fill_manual(values = c("#005f73", "#2B638B", "#94d2bd", "#ee9b00")) +
  scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.1))) +
  labs(
    title = "Beneficiaries by Gender and Age Group",
    x = "Category",
    y = "Number of Beneficiaries"
  ) +
  theme(
    legend.position = "none"
  )
```

::: callout-note
## Gender Balance

The data shows **`r beneficiary_totals$'Female (%)'`%** female beneficiaries and **`r beneficiary_totals$'Male (%)'`%** male beneficiaries across all projects.
:::

# Project Status Overview

```{r project-status}
#| label: project-status

status_summary <- project_summary %>%
  count(ProjectStatus, name = "Count") %>%
  arrange(desc(Count))

kable(status_summary,
      format.args = list(big.mark = ","),
      caption = "Project Status Distribution")
```

# 2025 Funding Analysis {#sec-2025}

## Top Organizations

```{r funding-2025}
#| label: funding-2025
#| fig-width: 12
#| fig-height: 8

# Filter for 2025 projects and aggregate by organization
funding_2025 <- project_summary %>%
  filter(AllocationYear == 2025) %>%
  group_by(OrganizationName, OrganizationType) %>%
  summarise(
    `Total Funding ($)` = sum(Budget, na.rm = TRUE),
    `Number of Projects` = n(),
    .groups = 'drop'
  ) %>%
  arrange(desc(`Total Funding ($)`))

# Total 2025 funding
total_2025 <- sum(funding_2025$`Total Funding ($)`)

# Display top 20 organizations
kable(head(funding_2025, 20),
      format.args = list(big.mark = ","),
      caption = "Top 20 Organizations by Funding in 2025")

# Visualization: Top 15 Organizations
funding_2025 %>%
  head(15) %>%
  ggplot(aes(x = reorder(OrganizationName, `Total Funding ($)`), 
             y = `Total Funding ($)` / 1e6,
             fill = OrganizationType)) +
  geom_col() +
  geom_text(aes(label = paste0("$", format(round(`Total Funding ($)` / 1e6, 1), nsmall = 1), "M")),
            hjust = -0.1,
            size = 3.5,
            color = "black") +
  coord_flip() +
  scale_fill_manual(values = c("#005f73", "#2B638B", "#94d2bd", "#ee9b00")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = "Top 15 Organizations by Funding - 2025",
    subtitle = paste("Total 2025 Funding: $", format(round(total_2025/1e6, 1), nsmall = 1), "M"),
    x = "Organization",
    y = "Funding (Millions USD)",
    fill = "Organization Type"
  ) +
  theme(
    legend.position = "bottom",
    axis.text.y = element_text(size = 10)
  )
```

## Funding by Organization Type (2025)

```{r funding-type-2025}
#| label: funding-type-2025
#| fig-width: 10
#| fig-height: 6

funding_by_type_2025 <- project_summary %>%
  filter(AllocationYear == 2025) %>%
  group_by(OrganizationType) %>%
  summarise(
    `Total Funding ($)` = sum(Budget, na.rm = TRUE),
    `Number of Organizations` = n_distinct(OrganizationName),
    `Number of Projects` = n(),
    `Average Project Size ($)` = round(`Total Funding ($)` / `Number of Projects`, 0),
    .groups = 'drop'
  ) %>%
  arrange(desc(`Total Funding ($)`))

kable(funding_by_type_2025,
      format.args = list(big.mark = ","),
      caption = "2025 Funding by Organization Type")

# Pie chart for organization type distribution
funding_by_type_2025 %>%
  ggplot(aes(x = "", y = `Total Funding ($)`, fill = OrganizationType)) +
  geom_col(width = 1, color = "white") +
  geom_text(aes(label = paste0(OrganizationType, "\n$", 
                                format(round(`Total Funding ($)`/1e6, 1), nsmall = 1), "M")),
            position = position_stack(vjust = 0.5),
            size = 3.5) +
  coord_polar("y", start = 0) +
  scale_fill_manual(values = c("#005f73", "#2B638B", "#94d2bd", "#ee9b00")) +
  labs(
    title = "2025 Funding Distribution by Organization Type",
    fill = "Organization Type"
  ) +
  theme_void() +
  theme(
    legend.position = "right",
    plot.title = element_text(face = "bold", color = "#005f73", hjust = 0.5, size = 16)
  )
```

# Conclusions

## Summary Statistics

```{r final-summary}
#| label: final-summary

summary_stats <- tibble(
  Metric = c(
    "Total Projects",
    "Total Organizations",
    "Total Budget (All Years)",
    "Total Beneficiaries",
    "2025 Funding",
    "Average Project Budget"
  ),
  Value = c(
    format(nrow(project_summary), big.mark = ","),
    format(n_distinct(project_summary$OrganizationName), big.mark = ","),
    paste0("$", format(round(sum(project_summary$Budget, na.rm = TRUE)/1e6, 1), nsmall = 1), "M"),
    format(beneficiary_totals$Total, big.mark = ","),
    paste0("$", format(round(total_2025/1e6, 1), nsmall = 1), "M"),
    paste0("$", format(round(mean(project_summary$Budget, na.rm = TRUE)/1e3, 0), big.mark = ","), "K")
  )
)

kable(summary_stats,
      caption = "Key Metrics Summary")
```

## Key Insights

1.  **Organizational Diversity**: The fund supports multiple types of organizations including UN agencies, international NGOs, and national NGOs, ensuring broad implementation capacity.

2.  **Consistent Funding**: Projects show steady funding trends over the years with significant allocations in 2025.

3.  **Beneficiary Reach**: The program has reached a substantial number of beneficiaries across different demographic groups.

4.  **Gender Consideration**: The funding demonstrates attention to gender-balanced programming with relatively even distribution between male and female beneficiaries.

------------------------------------------------------------------------

*Report generated on `r Sys.Date()` using Quarto*