---
title: "Ukraine Humanitarian Pooled Fund Analysis"
subtitle: "Funding Allocation and Impact Assessment"
author: "Jesus Baena"
date: "2026-02-13"
format:
  html:
    theme: yeti
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-summary: "Show code"
    embed-resources: true
    smooth-scroll: true
    number-sections: true
    css: custom-style.css
editor: visual
execute:
  warning: false
  message: false
  echo: true
  fig-align: center
  fig-width: 8
  fig-height: 5
  fig-dpi: 150
---

# Introduction

This report provides a comprehensive analysis of the Ukraine Humanitarian Pooled Fund, **examining funding allocations and project implementation from 2019 to 2025**. These findings are contextualized within the global CBPF landscape, leveraging historical data from 28 country pooled funds spanning **2014 to 2025** (the mature operational period) to provide comparative insights and identify emerging trends in humanitarian financing.

## Key Findings

-   **Total Projects Analyzed**: Analysis spans multiple allocation years
-   **Primary Focus**: 2025 funding distribution and organizational performance
-   **Data Sources**: Official UHF allocation and project data

# Data Overview

Data is retrieved from the official repository [data explorer](https://cbpf.data.unocha.org/dataexplorer.html), it was downloaded on Febrary 2026 and it is formed by raw .csv files. The raw data spans from 2010 onward, but **analysis focuses on 2014-2025** to ensure comparability (2010-2013 was the pilot phase with limited geographic coverage). Data was cut at the end of 2025 to use static data for this analysis.

In total the data comprises:

-   28 recipient countries

-   16,304 projects

-   xx allocations

```{r setup}
#| label: setup
#| include: false

# Load required libraries
library(dplyr)
library(ggplot2)
library(tidyr)
library(readr)
library(lubridate)
library(stringr)
library(knitr)
library(scales)

# Set working directory
setwd("/home/jbi-debian/Git/2026-ukraine-poolfund-analysis")

# --- Consistent ggplot2 Theme ---
theme_set(
  theme_minimal(base_size = 14, base_family = "") +
  theme(
    plot.title = element_text(face = "bold", color = "#005f73", hjust = 0.5, size = 16),
    plot.subtitle = element_text(color = "#343a40", hjust = 0.5, margin = margin(b=10)),
    plot.caption = element_text(color = "grey50", face = "italic", size = 9),
    axis.title = element_text(face = "bold", size = 12),
    axis.text = element_text(color = "grey30", size = 10),
    legend.position = "top",
    legend.title = element_text(face = "bold", size = 11),
    panel.grid.minor = element_blank()
  )
)

# Color palette matching the theme
theme_colors <- list(
  primary = "#005f73",
  secondary = "#2B638B",
  accent = "#c7254e",
  light_bg = "#eef7f7"
)
```

```{r load-data}
#| label: load-data

# Load all data files - Latest download from February 19, 2026
# Using complete global dataset (ALL versions) for all 28 CBPF countries
project_summary <- read_csv("Raw-data/ProjectSummary_ALL_20260219_100323_UTC.csv", 
                            show_col_types = FALSE)
project_summary_location <- read_csv("Raw-data/ProjectSummaryWithLocation_ALL_20260219_100323_UTC.csv", 
                                     show_col_types = FALSE)
pipeline_summary <- read_csv("Raw-data/PipelineProjectSummary_ALL_20260219_125012_UTC.csv", 
                             show_col_types = FALSE)
pipeline_cluster <- read_csv("Raw-data/PipelineProjectCluster_ALL_20260219_100323_UTC.csv", 
                             show_col_types = FALSE)
clusters <- read_csv("Raw-data/Cluster_ALL_20260219_100323_UTC.csv", 
                     show_col_types = FALSE)
contributions <- read_csv("Raw-data/Contribution_ALL_20260219_100323_UTC.csv", 
                          show_col_types = FALSE)
```

## Data Preparation

The raw data is well-structured and consistent with expected values. Minor transformations were applied to standardise date formats across datasets and to filter out 2026 entries, keeping the analysis within the 2014–2025 window. Helper functions `safe_date_format` and `safe_year_format` handle edge cases in date parsing robustly.

::: {.callout-note collapse="true"}
## Data Preparation Details

**Date conversions**

| Dataset | Date columns converted |
|---|---|
| `ProjectSummary` | `ProjectStartDate`, `ProjectEndDate` → `mdy_hms()` |
| `ProjectSummaryWithLocation` | `ActualStartDate`, `ActualEndDate` → `mdy_hms()` |
| `PipelineProjectSummary` | `ActualStartDate`, `ActualEndDate` → `mdy_hms()` |
| `pipeline_cluster`, `clusters` | Use numeric `AllocationYear` — no conversion needed |
| `contributions` | `PledgeDate` parsed on demand during analysis |

**Temporal cutoff — analysis endpoint: December 31, 2025**

- `AllocationYear != 2026` applied to: `project_summary`, `project_summary_location`, `pipeline_summary`, `pipeline_cluster`
- `FiscalYear != 2026` applied to: `contributions`

**Helper functions**

- `safe_date_format()` — returns the min/max of a date vector, filtered to the 2000–2025 range; returns `"N/A"` if no valid dates remain
- `safe_year_format()` — same logic for numeric year vectors; returns the result as a character string
:::

```{r data-preparation}
#| label: data-preparation

# ============================================================
# PART 1: Date and Type Conversions
# ============================================================

# Parse date columns for project datasets
project_summary <- project_summary %>%
  mutate(
    ProjectStartDate = mdy_hms(ProjectStartDate),
    ProjectEndDate = mdy_hms(ProjectEndDate)
  )

# Note: project_summary_location and pipeline_summary use ActualStartDate
project_summary_location <- project_summary_location %>%
  mutate(
    ActualStartDate = mdy_hms(ActualStartDate),
    ActualEndDate = mdy_hms(ActualEndDate)
  )

pipeline_summary <- pipeline_summary %>%
  mutate(
    ActualStartDate = mdy_hms(ActualStartDate),
    ActualEndDate = mdy_hms(ActualEndDate)
  )

# Note: pipeline_cluster and clusters use AllocationYear (numeric)
# Note: contributions uses PledgeDate (will be parsed when needed)

# ============================================================
# PART 2: Filter Out 2026 Data
# ============================================================
# Remove all data from 2026 to maintain analysis cutoff at end of 2025

project_summary <- project_summary %>%
  filter(AllocationYear != 2026)

project_summary_location <- project_summary_location %>%
  filter(AllocationYear != 2026)

pipeline_summary <- pipeline_summary %>%
  filter(AllocationYear != 2026)

pipeline_cluster <- pipeline_cluster %>%
  filter(AllocationYear != 2026)

contributions <- contributions %>%
  filter(FiscalYear != 2026)

# ============================================================
# PART 3: Helper Functions
# ============================================================

# Safely format date ranges, filtering invalid dates
safe_date_format <- function(date_vec, func = min) {
  valid_dates <- date_vec[!is.na(date_vec) & 
                          year(date_vec) >= 2000 & 
                          year(date_vec) <= 2025]
  
  if (length(valid_dates) == 0) return("N/A")
  
  result <- func(valid_dates)
  return(format(result, "%B %d, %Y"))
}

# Format year ranges for datasets that use AllocationYear
safe_year_format <- function(year_vec, func = min) {
  valid_years <- year_vec[!is.na(year_vec) & year_vec >= 2000 & year_vec <= 2025]
  
  if (length(valid_years) == 0) return("N/A")
  
  return(as.character(func(valid_years)))
}
```

```{r dataset-overview}
#| label: dataset-overview

data_summary <- tibble(
  Dataset = c("Project Summary", "Project Summary (Location)", 
              "Pipeline Summary", "Pipeline Cluster", "Clusters", "Contributions"),
  Rows = c(nrow(project_summary), nrow(project_summary_location),
           nrow(pipeline_summary), nrow(pipeline_cluster), nrow(clusters), 
           nrow(contributions)),
  Columns = c(ncol(project_summary), ncol(project_summary_location),
              ncol(pipeline_summary), ncol(pipeline_cluster), ncol(clusters), 
              ncol(contributions)),
  Countries = c(
    n_distinct(project_summary$PooledFundName),
    n_distinct(project_summary_location$PooledFundName),
    n_distinct(pipeline_summary$PooledFundName),
    n_distinct(pipeline_cluster$PooledFundName),
    n_distinct(clusters$PooledFundName),
    n_distinct(contributions$PooledFundName)
  ),
  `Earliest Entry` = c(
    safe_date_format(project_summary$ProjectStartDate, min),
    safe_date_format(project_summary_location$ActualStartDate, min),
    safe_date_format(pipeline_summary$ActualStartDate, min),
    safe_year_format(pipeline_cluster$AllocationYear, min),
    safe_year_format(clusters$AllocationYear, min),
    safe_date_format(mdy(contributions$PledgeDate), min)
  ),
  `Latest Entry` = c(
    safe_date_format(project_summary$ProjectStartDate, max),
    safe_date_format(project_summary_location$ActualStartDate, max),
    safe_date_format(pipeline_summary$ActualStartDate, max),
    safe_year_format(pipeline_cluster$AllocationYear, max),
    safe_year_format(clusters$AllocationYear, max),
    safe_date_format(mdy(contributions$PledgeDate), max)
  )
)

kable(data_summary, 
      format.args = list(big.mark = ","),
      caption = "Overview of Available Datasets")
```



### Pipeline Data Note

::: callout-important
## Critical Data Limitation: Pipeline Data 

The data site offers an endpoint to retrieve projects on the pipeline (under review, cancelled, etc...). After multiple attempts to retrieve comprehensive pipeline data from the CBPF API, it was confirmed that the **PipelineProjectSummary endpoint returns only 581 entries from scattered years** (2013-2015, 2022, 2025-2026), representing just a fraction of all project submissions across the CBPF system's history and with some inconsistencies (note the earlier enrty in 2002, when Pooled Funds did not exist).

- **Ukraine-specific**: Only **1 entry** from 2022 allocation (out of 234 approved Ukraine projects in ProjectSummary)
- **Cannot calculate meaningful approval/rejection rates**: Missing data prevents comprehensive submission vs approval analysis

**What this means:** - We cannot analyze Ukraine’s proposal-to-approval ratio - We cannot determine how many Ukraine projects were rejected or cancelled - We cannot compare Ukraine’s approval success rate to other pooled funds - Analysis must focus on approved projects from the ProjectSummary dataset
:::


# Global summary 
There have been 46 pool funds established, benefiting 42 unique countries. Four countries have had 2 pool funds due to situations like cross-border operations: Colombia, Haiti, Pakistan and Sri Lanka. 

The data shows the humanitarian gap visible, in recent years, the number of countries receiving funding has increased, but the total budget allocated has gone in the wrong direction.  

```{r global-budget-countries}
#| label: global-budget-countries
#| fig-width: 12
#| fig-height: 7

# Aggregate data by year - separate Ukraine from other countries
# Keep all years (2010-2025) to show pilot phase vs mature operational period
global_summary_detailed <- project_summary %>%
  filter(!is.na(AllocationYear), !is.na(Budget)) %>%
  mutate(Country_Group = if_else(PooledFundName == "Ukraine", "Ukraine", "Other Countries")) %>%
  group_by(AllocationYear, Country_Group) %>%
  summarise(Budget = sum(Budget, na.rm = TRUE), .groups = 'drop') %>%
  arrange(AllocationYear, Country_Group)

# Get total budget by year for labels
global_totals <- global_summary_detailed %>%
  group_by(AllocationYear) %>%
  summarise(
    Total_Budget = sum(Budget),
    .groups = 'drop'
  )

# Get country count by year for the line
country_counts <- project_summary %>%
  filter(!is.na(AllocationYear)) %>%
  group_by(AllocationYear) %>%
  summarise(Countries = n_distinct(PooledFundName), .groups = 'drop')

# Merge totals with country counts
global_summary <- global_totals %>%
  left_join(country_counts, by = "AllocationYear")

# Create scaling factor for secondary axis
scale_factor <- max(global_summary$Total_Budget) / max(global_summary$Countries) / 1e6

# Create combination chart with Ukraine highlighted
ggplot() +
  # Stacked bars for budget (Ukraine in yellow, others in teal)
  geom_col(data = global_summary_detailed,
           aes(x = as.factor(AllocationYear), y = Budget/1e6, fill = Country_Group),
           alpha = 0.8, width = 0.7) +
  # Line for countries
  geom_line(data = global_summary,
            aes(x = as.factor(AllocationYear), y = Countries * scale_factor, group = 1),
            color = "#e63946", linewidth = 1.2) +
  geom_point(data = global_summary,
             aes(x = as.factor(AllocationYear), y = Countries * scale_factor),
             color = "#e63946", size = 3) +
  # Labels at top of bars
  geom_text(data = global_summary,
            aes(x = as.factor(AllocationYear), y = Total_Budget/1e6,
                label = paste0("$", round(Total_Budget/1e6, 0), "M")),
            vjust = -0.5, size = 3.5, fontface = "bold", color = "#333333") +
  # Color scale
  scale_fill_manual(
    values = c("Ukraine" = "#f4a261", "Other Countries" = "#005f73"),
    name = "Region"
  ) +
  # Primary y-axis
  scale_y_continuous(
    name = "Total Budget Allocated (Million USD)",
    labels = scales::comma,
    limits = c(0, NA),
    expand = expansion(mult = c(0, 0.1)),
    # Secondary y-axis
    sec.axis = sec_axis(~ . / scale_factor, 
                        name = "Number of Countries",
                        breaks = scales::pretty_breaks(n = 6))
  ) +
  labs(
    title = "Global CBPF Funding Allocation and Country Coverage by Year (2010-2025)",
    subtitle = "Budget shown in bars (Ukraine highlighted in yellow), country count shown as line • Pilot phase (2010-2013) shows limited scale",
    x = "Allocation Year"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 12, color = "#666666"),
    axis.title.y.left = element_text(color = "#005f73", face = "bold", size = 12),
    axis.title.y.right = element_text(color = "#e63946", face = "bold", size = 12),
    axis.text = element_text(size = 11),
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.minor = element_blank(),
    legend.position = "bottom",
    legend.title = element_text(face = "bold")
  )
```

::: callout-important
The 2010-2013 period represents the early pilot phase of CBPF with limited geographic coverage (primarily one country) and significantly lower funding levels. A major scale-up occurred in 2014 when the system expanded to multiple countries and reached operational maturity. **To ensure data comparability, subsequent analysis focuses on the mature operational period from 2014 to 2025.**
:::


## Budget Distribution by Country (2022-2025)

This visualization shows how CBPF funding was distributed across countries for each year from 2022 to 2025. Ukraine is highlighted in yellow to track its position among the top recipients throughout this period.

```{r budget-by-country-yearly}
#| label: budget-by-country-yearly
#| fig-width: 14
#| fig-height: 11

# Get top countries per year
country_budget_yearly <- project_summary %>%
  filter(AllocationYear >= 2022, AllocationYear <= 2025) %>%
  filter(!is.na(Budget), !is.na(PooledFundName)) %>%
  group_by(AllocationYear, PooledFundName) %>%
  summarise(
    Total_Budget = sum(Budget, na.rm = TRUE),
    Projects = n(),
    .groups = 'drop'
  )

# Get top 12 countries per year (or include Ukraine if not in top 12)
top_countries_per_year <- country_budget_yearly %>%
  group_by(AllocationYear) %>%
  arrange(desc(Total_Budget)) %>%
  slice_head(n = 12) %>%
  ungroup()

# Ensure Ukraine is included in all years
ukraine_data <- country_budget_yearly %>%
  filter(PooledFundName == "Ukraine")

# Combine and remove duplicates
country_budget_yearly_filtered <- bind_rows(top_countries_per_year, ukraine_data) %>%
  distinct(AllocationYear, PooledFundName, .keep_all = TRUE)

# Add highlighting, labels, and ordering within each year
country_budget_yearly_filtered <- country_budget_yearly_filtered %>%
  mutate(
    Highlight = if_else(PooledFundName == "Ukraine", "Ukraine", "Other")
  ) %>%
  group_by(AllocationYear) %>%
  mutate(
    Rank = rank(-Total_Budget),
    Label = if_else(Rank <= 8 | PooledFundName == "Ukraine",
                    paste0("$", round(Total_Budget/1e6, 0), "M"),
                    "")
  ) %>%
  ungroup() %>%
  # Create a compound key for proper ordering within facets
  mutate(
    Year_Country = paste(AllocationYear, PooledFundName, sep = "_"),
    Year_Country = reorder(Year_Country, Total_Budget)
  )

# Create faceted horizontal bar chart
ggplot(country_budget_yearly_filtered, 
       aes(x = Total_Budget/1e6, 
           y = Year_Country)) +
  geom_col(aes(fill = Highlight), alpha = 0.85, width = 0.75) +
  geom_text(aes(label = Label), hjust = -0.1, size = 3, fontface = "bold", color = "#333333") +
  scale_fill_manual(
    values = c("Ukraine" = "#f4a261", "Other" = "#005f73"),
    guide = "none"
  ) +
  scale_y_discrete(labels = function(x) gsub("^\\d{4}_", "", x)) +
  scale_x_continuous(
    labels = scales::comma,
    expand = expansion(mult = c(0, 0.15))
  ) +
  facet_wrap(~ AllocationYear, ncol = 2, scales = "free_y") +
  labs(
    title = "Top Countries by CBPF Budget Allocation per Year (2022-2025)",
    subtitle = "Ukraine highlighted in yellow • Shows top recipients each year and how rankings shift",
    x = "Budget (Million USD)",
    y = NULL
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 16, margin = margin(b = 5)),
    plot.subtitle = element_text(size = 11, color = "#666666", margin = margin(b = 15)),
    axis.text.y = element_text(size = 9.5),
    axis.text.x = element_text(size = 9),
    axis.title.x = element_text(face = "bold", size = 11, margin = margin(t = 10)),
    strip.text = element_text(face = "bold", size = 13, margin = margin(b = 10)),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    panel.spacing = unit(1.5, "lines"),
    plot.margin = margin(15, 15, 15, 15)
  )
```

Afganistan is the only country that has received more funding than Ukraine in 2022, this is due to commitments made at the global level following the Taliban takeover, probably made earlier than February 2022 as contributions to the appeal. 

Also note, that this cannot be taken as an indicator of the *global humanitarian attention*, CBPFs is a flexible global mechanism that can balance the funding situation between different crises, considering the attention given by other donors.  

# Allocation Flow Analysis

Country-Based Pool Funds have the two ways of publish allocations: the standard allocation and the reserve allocation. 

-   **Standard Allocation**: Regular funding cycles based on strategic response plans and humanitarian response planning
-   **Reserve Allocation**: Emergency funding for urgent needs, rapid response to sudden onset crises, and critical gaps

Key Differences from Standard Allocations:

- **Speed:** The timeline is compressed. A Reserve Allocation can be processed from launch to signature in as little as 48-72 hours in extreme cases, though 2-3 weeks is more common.
- **Targeting:** The strategy is highly prescriptive. The HC may define not just the geographic area but the specific activities required (e.g., "Provision of emergency shelter kits in District X").
- **Partner Selection:** While Standard Allocations are open calls, Reserve Allocations can be "closed" or restricted. The HC may invite specific partners who are known to have presence and capacity in the affected area to submit proposals, thereby bypassing the open competitive process to save time.


## Budget Distribution by Allocation Source

```{r allocation-count-by-source}
#| label: allocation-count-by-source
#| fig-width: 12
#| fig-height: 9

# Ukraine allocation summary
ukraine_allocation_summary <- project_summary %>%
  filter(AllocationYear >= 2014, 
         !is.na(AllocationSourceName),
         PooledFundName == "Ukraine") %>%
  group_by(AllocationSourceName) %>%
  summarise(
    Projects = n(),
    Budget = sum(Budget, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  mutate(
    Region = "Ukraine",
    # Force factor ordering: Standard first, Reserve second
    AllocationSourceName = factor(AllocationSourceName, levels = c("Standard", "Reserve"))
  ) %>%
  arrange(AllocationSourceName)

# Rest of World allocation summary (excluding Ukraine)
rest_of_world_summary <- project_summary %>%
  filter(AllocationYear >= 2014, 
         !is.na(AllocationSourceName),
         PooledFundName != "Ukraine") %>%
  group_by(AllocationSourceName) %>%
  summarise(
    Projects = n(),
    Budget = sum(Budget, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  mutate(
    Region = "Rest of World",
    # Force factor ordering: Standard first, Reserve second
    AllocationSourceName = factor(AllocationSourceName, levels = c("Standard", "Reserve"))
  ) %>%
  arrange(AllocationSourceName)

# Combine both datasets and ensure ordering
comparison_data <- bind_rows(ukraine_allocation_summary, rest_of_world_summary) %>%
  arrange(Region, AllocationSourceName)

# Recalculate percentages within each region
comparison_data <- comparison_data %>%
  group_by(Region) %>%
  mutate(
    Projects_Pct = round((Projects / sum(Projects)) * 100, 1),
    Budget_Pct = round((Budget / sum(Budget)) * 100, 1)
  ) %>%
  ungroup() %>%
  arrange(Region, AllocationSourceName)

# Budget proportions chart - Standard at bottom (dark teal), Reserve on top (light teal)
ggplot(comparison_data, 
       aes(x = Region, y = Budget_Pct, fill = AllocationSourceName)) +
  geom_col(position = position_stack(reverse = FALSE), width = 0.6) +
  geom_text(aes(label = paste0(AllocationSourceName, "\n", Budget_Pct, "%\n($", round(Budget/1e6, 0), "M)")),
            position = position_stack(vjust = 0.5, reverse = FALSE),
            color = "white",
            fontface = "bold",
            size = 3.8,
            lineheight = 0.85) +
  scale_fill_manual(values = c("Standard" = "#005f73", "Reserve" = "#94d2bd"),
                    name = "Allocation Type",
                    breaks = c("Standard", "Reserve")) +
  scale_y_continuous(labels = function(x) paste0(x, "%"),
                     expand = expansion(mult = c(0, 0.02))) +
  labs(
    title = "Budget Distribution by Allocation Type: Ukraine vs Rest of World (2014-2025)",
    subtitle = "Standard allocations (dark teal) at bottom, Reserve allocations (light teal) on top",
    x = NULL,
    y = "Percentage of Budget"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 11, color = "#666666"),
    panel.grid.major.x = element_blank(),
    axis.text.x = element_text(size = 12, face = "bold"),
    legend.position = "top",
    legend.text = element_text(size = 11, face = "bold")
  )
```

```{r allocation-source-by-year}
#| label: allocation-source-by-year
#| fig-width: 10
#| fig-height: 6

# Calculate budget by allocation source and year for UKRAINE only (2019-2025)
ukraine_allocation_by_year <- project_summary %>%
  filter(PooledFundName == "Ukraine", 
         AllocationYear >= 2019,
         !is.na(AllocationSourceName)) %>%
  group_by(AllocationYear, AllocationSourceName) %>%
  summarise(Total_Budget = sum(Budget, na.rm = TRUE), .groups = 'drop') %>%
  # Force factor ordering BEFORE calculating percentages
  mutate(
    AllocationSourceName = factor(AllocationSourceName, levels = c("Standard", "Reserve"))
  ) %>%
  arrange(AllocationYear, AllocationSourceName) %>%
  group_by(AllocationYear) %>%
  mutate(
    Year_Total = sum(Total_Budget),
    Percentage = round((Total_Budget / Year_Total) * 100, 1)
  ) %>%
  ungroup() %>%
  arrange(AllocationYear, AllocationSourceName)

# Visualization - line chart showing percentage over time
# Convert to millions for display
ukraine_allocation_by_year <- ukraine_allocation_by_year %>%
  mutate(Budget_Millions = Total_Budget / 1e6)

ggplot(ukraine_allocation_by_year, 
       aes(x = AllocationYear, y = Percentage, color = AllocationSourceName, group = AllocationSourceName)) +
  geom_line(linewidth = 1.5) +
  geom_point(size = 4) +
  geom_text(aes(label = paste0(Percentage, "%")),
            vjust = -1.2, 
            size = 3.5,
            fontface = "bold",
            show.legend = FALSE) +
  scale_color_manual(values = c("Standard" = "#005f73", "Reserve" = "#94d2bd"),
                     name = "Allocation Type",
                     breaks = c("Standard", "Reserve")) +
  scale_y_continuous(labels = function(x) paste0(x, "%"),
                     limits = c(0, 100),
                     breaks = seq(0, 100, 20),
                     expand = expansion(mult = c(0.05, 0.15))) +
  scale_x_continuous(breaks = seq(2019, 2025, 1)) +
  labs(
    title = "Ukraine Allocation Type Distribution Over Time (2019-2025)",
    subtitle = "Percentage of budget allocated through Standard vs Reserve modalities by year",
    x = "Allocation Year",
    y = "Percentage of Total Budget"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 11, color = "#666666"),
    legend.position = "top",
    legend.text = element_text(size = 11, face = "bold"),
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(size = 11)
  )
```


Although the balance between standard and reserve allocations appears to be correcting in 2025, it is important to note that since the counter-offensives of 2022, there have not been major events — such as large-scale population movements or recaptured territories — that would justify the recurrent use of the reserve modality. **Ukraine's humanitarian needs can be considered sustainably high, but not necessarily volatile.**

The global CBPF guidelines state that **only exceptionally** should the reserve allocation be used for non-emergency purposes: *"the Reserve Allocation may also be used to support special initiatives which may not be rapid onset or unforeseen per se..."* (CBPF Global Guidelines, art. 136).





------------------------------------------------------------------------

*Report generated on `r Sys.Date()` using Quarto*