---
title: "Ukraine Humanitarian Pooled Fund Analysis"
subtitle: "Funding Allocation and Grant Characteristics with emphasis on the health sector"
author: "Jesus Baena"
date: "2026-02-13"
format:
  html:
    theme: yeti
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-summary: "Show code"
    embed-resources: true
    smooth-scroll: true
    number-sections: true
    css: custom-style.css
    grid:
      sidebar-width: 300px
      body-width: 1100px
      margin-width: 0px
editor: visual
execute:
  warning: false
  message: false
  echo: true
  fig-align: center
  fig-width: 8
  fig-height: 5
  fig-dpi: 150
---

# Introduction

This report provides a comprehensive analysis of the Ukraine Humanitarian Pooled Fund, **examining funding allocations and project implementation from 2019 to 2025**. These findings are contextualized within the global CBPF landscape, leveraging historical data from 28 country pooled funds spanning **2014 to 2025** (the mature operational period) to provide comparative insights and identify emerging trends in humanitarian financing.

## Key Findings

-   **Higher use of the reserve allocation**: Although this is correcting in 2025, overall Ukraine has made more use of the reserve allocation than any other country-based pooled fund since this type of funding exists.
-   **Great concentration of the funding**: Ukraine grants are significantly more concentrated in a few partners than the rest of the world. 
-   **Data Sources**: Official UHF allocation and project data

# Data Overview

Data is retrieved from the official repository [data explorer](https://cbpf.data.unocha.org/dataexplorer.html), it was downloaded on Febrary 2026 and it is formed by raw .csv files. The raw data spans from 2010 onward, but **analysis focuses on 2014-2025** to ensure comparability (2010-2013 was the pilot phase with limited geographic coverage). Data was cut at the end of 2025 to use static data for this analysis.

In total the data comprises:

-   28 recipient countries

-   16,304 projects

-   xx allocations

```{r setup}
#| include: false

# Load required libraries
library(dplyr)
library(ggplot2)
library(tidyr)
library(readr)
library(lubridate)
library(stringr)
library(knitr)
library(scales)
library(patchwork)
library(ggtext)
library(gghighlight)
library(ggrepel)

# Set working directory
setwd("/home/jbi/Git/2026-ukraine-poolfund-analysis")

# --- Consistent ggplot2 Theme ---
theme_set(
  theme_minimal(base_size = 14, base_family = "") +
  theme(
    plot.title = element_text(face = "bold", color = "#005f73", hjust = 0.5, size = 16),
    plot.subtitle = element_text(color = "#343a40", hjust = 0.5, margin = margin(b=10)),
    plot.caption = element_text(color = "grey50", face = "italic", size = 9),
    axis.title = element_text(face = "bold", size = 12),
    axis.text = element_text(color = "grey30", size = 10),
    legend.position = "top",
    legend.title = element_text(face = "bold", size = 11),
    panel.grid.minor = element_blank()
  )
)

# Color palette matching the theme
theme_colors <- list(
  primary = "#005f73",
  secondary = "#2B638B",
  accent = "#c7254e",
  light_bg = "#eef7f7"
)
```

```{r load-data}

# Load all data files - Latest download from February 19, 2026
# Using complete global dataset (ALL versions) for all 28 CBPF countries
project_summary <- read_csv("Raw-data/ProjectSummary_ALL_20260219_100323_UTC.csv", 
                            show_col_types = FALSE)
project_summary_location <- read_csv("Raw-data/ProjectSummaryWithLocation_ALL_20260219_100323_UTC.csv", 
                                     show_col_types = FALSE)
pipeline_summary <- read_csv("Raw-data/PipelineProjectSummary_ALL_20260219_125012_UTC.csv", 
                             show_col_types = FALSE)
pipeline_cluster <- read_csv("Raw-data/PipelineProjectCluster_ALL_20260219_100323_UTC.csv", 
                             show_col_types = FALSE)
clusters <- read_csv("Raw-data/Cluster_ALL_20260219_100323_UTC.csv", 
                     show_col_types = FALSE)
contributions <- read_csv("Raw-data/Contribution_ALL_20260219_100323_UTC.csv", 
                          show_col_types = FALSE)
```

## Data Preparation

The raw data is well-structured and consistent with expected values. Minor transformations were applied to standardise date formats across datasets and to filter out 2026 entries, keeping the analysis within the 2014–2025 window. Helper functions `safe_date_format` and `safe_year_format` handle edge cases in date parsing robustly.

::: {.callout-note collapse="true"}
## Data Preparation Details

**Date conversions**

| Dataset | Date columns converted |
|---|---|
| `ProjectSummary` | `ProjectStartDate`, `ProjectEndDate` → `mdy_hms()` |
| `ProjectSummaryWithLocation` | `ActualStartDate`, `ActualEndDate` → `mdy_hms()` |
| `PipelineProjectSummary` | `ActualStartDate`, `ActualEndDate` → `mdy_hms()` |
| `pipeline_cluster`, `clusters` | Use numeric `AllocationYear` — no conversion needed |
| `contributions` | `PledgeDate` parsed on demand during analysis |

**Temporal cutoff — analysis endpoint: December 31, 2025**

- `AllocationYear != 2026` applied to: `project_summary`, `project_summary_location`, `pipeline_summary`, `pipeline_cluster`
- `FiscalYear != 2026` applied to: `contributions`

**Helper functions**

- `safe_date_format()` — returns the min/max of a date vector, filtered to the 2000–2025 range; returns `"N/A"` if no valid dates remain
- `safe_year_format()` — same logic for numeric year vectors; returns the result as a character string
:::

```{r data-preparation}

# ============================================================
# PART 1: Date and Type Conversions
# ============================================================

# Parse date columns for project datasets
project_summary <- project_summary %>%
  mutate(
    ProjectStartDate = mdy_hms(ProjectStartDate),
    ProjectEndDate = mdy_hms(ProjectEndDate)
  )

# Note: project_summary_location and pipeline_summary use ActualStartDate
project_summary_location <- project_summary_location %>%
  mutate(
    ActualStartDate = mdy_hms(ActualStartDate),
    ActualEndDate = mdy_hms(ActualEndDate)
  )

pipeline_summary <- pipeline_summary %>%
  mutate(
    ActualStartDate = mdy_hms(ActualStartDate),
    ActualEndDate = mdy_hms(ActualEndDate)
  )

# Note: pipeline_cluster and clusters use AllocationYear (numeric)
# Note: contributions uses PledgeDate (will be parsed when needed)

# ============================================================
# PART 2: Filter Out 2026 Data
# ============================================================
# Remove all data from 2026 to maintain analysis cutoff at end of 2025

project_summary <- project_summary %>%
  filter(AllocationYear != 2026)

project_summary_location <- project_summary_location %>%
  filter(AllocationYear != 2026)

pipeline_summary <- pipeline_summary %>%
  filter(AllocationYear != 2026)

pipeline_cluster <- pipeline_cluster %>%
  filter(AllocationYear != 2026)

contributions <- contributions %>%
  filter(FiscalYear != 2026)

# ============================================================
# PART 3: Helper Functions
# ============================================================

# Safely format date ranges, filtering invalid dates
safe_date_format <- function(date_vec, func = min) {
  valid_dates <- date_vec[!is.na(date_vec) & 
                          year(date_vec) >= 2000 & 
                          year(date_vec) <= 2025]
  
  if (length(valid_dates) == 0) return("N/A")
  
  result <- func(valid_dates)
  return(format(result, "%B %d, %Y"))
}

# Format year ranges for datasets that use AllocationYear
safe_year_format <- function(year_vec, func = min) {
  valid_years <- year_vec[!is.na(year_vec) & year_vec >= 2000 & year_vec <= 2025]
  
  if (length(valid_years) == 0) return("N/A")
  
  return(as.character(func(valid_years)))
}
```

```{r dataset-overview}

data_summary <- tibble(
  Dataset = c("Project Summary", "Project Summary (Location)", 
              "Pipeline Summary", "Pipeline Cluster", "Clusters", "Contributions"),
  Rows = c(nrow(project_summary), nrow(project_summary_location),
           nrow(pipeline_summary), nrow(pipeline_cluster), nrow(clusters), 
           nrow(contributions)),
  Columns = c(ncol(project_summary), ncol(project_summary_location),
              ncol(pipeline_summary), ncol(pipeline_cluster), ncol(clusters), 
              ncol(contributions)),
  Countries = c(
    n_distinct(project_summary$PooledFundName),
    n_distinct(project_summary_location$PooledFundName),
    n_distinct(pipeline_summary$PooledFundName),
    n_distinct(pipeline_cluster$PooledFundName),
    n_distinct(clusters$PooledFundName),
    n_distinct(contributions$PooledFundName)
  ),
  `Earliest Entry` = c(
    safe_date_format(project_summary$ProjectStartDate, min),
    safe_date_format(project_summary_location$ActualStartDate, min),
    safe_date_format(pipeline_summary$ActualStartDate, min),
    safe_year_format(pipeline_cluster$AllocationYear, min),
    safe_year_format(clusters$AllocationYear, min),
    safe_date_format(mdy(contributions$PledgeDate), min)
  ),
  `Latest Entry` = c(
    safe_date_format(project_summary$ProjectStartDate, max),
    safe_date_format(project_summary_location$ActualStartDate, max),
    safe_date_format(pipeline_summary$ActualStartDate, max),
    safe_year_format(pipeline_cluster$AllocationYear, max),
    safe_year_format(clusters$AllocationYear, max),
    safe_date_format(mdy(contributions$PledgeDate), max)
  )
)

kable(data_summary, 
      format.args = list(big.mark = ","),
      caption = "Overview of Available Datasets")
```



### Pipeline Data Note

::: callout-important
## Critical Data Limitation: Pipeline Data 

The data site offers an endpoint to retrieve projects on the pipeline (under review, cancelled, etc...). After multiple attempts to retrieve comprehensive pipeline data from the CBPF API, it was confirmed that the **PipelineProjectSummary endpoint returns only 581 entries from scattered years** (2013-2015, 2022, 2025-2026), representing just a fraction of all project submissions across the CBPF system's history and with some inconsistencies (note the earlier enrty in 2002, when Pooled Funds did not exist).

- **Ukraine-specific**: Only **1 entry** from 2022 allocation (out of 234 approved Ukraine projects in ProjectSummary)
- **Cannot calculate meaningful approval/rejection rates**: Missing data prevents comprehensive submission vs approval analysis

**What this means:** - We cannot analyze Ukraine’s proposal-to-approval ratio - We cannot determine how many Ukraine projects were rejected or cancelled - We cannot compare Ukraine’s approval success rate to other pooled funds - Analysis must focus on approved projects from the ProjectSummary dataset
:::


# Global summary 
There have been 46 pool funds established, benefiting 42 unique countries. Four countries have had 2 pool funds due to situations like cross-border operations: Colombia, Haiti, Pakistan and Sri Lanka. 

The data shows the humanitarian gap visible, in recent years, the number of countries receiving funding has increased, but the total budget allocated has gone in the wrong direction.  

```{r global-budget-countries}
#| fig-width: 12
#| fig-height: 7

# Aggregate data by year - separate Ukraine from other countries
# Keep all years (2010-2025) to show pilot phase vs mature operational period
global_summary_detailed <- project_summary %>%
  filter(!is.na(AllocationYear), !is.na(Budget)) %>%
  mutate(Country_Group = if_else(PooledFundName == "Ukraine", "Ukraine", "Other Countries")) %>%
  group_by(AllocationYear, Country_Group) %>%
  summarise(Budget = sum(Budget, na.rm = TRUE), .groups = 'drop') %>%
  arrange(AllocationYear, Country_Group)

# Get total budget by year for labels
global_totals <- global_summary_detailed %>%
  group_by(AllocationYear) %>%
  summarise(
    Total_Budget = sum(Budget),
    .groups = 'drop'
  )

# Get country count by year for the line
country_counts <- project_summary %>%
  filter(!is.na(AllocationYear)) %>%
  group_by(AllocationYear) %>%
  summarise(Countries = n_distinct(PooledFundName), .groups = 'drop')

# Merge totals with country counts
global_summary <- global_totals %>%
  left_join(country_counts, by = "AllocationYear")

# Create scaling factor for secondary axis
scale_factor <- max(global_summary$Total_Budget) / max(global_summary$Countries) / 1e6

# Create combination chart with Ukraine highlighted
ggplot() +
  # Stacked bars for budget (Ukraine in yellow, others in teal)
  geom_col(data = global_summary_detailed,
           aes(x = as.factor(AllocationYear), y = Budget/1e6, fill = Country_Group),
           alpha = 0.8, width = 0.7) +
  # Line for countries
  geom_line(data = global_summary,
            aes(x = as.factor(AllocationYear), y = Countries * scale_factor, group = 1),
            color = "#e63946", linewidth = 1.2) +
  geom_point(data = global_summary,
             aes(x = as.factor(AllocationYear), y = Countries * scale_factor),
             color = "#e63946", size = 3) +
  # Labels at top of bars
  geom_text(data = global_summary,
            aes(x = as.factor(AllocationYear), y = Total_Budget/1e6,
                label = paste0("$", round(Total_Budget/1e6, 0), "M")),
            vjust = -0.5, size = 3.5, fontface = "bold", color = "#333333") +
  # Color scale
  scale_fill_manual(
    values = c("Ukraine" = "#f4a261", "Other Countries" = "#005f73"),
    name = "Region"
  ) +
  # Primary y-axis
  scale_y_continuous(
    name = "Total Budget Allocated (Million USD)",
    labels = scales::comma,
    limits = c(0, NA),
    expand = expansion(mult = c(0, 0.1)),
    # Secondary y-axis
    sec.axis = sec_axis(~ . / scale_factor, 
                        name = "Number of Countries",
                        breaks = scales::pretty_breaks(n = 6))
  ) +
  labs(
    title = "Global CBPF Funding Allocation and Country Coverage by Year (2010-2025)",
    subtitle = "Budget shown in bars (Ukraine highlighted in yellow), country count shown as line • Pilot phase (2010-2013) shows limited scale",
    x = "Allocation Year"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 12, color = "#666666"),
    axis.title.y.left = element_text(color = "#005f73", face = "bold", size = 12),
    axis.title.y.right = element_text(color = "#e63946", face = "bold", size = 12),
    axis.text = element_text(size = 11),
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.minor = element_blank(),
    legend.position = "bottom",
    legend.title = element_text(face = "bold")
  )
```

::: callout-important
The 2010-2013 period represents the early pilot phase of CBPF with limited geographic coverage (primarily one country) and significantly lower funding levels. A major scale-up occurred in 2014 when the system expanded to multiple countries and reached operational maturity. **To ensure data comparability, subsequent analysis focuses on the mature operational period from 2014 to 2025.**
:::


## Budget Distribution by Country (2022-2025)

This visualization shows how CBPF funding was distributed across countries for each year from 2022 to 2025. Ukraine is highlighted in yellow to track its position among the top recipients throughout this period.

```{r budget-by-country-yearly}
#| fig-width: 14
#| fig-height: 11

# Get top countries per year
country_budget_yearly <- project_summary %>%
  filter(AllocationYear >= 2022, AllocationYear <= 2025) %>%
  filter(!is.na(Budget), !is.na(PooledFundName)) %>%
  group_by(AllocationYear, PooledFundName) %>%
  summarise(
    Total_Budget = sum(Budget, na.rm = TRUE),
    Projects = n(),
    .groups = 'drop'
  )

# Get top 12 countries per year (or include Ukraine if not in top 12)
top_countries_per_year <- country_budget_yearly %>%
  group_by(AllocationYear) %>%
  arrange(desc(Total_Budget)) %>%
  slice_head(n = 12) %>%
  ungroup()

# Ensure Ukraine is included in all years
ukraine_data <- country_budget_yearly %>%
  filter(PooledFundName == "Ukraine")

# Combine and remove duplicates
country_budget_yearly_filtered <- bind_rows(top_countries_per_year, ukraine_data) %>%
  distinct(AllocationYear, PooledFundName, .keep_all = TRUE)

# Add highlighting, labels, and ordering within each year
country_budget_yearly_filtered <- country_budget_yearly_filtered %>%
  mutate(
    Highlight = if_else(PooledFundName == "Ukraine", "Ukraine", "Other")
  ) %>%
  group_by(AllocationYear) %>%
  mutate(
    Rank = rank(-Total_Budget),
    Label = if_else(Rank <= 8 | PooledFundName == "Ukraine",
                    paste0("$", round(Total_Budget/1e6, 0), "M"),
                    "")
  ) %>%
  ungroup() %>%
  # Create a compound key for proper ordering within facets
  mutate(
    Year_Country = paste(AllocationYear, PooledFundName, sep = "_"),
    Year_Country = reorder(Year_Country, Total_Budget)
  )

# Create faceted horizontal bar chart
ggplot(country_budget_yearly_filtered, 
       aes(x = Total_Budget/1e6, 
           y = Year_Country)) +
  geom_col(aes(fill = Highlight), alpha = 0.85, width = 0.75) +
  geom_text(aes(label = Label), hjust = -0.1, size = 3, fontface = "bold", color = "#333333") +
  scale_fill_manual(
    values = c("Ukraine" = "#f4a261", "Other" = "#005f73"),
    guide = "none"
  ) +
  scale_y_discrete(labels = function(x) gsub("^\\d{4}_", "", x)) +
  scale_x_continuous(
    labels = scales::comma,
    expand = expansion(mult = c(0, 0.15))
  ) +
  facet_wrap(~ AllocationYear, ncol = 2, scales = "free_y") +
  labs(
    title = "Top Countries by CBPF Budget Allocation per Year (2022-2025)",
    subtitle = "Ukraine highlighted in yellow • Shows top recipients each year and how rankings shift",
    x = "Budget (Million USD)",
    y = NULL
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 16, margin = margin(b = 5)),
    plot.subtitle = element_text(size = 11, color = "#666666", margin = margin(b = 15)),
    axis.text.y = element_text(size = 9.5),
    axis.text.x = element_text(size = 9),
    axis.title.x = element_text(face = "bold", size = 11, margin = margin(t = 10)),
    strip.text = element_text(face = "bold", size = 13, margin = margin(b = 10)),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    panel.spacing = unit(1.5, "lines"),
    plot.margin = margin(15, 15, 15, 15)
  )
```

Afganistan is the only country that has received more funding than Ukraine in 2022, this is due to commitments made at the global level following the Taliban takeover, probably made earlier than February 2022 as contributions to the appeal. 

Also note, that this cannot be taken as an indicator of the *global humanitarian attention*, CBPFs is a flexible global mechanism that can balance the funding situation between different crises, considering the attention given by other donors.  

# Allocation Flow Analysis

Country-Based Pool Funds have the two ways of publish allocations: the standard allocation and the reserve allocation. 

-   **Standard Allocation**: Regular funding cycles based on strategic response plans and humanitarian response planning
-   **Reserve Allocation**: Emergency funding for urgent needs, rapid response to sudden onset crises, and critical gaps

Key Differences from Standard Allocations:

- **Speed:** The timeline is compressed. A Reserve Allocation can be processed from launch to signature in as little as 48-72 hours in extreme cases, though 2-3 weeks is more common.
- **Targeting:** The strategy is highly prescriptive. The HC may define not just the geographic area but the specific activities required (e.g., "Provision of emergency shelter kits in District X").
- **Partner Selection:** While Standard Allocations are open calls, Reserve Allocations can be "closed" or restricted. The HC may invite specific partners who are known to have presence and capacity in the affected area to submit proposals, thereby bypassing the open competitive process to save time.


## Budget Distribution by Allocation Source

**UHF has made more use of the reserve allocation than any other country-based pooled fund**, with a significant portion of its funding coming from this modality. This chart compares the distribution of standard vs reserve allocations for Ukraine against the rest of the world, showing that while Ukraine has a higher reliance on reserve allocations, there is a trend towards more balanced funding in 2025.

```{r allocation-count-by-source}
#| fig-width: 12
#| fig-height: 9

# Ukraine allocation summary
ukraine_allocation_summary <- project_summary %>%
  filter(AllocationYear >= 2014, 
         !is.na(AllocationSourceName),
         PooledFundName == "Ukraine") %>%
  group_by(AllocationSourceName) %>%
  summarise(
    Projects = n(),
    Budget = sum(Budget, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  mutate(
    Region = "Ukraine",
    # Force factor ordering: Standard first, Reserve second
    AllocationSourceName = factor(AllocationSourceName, levels = c("Standard", "Reserve"))
  ) %>%
  arrange(AllocationSourceName)

# Rest of World allocation summary (excluding Ukraine)
rest_of_world_summary <- project_summary %>%
  filter(AllocationYear >= 2014, 
         !is.na(AllocationSourceName),
         PooledFundName != "Ukraine") %>%
  group_by(AllocationSourceName) %>%
  summarise(
    Projects = n(),
    Budget = sum(Budget, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  mutate(
    Region = "Rest of World",
    # Force factor ordering: Standard first, Reserve second
    AllocationSourceName = factor(AllocationSourceName, levels = c("Standard", "Reserve"))
  ) %>%
  arrange(AllocationSourceName)

# Combine both datasets and ensure ordering
comparison_data <- bind_rows(ukraine_allocation_summary, rest_of_world_summary) %>%
  arrange(Region, AllocationSourceName)

# Recalculate percentages within each region
comparison_data <- comparison_data %>%
  group_by(Region) %>%
  mutate(
    Projects_Pct = round((Projects / sum(Projects)) * 100, 1),
    Budget_Pct = round((Budget / sum(Budget)) * 100, 1)
  ) %>%
  ungroup() %>%
  arrange(Region, AllocationSourceName)

# Budget proportions chart - Standard at bottom (dark teal), Reserve on top (light teal)
ggplot(comparison_data, 
       aes(x = Region, y = Budget_Pct, fill = AllocationSourceName)) +
  geom_col(position = position_stack(reverse = FALSE), width = 0.6) +
  geom_text(aes(label = paste0(AllocationSourceName, "\n", Budget_Pct, "%\n($", round(Budget/1e6, 0), "M)")),
            position = position_stack(vjust = 0.5, reverse = FALSE),
            color = "white",
            fontface = "bold",
            size = 3.8,
            lineheight = 0.85) +
  scale_fill_manual(values = c("Standard" = "#005f73", "Reserve" = "#94d2bd"),
                    name = "Allocation Type",
                    breaks = c("Standard", "Reserve")) +
  scale_y_continuous(labels = function(x) paste0(x, "%"),
                     expand = expansion(mult = c(0, 0.02))) +
  labs(
    title = "Budget Distribution by Allocation Type: Ukraine vs Rest of World (2014-2025)",
    subtitle = "Standard allocations (dark teal) at bottom, Reserve allocations (light teal) on top",
    x = NULL,
    y = "Percentage of Budget"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 11, color = "#666666"),
    panel.grid.major.x = element_blank(),
    axis.text.x = element_text(size = 12, face = "bold"),
    legend.position = "top",
    legend.text = element_text(size = 11, face = "bold")
  )
```

```{r allocation-source-by-year}
#| fig-width: 10
#| fig-height: 6

# Calculate budget by allocation source and year for UKRAINE only (2019-2025)
ukraine_allocation_by_year <- project_summary %>%
  filter(PooledFundName == "Ukraine", 
         AllocationYear >= 2019,
         !is.na(AllocationSourceName)) %>%
  group_by(AllocationYear, AllocationSourceName) %>%
  summarise(Total_Budget = sum(Budget, na.rm = TRUE), .groups = 'drop') %>%
  # Force factor ordering BEFORE calculating percentages
  mutate(
    AllocationSourceName = factor(AllocationSourceName, levels = c("Standard", "Reserve"))
  ) %>%
  arrange(AllocationYear, AllocationSourceName) %>%
  group_by(AllocationYear) %>%
  mutate(
    Year_Total = sum(Total_Budget),
    Percentage = round((Total_Budget / Year_Total) * 100, 1)
  ) %>%
  ungroup() %>%
  arrange(AllocationYear, AllocationSourceName)

# Visualization - line chart showing percentage over time
# Convert to millions for display
ukraine_allocation_by_year <- ukraine_allocation_by_year %>%
  mutate(Budget_Millions = Total_Budget / 1e6)

ggplot(ukraine_allocation_by_year, 
       aes(x = AllocationYear, y = Percentage, color = AllocationSourceName, group = AllocationSourceName)) +
  geom_line(linewidth = 1.5) +
  geom_point(size = 4) +
  geom_text(aes(label = paste0(Percentage, "%")),
            vjust = -1.2, 
            size = 3.5,
            fontface = "bold",
            show.legend = FALSE) +
  scale_color_manual(values = c("Standard" = "#005f73", "Reserve" = "#94d2bd"),
                     name = "Allocation Type",
                     breaks = c("Standard", "Reserve")) +
  scale_y_continuous(labels = function(x) paste0(x, "%"),
                     limits = c(0, 100),
                     breaks = seq(0, 100, 20),
                     expand = expansion(mult = c(0.05, 0.15))) +
  scale_x_continuous(breaks = seq(2019, 2025, 1)) +
  labs(
    title = "Ukraine Allocation Type Distribution Over Time (2019-2025)",
    subtitle = "Percentage of budget allocated through Standard vs Reserve modalities by year",
    x = "Allocation Year",
    y = "Percentage of Total Budget"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 11, color = "#666666"),
    legend.position = "top",
    legend.text = element_text(size = 11, face = "bold"),
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(size = 11)
  )
```


Although the balance between standard and reserve allocations appears to be correcting in 2025, it is important to note that since the counter-offensives of 2022, there have not been major events — such as large-scale population movements or recaptured territories — that would justify the recurrent use of the reserve modality. **Ukraine's humanitarian needs can be considered sustainably high, but not necessarily volatile.**

The global CBPF guidelines state that **only exceptionally** should the reserve allocation be used for non-emergency purposes: *"the Reserve Allocation may also be used to support special initiatives which may not be rapid onset or unforeseen per se..."* (CBPF Global Guidelines, art. 136).


## Budget Distribution by Type of Organization

CBPF allocations reach beneficiaries through four types of implementing partners: International NGOs (INGOs), National NGOs (NNGOs), UN Agencies, and a residual "Others" category. The balance between these partner types reflects both operational realities on the ground and broader policy commitments — notably the Grand Bargain pledge to channel at least 25% of humanitarian funding to local and national actors.

```{r org-type-combined}
#| fig-width: 14
#| fig-height: 7

org_colors <- c("National NGO"      = "#2a9d8f",
                "International NGO" = "#005f73",
                "UN Agency"         = "#e9c46a",
                "Others"            = "#adb5bd")

# --- Left panel: proportional comparison Ukraine vs Rest of World ---
org_type_ukraine <- project_summary %>%
  filter(AllocationYear >= 2014, !is.na(OrganizationType),
         PooledFundName == "Ukraine") %>%
  group_by(OrganizationType) %>%
  summarise(Budget = sum(Budget, na.rm = TRUE), Projects = n(), .groups = "drop") %>%
  mutate(Region = "Ukraine")

org_type_row <- project_summary %>%
  filter(AllocationYear >= 2014, !is.na(OrganizationType),
         PooledFundName != "Ukraine") %>%
  group_by(OrganizationType) %>%
  summarise(Budget = sum(Budget, na.rm = TRUE), Projects = n(), .groups = "drop") %>%
  mutate(Region = "Rest of World")

org_type_comparison <- bind_rows(org_type_ukraine, org_type_row) %>%
  group_by(Region) %>%
  mutate(
    Budget_Pct = round((Budget / sum(Budget)) * 100, 1),
    OrganizationType = factor(OrganizationType, levels = names(org_colors))
  ) %>%
  ungroup()

p_comparison <- ggplot(org_type_comparison,
       aes(x = Region, y = Budget_Pct, fill = OrganizationType)) +
  geom_col(position = position_stack(reverse = TRUE), width = 0.55) +
  geom_text(aes(label = paste0(OrganizationType, "\n", Budget_Pct, "%")),
            position = position_stack(vjust = 0.5, reverse = TRUE),
            color = "white", fontface = "bold", size = 3.2, lineheight = 0.9) +
  scale_fill_manual(values = org_colors, name = "Organization Type") +
  scale_y_continuous(labels = function(x) paste0(x, "%"),
                     expand = expansion(mult = c(0, 0.02))) +
  labs(title = "Ukraine vs Rest of World",
       subtitle = "Budget share by org type (2014-2025)",
       x = NULL, y = "Percentage of Budget") +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(size = 10, color = "#666666"),
    panel.grid.major.x = element_blank(),
    axis.text.x = element_text(size = 11, face = "bold"),
    legend.position = "none"
  )

# --- Right panel: Ukraine absolute budget over time ---
ukraine_org_year <- project_summary %>%
  filter(PooledFundName == "Ukraine", AllocationYear >= 2022,
         !is.na(OrganizationType)) %>%
  group_by(AllocationYear, OrganizationType) %>%
  summarise(Budget = sum(Budget, na.rm = TRUE), .groups = "drop") %>%
  mutate(OrganizationType = factor(OrganizationType, levels = names(org_colors)))

p_over_time <- ggplot(ukraine_org_year,
       aes(x = as.factor(AllocationYear), y = Budget / 1e6, fill = OrganizationType)) +
  geom_col(position = position_stack(reverse = TRUE), width = 0.7, alpha = 0.9) +
  scale_fill_manual(values = org_colors, name = "Organization Type") +
  scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.08))) +
  labs(title = "Ukraine Budget by Organization Type Over Time",
       subtitle = "USD millions (2022-2025)",
       x = "Allocation Year", y = "Budget (USD millions)") +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(size = 10, color = "#666666"),
    axis.text.x = element_text(size = 9, angle = 45, hjust = 1),
    legend.position = "top",
    legend.text = element_text(size = 9)
  )

p_comparison + p_over_time + plot_layout(widths = c(1, 2))
```

### Localization Trend: National NGO Share Over Time

```{r localization-trend}
#| fig-width: 12
#| fig-height: 6

# Calculate NNGO % of budget per year for Ukraine and global average
localization_ukraine <- project_summary %>%
  filter(AllocationYear >= 2014, !is.na(OrganizationType)) %>%
  group_by(AllocationYear, PooledFundName) %>%
  mutate(Year_Fund_Total = sum(Budget, na.rm = TRUE)) %>%
  filter(OrganizationType == "National NGO") %>%
  summarise(
    NNGO_Budget = sum(Budget, na.rm = TRUE),
    Year_Fund_Total = first(Year_Fund_Total),
    .groups = "drop"
  ) %>%
  mutate(NNGO_Pct = round((NNGO_Budget / Year_Fund_Total) * 100, 1))

# Global average per year (all funds EXCEPT Ukraine)
global_nngo <- localization_ukraine %>%
  filter(PooledFundName != "Ukraine") %>%
  group_by(AllocationYear) %>%
  summarise(
    NNGO_Pct = round(sum(NNGO_Budget) / sum(Year_Fund_Total) * 100, 1),
    .groups = "drop"
  ) %>%
  mutate(Series = "Global Average")

ukraine_nngo <- localization_ukraine %>%
  filter(PooledFundName == "Ukraine") %>%
  select(AllocationYear, NNGO_Pct) %>%
  mutate(Series = "Ukraine UHF")

localization_plot <- bind_rows(ukraine_nngo, global_nngo)

ggplot(localization_plot,
       aes(x = AllocationYear, y = NNGO_Pct, color = Series, group = Series)) +
  geom_hline(yintercept = 25, linetype = "dashed", color = "#e63946", linewidth = 0.8) +
  annotate("text", x = 2014.2, y = 26.5, label = "Grand Bargain 25% target",
           color = "#e63946", size = 3.5, hjust = 0, fontface = "italic") +
  geom_line(linewidth = 1.5) +
  geom_point(size = 4) +
  geom_text(aes(label = paste0(NNGO_Pct, "%")),
            vjust = -1.2, size = 3.5, fontface = "bold", show.legend = FALSE) +
  scale_color_manual(
    values = c("Ukraine UHF" = "#005f73", "Global Average" = "#94d2bd"),
    name = NULL
  ) +
  scale_y_continuous(labels = function(x) paste0(x, "%"),
                     limits = c(0, 60),
                     breaks = seq(0, 60, 10),
                     expand = expansion(mult = c(0.02, 0.12))) +
  scale_x_continuous(breaks = seq(2014, 2025, 1)) +
  labs(
    title = "National NGO Budget Share Over Time: Ukraine vs Global Average (2014–2025)",
    subtitle = "Dashed red line marks the Grand Bargain 25% localization target",
    x = "Allocation Year", y = "National NGO % of Total Budget"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 11, color = "#666666"),
    legend.position = "top",
    legend.text = element_text(size = 11, face = "bold"),
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(size = 10, angle = 45, hjust = 1)
  )
```


## Budget Distribution by Sector (Cluster)

This section examines how humanitarian funding is distributed across sectors, comparing the global CBPF portfolio with Ukraine's UHF allocations (2014–2025). Sector labels in the CBPF system correspond to inter-agency clusters (e.g., Health, WASH, Protection).

```{r cluster-prep}
#| include: false

# Build sector-level aggregates from the clusters dataset (already filtered to ≤2025)
clusters_clean <- clusters %>%
  filter(AllocationYear != 2026, !is.na(Cluster), Cluster != "")

# Standardise minor cluster labels for readability
clusters_clean <- clusters_clean %>%
  mutate(Cluster = case_when(
    Cluster == "Camp Coordination / Management" ~ "Camp Coordination/Mgmt",
    Cluster == "Coordination and Support Services" ~ "Coord. & Support Services",
    Cluster == "Emergency Shelter and NFI" ~ "Emergency Shelter & NFI",
    Cluster == "Emergency Telecommunications" ~ "Emergency Telecoms",
    Cluster == "Water, Sanitation and Hygiene" ~ "WASH",
    TRUE ~ Cluster
  ))

# ---------- Global aggregate (all funds, 2014-2025) ----------
global_cluster <- clusters_clean %>%
  filter(AllocationYear >= 2014) %>%
  group_by(Cluster) %>%
  summarise(Total = sum(ClusterBudget, na.rm = TRUE), .groups = "drop") %>%
  mutate(Share = Total / sum(Total) * 100,
         Source = "Global CBPF")

# ---------- Ukraine aggregate (2014-2025) ----------
ukraine_cluster <- clusters_clean %>%
  filter(PooledFundName == "Ukraine", AllocationYear >= 2014) %>%
  group_by(Cluster) %>%
  summarise(Total = sum(ClusterBudget, na.rm = TRUE), .groups = "drop") %>%
  mutate(Share = Total / sum(Total) * 100,
         Source = "Ukraine UHF")

# Ordered by global share for consistent axis
cluster_order <- global_cluster %>%
  arrange(Share) %>%
  pull(Cluster)

# Combined for comparison chart
comparison_df <- bind_rows(global_cluster, ukraine_cluster) %>%
  mutate(Cluster = factor(Cluster, levels = cluster_order))
```

### Global vs Ukraine: Sector Budget Share

```{r cluster-comparison}
#| fig-width: 12
#| fig-height: 7

# Colour palette: one shade per source
source_colors <- c("Global CBPF" = "#94d2bd", "Ukraine UHF" = "#005f73")

ggplot(comparison_df,
       aes(x = Cluster, y = Share, fill = Source)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.65, alpha = 0.9) +
  geom_text(aes(label = paste0(round(Share, 1), "%")),
            position = position_dodge(width = 0.75),
            hjust = -0.15, size = 3.2, fontface = "bold") +
  coord_flip() +
  scale_fill_manual(values = source_colors, name = NULL) +
  scale_y_continuous(
    labels = function(x) paste0(x, "%"),
    expand = expansion(mult = c(0, 0.25))
  ) +
  labs(
    title = "Sector Budget Share: Global CBPF vs Ukraine UHF (2014–2025)",
    subtitle = "Percentage of total allocated budget per sector",
    x = NULL, y = "Share of Total Budget (%)",
    caption = "Source: CBPF Cluster dataset, Feb 2026 extract"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold", size = 14, color = "#005f73"),
    plot.subtitle = element_text(size = 11, color = "#555555"),
    legend.position = "top",
    legend.text   = element_text(size = 11, face = "bold"),
    panel.grid.major.y = element_blank(),
    panel.grid.minor   = element_blank(),
    axis.text.y        = element_text(size = 11)
  )
```

::: {.callout-note}
**Key observations:** Emergency Shelter & NFI and Protection dominate Ukraine's portfolio—reflecting the conflict-driven displacement crisis—while globally WASH and Health are the largest sectors. Ukraine's WASH and Health shares are notably lower than the global average, highlighting a context-specific funding pattern.
:::

### Ukraine Sector Budget Evolution Over Time

```{r cluster-ukraine-trend}
#| fig-width: 12
#| fig-height: 6

# Yearly breakdown for Ukraine per cluster
ukraine_yearly <- clusters_clean %>%
  filter(PooledFundName == "Ukraine", AllocationYear >= 2022) %>%
  group_by(AllocationYear, Cluster) %>%
  summarise(Budget_M = sum(ClusterBudget, na.rm = TRUE) / 1e6, .groups = "drop")

# Keep sector order consistent with global ranking
sector_palette <- c(
  "WASH"                         = "#0077B6",
  "Health"                       = "#e63946",  # highlighted
  "Food Security"                = "#90E0EF",
  "Emergency Shelter & NFI"      = "#005f73",
  "Protection"                   = "#0A9396",
  "Nutrition"                    = "#94D2BD",
  "Education"                    = "#E9C46A",
  "Camp Coordination/Mgmt"       = "#F4A261",
  "Multi-purpose CASH"           = "#E76F51",
  "Coord. & Support Services"    = "#ae2012",
  "Multi-Sector"                 = "#9b2226",
  "Early Recovery"               = "#6d6875",
  "Logistics"                    = "#b5b5b5",
  "Emergency Telecoms"           = "#d4d4d4",
  "COVID-19"                     = "#ccc5b9"
)

ukraine_yearly <- ukraine_yearly %>%
  mutate(Cluster = factor(Cluster, levels = rev(cluster_order)))

ggplot(ukraine_yearly,
       aes(x = as.factor(AllocationYear), y = Budget_M, fill = Cluster)) +
  geom_col(position = position_stack(reverse = TRUE), width = 0.75, alpha = 0.92) +
  scale_fill_manual(values = sector_palette, name = "Sector") +
  scale_y_continuous(
    labels = scales::comma,
    expand = expansion(mult = c(0, 0.06))
  ) +
  labs(
    title = "Ukraine UHF: Sector Budget Distribution Over Time (2022–2025)",
    subtitle = "USD millions, stacked by cluster  |  \u2665 Health sector highlighted",
    x = "Allocation Year", y = "Budget (USD millions)",
    caption = "Source: CBPF Cluster dataset, Feb 2026 extract"
  ) +
  guides(fill = guide_legend(nrow = 3, byrow = TRUE)) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold", size = 14, color = "#005f73"),
    plot.subtitle = element_text(size = 11, color = "#555555"),
    legend.position = "bottom",
    legend.text   = element_text(size = 9),
    panel.grid.minor   = element_blank(),
    axis.text.x        = element_text(size = 10, angle = 45, hjust = 1)
  )
```

### Divergence from Global Average

```{r cluster-divergence}
#| fig-width: 11
#| fig-height: 6

# Calculate difference: Ukraine share minus global share
divergence_df <- ukraine_cluster %>%
  select(Cluster, Ukraine_Share = Share) %>%
  left_join(
    global_cluster %>% select(Cluster, Global_Share = Share),
    by = "Cluster"
  ) %>%
  mutate(
    Global_Share  = replace_na(Global_Share, 0),
    Diff          = Ukraine_Share - Global_Share,
    Direction     = if_else(Diff >= 0, "Above global average", "Below global average"),
    Cluster       = factor(Cluster, levels = cluster_order)
  )

ggplot(divergence_df,
       aes(x = Cluster, y = Diff, fill = Direction)) +
  geom_col(width = 0.7, alpha = 0.9) +
  geom_hline(yintercept = 0, linewidth = 0.8, color = "grey30") +
  geom_text(aes(label = paste0(if_else(Diff > 0, "+", ""), round(Diff, 1), " pp"),
                hjust = if_else(Diff >= 0, -0.1, 1.1)),
            size = 3.4, fontface = "bold") +
  coord_flip() +
  scale_fill_manual(
    values = c("Above global average" = "#005f73", "Below global average" = "#e76f51"),
    name = NULL
  ) +
  scale_y_continuous(
    labels = function(x) paste0(x, " pp"),
    expand = expansion(mult = c(0.25, 0.25))
  ) +
  labs(
    title = "Ukraine UHF Sector Share vs Global CBPF Average (2014–2025)",
    subtitle = "Percentage-point difference (Ukraine minus global); pp = percentage points",
    x = NULL, y = "Difference (percentage points)",
    caption = "Source: CBPF Cluster dataset, Feb 2026 extract"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold", size = 14, color = "#005f73"),
    plot.subtitle = element_text(size = 11, color = "#555555"),
    legend.position = "top",
    legend.text   = element_text(size = 11, face = "bold"),
    panel.grid.major.y = element_blank(),
    panel.grid.minor   = element_blank(),
    axis.text.y        = element_text(size = 11)
  )
```



# Grant Characteristics and Patterns




------------------------------------------------------------------------

# Grant Size Distribution

This chapter examines how large individual grants are across the CBPF system, comparing global patterns with Ukraine. Grant size reflects strategic choices: larger grants reduce transaction costs but concentrate risk; smaller grants enable broader partner diversity. Ukraine, as a high-intensity conflict fund, may follow different patterns.

```{r grant-prep}
#| include: false

# Base dataset: all valid grants 2014-2025
grants <- project_summary %>%
  filter(AllocationYear >= 2014, !is.na(Budget), Budget > 0) %>%
  mutate(
    Context = if_else(PooledFundName == "Ukraine", "Ukraine UHF", "Global CBPF"),
    OrgType = case_when(
      OrganizationType == "International NGO" ~ "INGO",
      OrganizationType == "National NGO"      ~ "NNGO",
      OrganizationType == "UN Agency"         ~ "UN Agency",
      TRUE                                    ~ NA_character_
    )
  ) %>%
  filter(!is.na(OrgType))

# Join primary cluster (highest budget percentage) from cluster dataset
cl_primary <- clusters %>%
  filter(AllocationYear >= 2014) %>%
  group_by(ChfId) %>%
  slice_max(Percentage, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(ChfId, Cluster) %>%
  mutate(Cluster = case_when(
    Cluster == "Water, Sanitation and Hygiene"      ~ "WASH",
    Cluster == "Emergency Shelter and NFI"          ~ "Shelter & NFI",
    Cluster == "Camp Coordination / Management"     ~ "Camp Coord./Mgmt",
    Cluster == "Coordination and Support Services"  ~ "Coord. & Support",
    Cluster == "Emergency Telecommunications"       ~ "Telecoms",
    TRUE ~ Cluster
  ))

grants_cl <- grants %>%
  left_join(cl_primary, by = "ChfId") %>%
  mutate(Cluster = replace_na(Cluster, "Unknown"),
         is_health = Cluster == "Health")

# Shared colour scheme
ctx_colors <- c("Global CBPF" = "#94d2bd", "Ukraine UHF" = "#005f73")
org_order   <- c("UN Agency", "INGO", "NNGO")
```

## Grant Size: Global vs Ukraine

The density plot below uses a **log₁₀ scale** to handle the wide spread of grant values (from ~\$25K to \$29M). Both distributions are right-skewed; the key question is whether Ukraine's curve sits systematically to the right.

```{r grant-density}
#| fig-width: 11
#| fig-height: 5

# Summary stats for annotation
medians <- grants %>%
  group_by(Context) %>%
  summarise(med = median(Budget), .groups = "drop")

fmt_usd <- function(x) {
  ifelse(x >= 1e6,
    paste0("$", formatC(x / 1e6, format = "f", digits = 2), "M"),
    paste0("$", scales::comma(round(x / 1e3)), "K")
  )
}

ggplot(grants, aes(x = Budget, fill = Context, color = Context)) +
  geom_density(alpha = 0.35, linewidth = 0.9, adjust = 0.8) +
  geom_vline(data = medians,
             aes(xintercept = med, color = Context),
             linetype = "dashed", linewidth = 1.1) +
  geom_label(data = medians,
             aes(x = med, y = Inf,
                 label = paste0("Median\n", fmt_usd(med)),
                 color = Context),
             vjust = 1.3, hjust = -0.08, size = 3.5, fontface = "bold",
             fill = "white", label.size = 0.3, show.legend = FALSE) +
  scale_x_log10(
    labels = scales::label_dollar(scale_cut = scales::cut_short_scale()),
    breaks = c(1e4, 5e4, 1e5, 5e5, 1e6, 5e6, 1e7)
  ) +
  scale_fill_manual(values  = ctx_colors, name = NULL) +
  scale_color_manual(values = ctx_colors, name = NULL) +
  labs(
    title    = "Grant Size Distribution: Global CBPF vs Ukraine UHF (2014–2025)",
    subtitle = "Log₁₀ scale — dashed lines mark medians. Ukraine grants are substantially larger.",
    x = "Grant size (USD, log scale)", y = "Density",
    caption = "Source: CBPF ProjectSummary dataset, Feb 2026 extract"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold", size = 14, color = "#005f73"),
    plot.subtitle = element_text(size = 11, color = "#555555"),
    legend.position = "top",
    legend.text   = element_text(size = 11, face = "bold"),
    panel.grid.minor = element_blank()
  )
```

::: {.callout-note}
**Ukraine grants are 2.6× larger at the median** — \$1.13M vs \$440K globally. This reflects both the scale of needs and Ukraine's profile as a large, well-funded conflict response where donors consolidate resources into fewer, larger grants.
:::

## Grant Size by Organisation Type

```{r grant-by-orgtype}
#| fig-width: 12
#| fig-height: 6

grants_org <- grants %>%
  filter(!is.na(OrgType)) %>%
  mutate(OrgType = factor(OrgType, levels = org_order))

ggplot(grants_org,
       aes(x = OrgType, y = Budget / 1e6, fill = Context)) +
  geom_boxplot(
    outlier.shape = 16, outlier.size = 0.8, outlier.alpha = 0.3,
    linewidth = 0.6, width = 0.55,
    position = position_dodge(width = 0.7)
  ) +
  stat_summary(
    fun = median,
    geom = "text",
    aes(label = scales::dollar(..y.. * 1e6, scale = 1e-3, suffix = "K",
                               accuracy = 1, prefix = "$"),
        color = Context),
    position = position_dodge(width = 0.7),
    vjust = -0.6, size = 3.0, fontface = "bold", show.legend = FALSE
  ) +
  scale_fill_manual(values  = ctx_colors, name = NULL) +
  scale_color_manual(values = ctx_colors, name = NULL) +
  scale_y_continuous(
    labels = scales::dollar_format(suffix = "M", accuracy = 0.1),
    trans  = "log10"
  ) +
  labs(
    title    = "Grant Size by Organisation Type: Global vs Ukraine (2014–2025)",
    subtitle = "Log₁₀ scale — boxes show IQR, labels mark medians",
    x = NULL, y = "Grant size (USD millions, log scale)",
    caption = "Source: CBPF ProjectSummary dataset, Feb 2026 extract"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold", size = 14, color = "#005f73"),
    plot.subtitle = element_text(size = 11, color = "#555555"),
    legend.position = "top",
    legend.text   = element_text(size = 11, face = "bold"),
    panel.grid.minor   = element_blank(),
    panel.grid.major.x = element_blank()
  )
```

## INGO vs NNGO Grant Size: A Deeper Look

Globally, INGOs receive grants about **28% larger** than NNGOs at the median. Ukraine's ratio (1.40×) is mid-table but masks a striking post-2022 shift.

### How Does Ukraine Compare Across All Funds?

```{r ingo-nngo-dumbbell}
#| fig-width: 12
#| fig-height: 9

# Per-fund medians for INGO and NNGO (min 10 grants each)
fund_ngotype <- project_summary %>%
  filter(AllocationYear >= 2014,
         !is.na(Budget), Budget > 0,
         OrganizationType %in% c("International NGO", "National NGO")) %>%
  group_by(PooledFundName, OrganizationType) %>%
  summarise(med = median(Budget), n = n(), .groups = "drop")

fund_wide <- fund_ngotype %>%
  pivot_wider(names_from = OrganizationType, values_from = c(med, n)) %>%
  rename(ingo_med = `med_International NGO`,
         nngo_med = `med_National NGO`,
         ingo_n   = `n_International NGO`,
         nngo_n   = `n_National NGO`) %>%
  filter(!is.na(ingo_med), !is.na(nngo_med),
         ingo_n >= 10, nngo_n >= 10) %>%
  mutate(
    ratio     = ingo_med / nngo_med,
    is_ukraine = PooledFundName == "Ukraine",
    PooledFundName = factor(PooledFundName, levels = PooledFundName[order(ratio)])
  )

# Long form for dumbbell
dumbbell_df <- fund_wide %>%
  select(PooledFundName, ingo_med, nngo_med, ratio, is_ukraine) %>%
  pivot_longer(cols = c(ingo_med, nngo_med),
               names_to = "OrgType", values_to = "Median") %>%
  mutate(OrgType = if_else(OrgType == "ingo_med", "INGO", "NNGO"))

ggplot() +
  # segment connecting NNGO to INGO per fund
  geom_segment(
    data = fund_wide,
    aes(x = nngo_med / 1e3, xend = ingo_med / 1e3,
        y = PooledFundName, yend = PooledFundName,
        color = is_ukraine),
    linewidth = 1.2, alpha = 0.7
  ) +
  # dots for each org type
  geom_point(
    data = dumbbell_df,
    aes(x = Median / 1e3, y = PooledFundName,
        shape = OrgType, color = is_ukraine),
    size = 3.5, stroke = 1.0
  ) +
  # ratio label to the right
  geom_text(
    data = fund_wide,
    aes(x = pmax(ingo_med, nngo_med) / 1e3,
        y = PooledFundName,
        label = paste0(round(ratio, 2), "×"),
        fontface = if_else(is_ukraine, "bold", "plain")),
    hjust = -0.25, size = 3.1, color = "grey30"
  ) +
  scale_x_continuous(
    labels = scales::dollar_format(suffix = "K", accuracy = 1),
    expand = expansion(mult = c(0.02, 0.18))
  ) +
  scale_color_manual(
    values = c("TRUE" = "#f4a261", "FALSE" = "#94d2bd"),
    guide  = "none"
  ) +
  scale_shape_manual(values = c("INGO" = 16, "NNGO" = 1), name = NULL) +
  labs(
    title    = "Median Grant Size: INGO vs NNGO per CBPF Fund (2014–2025)",
    subtitle = "Sorted by INGO/NNGO ratio (right labels)  |  <b style='color:#f4a261'>Ukraine highlighted</b>  |  filled = INGO, open = NNGO",
    x = "Median grant size (USD thousands)", y = NULL,
    caption = "Funds with < 10 grants per org type excluded"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title    = element_text(face = "bold", size = 14, color = "#005f73"),
    plot.subtitle = element_markdown(size = 10, color = "#555555"),
    panel.grid.major.y = element_blank(),
    panel.grid.minor   = element_blank(),
    legend.position    = "top",
    legend.text        = element_text(size = 11, face = "bold"),
    axis.text.y        = element_text(
      size  = ifelse(levels(fund_wide$PooledFundName) == "Ukraine", 12, 10),
      face  = ifelse(levels(fund_wide$PooledFundName) == "Ukraine", "bold", "plain"),
      color = ifelse(levels(fund_wide$PooledFundName) == "Ukraine", "#f4a261", "grey30")
    )
  )
```


What the chart presents: Each horizontal segment represents one CBPF fund. The open circle is the NNGO median grant, the filled circle is the INGO median. The further right, the larger the absolute grants. The ratio label is what matters for the comparison.

The headline finding: Ukraine's ratio is unremarkable
At 1.40×, Ukraine sits exactly in the middle of the 26 funds — 13th out of 26. This is counterintuitive given Ukraine's reputation as a progressive localization context. The ratio alone doesn't signal any particular INGO advantage.

But the absolute values tell a different story
Ukraine's NNGO median (~$999K) is likely higher in absolute terms than many funds' INGO medians. A Ukrainian NNGO receives grants that would be considered large even by INGO standards in Sudan or Nigeria. The localization story in Ukraine is not just about equality of ratio — it's that everyone operates at a much higher scale.

The high-ratio funds (Sudan 2.26×, Nigeria/Haiti 2.0×, Iraq 1.88×) are contexts where the NNGO ecosystem is less mature — smaller organizations, less absorptive capacity, so funders intentionally give them smaller grants to manage risk. The INGO premium there reflects structural capacity gaps.

The low-ratio end (Ethiopia 0.87×, Somalia 1.02×, Syria cross-border 1.07×) is where NNGO capacity has caught up or where large national NGOs dominate the landscape, compressing the gap.

The implication for Ukraine: The ratio of 1.40× aggregated over 2014–2025 understates how much has changed recently — that's exactly what the trend chart below it reveals, where the ratio collapsed to 0.84× in 2024. The aggregate figure is dragged up by the early pre-war years when Ukraine operated more like a typical fund.

### Ukraine: INGO/NNGO Grant Ratio Over Time

Since 2022, Ukraine's NNGO grants have grown faster than INGO grants. In 2023 and 2024 the ratio **fell below 1** — NNGOs received larger median grants than INGOs — a direct signal of the localization push intensifying after the full-scale invasion.

```{r ingo-nngo-ratio-trend}
#| fig-width: 12
#| fig-height: 5

# Year-by-year ratio for Ukraine and global average
ngo_ratio_all <- project_summary %>%
  filter(AllocationYear >= 2014,
         !is.na(Budget), Budget > 0,
         OrganizationType %in% c("International NGO", "National NGO")) %>%
  group_by(PooledFundName, AllocationYear, OrganizationType) %>%
  summarise(med = median(Budget), n = n(), .groups = "drop")

make_ratio <- function(df) {
  df %>%
    pivot_wider(names_from = OrganizationType, values_from = c(med, n)) %>%
    rename(ingo = `med_International NGO`, nngo = `med_National NGO`,
           n_i  = `n_International NGO`,  n_n  = `n_National NGO`) %>%
    filter(!is.na(ingo), !is.na(nngo), n_i >= 5, n_n >= 5) %>%
    mutate(ratio = ingo / nngo)
}

ukraine_ratio <- ngo_ratio_all %>%
  filter(PooledFundName == "Ukraine") %>%
  make_ratio() %>%
  mutate(Series = "Ukraine UHF")

global_ratio <- ngo_ratio_all %>%
  filter(PooledFundName != "Ukraine") %>%
  group_by(AllocationYear, OrganizationType) %>%
  summarise(med = median(med), n = sum(n), .groups = "drop") %>%
  mutate(PooledFundName = "Global") %>%
  make_ratio() %>%
  mutate(Series = "Global CBPF average")

ratio_trend <- bind_rows(ukraine_ratio, global_ratio)

series_colors <- c("Ukraine UHF" = "#005f73", "Global CBPF average" = "#94d2bd")

ggplot(ratio_trend,
       aes(x = AllocationYear, y = ratio, color = Series, group = Series)) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "#e63946",
             linewidth = 0.9) +
  annotate("text", x = 2014.2, y = 1.03,
           label = "Parity (INGO = NNGO)", color = "#e63946",
           size = 3.4, hjust = 0, fontface = "italic") +
  geom_ribbon(
    data = ratio_trend %>% filter(Series == "Ukraine UHF"),
    aes(ymin = 1, ymax = ratio, fill = ratio < 1),
    alpha = 0.15, color = NA
  ) +
  geom_line(linewidth = 1.5) +
  geom_point(size = 4) +
  geom_text(aes(label = round(ratio, 2)),
            vjust = -1.1, size = 3.3, fontface = "bold",
            show.legend = FALSE) +
  scale_color_manual(values = series_colors, name = NULL) +
  scale_fill_manual(values = c("FALSE" = "#005f73", "TRUE" = "#e63946"),
                    guide = "none") +
  scale_x_continuous(breaks = 2014:2025) +
  scale_y_continuous(
    labels = function(x) paste0(x, "×"),
    breaks = seq(0.7, 1.8, 0.1)
  ) +
  labs(
    title    = "INGO/NNGO Median Grant Ratio Over Time: Ukraine vs Global (2014–2025)",
    subtitle = "Ratio > 1 means INGOs get larger grants  |  Red shading = years where NNGOs received larger grants (ratio < 1)",
    x = "Allocation Year", y = "INGO median / NNGO median",
    caption = "Global = median of fund-level medians, excluding Ukraine"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold", size = 14, color = "#005f73"),
    plot.subtitle = element_text(size = 11, color = "#555555"),
    legend.position = "top",
    legend.text   = element_text(size = 11, face = "bold"),
    panel.grid.minor = element_blank(),
    axis.text.x      = element_text(angle = 45, hjust = 1, size = 10)
  )
```


A Ukrainian NNGO receives grants that would be considered large even by INGO standards in Sudan or Nigeria.

::: {.callout-note}
**Ukraine's localization in action**: while globally the INGO size premium has been stable at ~1.2–1.4×, Ukraine's ratio dropped to **0.94× in 2023** and **0.84× in 2024**, meaning NNGOs were receiving larger individual grants than INGOs. This likely reflects deliberate fund strategy to direct larger, more complex grants to national partners after the scale-up in 2022.
:::

## Grant Size by Sector — Health in Context

Using each project's primary cluster (the sector receiving the highest budget share), this chart compares median grant sizes across sectors. Health grants are highlighted in red.

```{r grant-by-sector}
#| fig-width: 12
#| fig-height: 7

# Compute median + IQR per cluster for both contexts combined
sector_stats <- grants_cl %>%
  filter(Cluster != "Unknown") %>%
  group_by(Cluster) %>%
  summarise(
    med    = median(Budget),
    q1     = quantile(Budget, 0.25),
    q3     = quantile(Budget, 0.75),
    n      = n(),
    .groups = "drop"
  ) %>%
  arrange(med) %>%
  mutate(
    Cluster    = factor(Cluster, levels = Cluster),
    is_health  = Cluster == "Health"
  )

# Separate Ukraine medians to overlay
ukraine_med <- grants_cl %>%
  filter(Cluster != "Unknown", PooledFundName == "Ukraine") %>%
  group_by(Cluster) %>%
  summarise(ukraine_med = median(Budget), .groups = "drop") %>%
  mutate(Cluster = factor(Cluster, levels = levels(sector_stats$Cluster)))

sector_plot_df <- sector_stats %>%
  left_join(ukraine_med, by = "Cluster")

ggplot(sector_plot_df,
       aes(x = Cluster, y = med / 1e3, fill = is_health)) +
  geom_col(width = 0.7, alpha = 0.88) +
  gghighlight(
    is_health,
    unhighlighted_params = list(fill = "#94d2bd", alpha = 0.6),
    use_direct_label = FALSE
  ) +
  # Ukraine medians as points
  geom_point(
    aes(y = ukraine_med / 1e3),
    shape = 23, size = 3.5, fill = "#005f73", color = "white", stroke = 0.8,
    na.rm = TRUE
  ) +
  geom_text(
    aes(label = paste0("$", scales::comma(round(med / 1e3)), "K")),
    hjust = -0.1, size = 3.2, fontface = "bold"
  ) +
  coord_flip() +
  scale_fill_manual(values = c("FALSE" = "#94d2bd", "TRUE" = "#e63946"),
                    guide = "none") +
  scale_y_continuous(
    labels = scales::dollar_format(suffix = "K", accuracy = 1),
    expand = expansion(mult = c(0, 0.3))
  ) +
  labs(
    title    = "Median Grant Size by Sector — Global CBPF (2014–2025)",
    subtitle = "<b style='color:#e63946'>♥ Health</b> highlighted  |  <b style='color:#005f73'>◆ Diamond</b> = Ukraine UHF median",
    x = NULL, y = "Median grant size (USD thousands)",
    caption = "Source: CBPF ProjectSummary + Cluster datasets, Feb 2026 extract"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold", size = 14, color = "#005f73"),
    plot.subtitle = element_markdown(size = 11, color = "#555555"),
    panel.grid.major.y = element_blank(),
    panel.grid.minor   = element_blank(),
    axis.text.y        = element_markdown(
      size  = ifelse(levels(sector_plot_df$Cluster) == "Health", 12, 11),
      face  = ifelse(levels(sector_plot_df$Cluster) == "Health", "bold", "plain"),
      color = ifelse(levels(sector_plot_df$Cluster) == "Health", "#e63946", "grey30")
    )
  )
```

::: {.callout-note}
**Health grants rank in the middle of the pack globally** (~\$422K median), well below sectors like Logistics, Multi-purpose Cash, and Shelter & NFI. The ◆ diamonds show that Ukraine's health grants are significantly larger than the global norm, consistent with the fund's overall pattern of larger, more consolidated grants.
:::

## Grant Size vs Number of Grants per Organisation {#sec-org-dotchart}

Each dot is one implementing organisation. The x-axis shows how many grants that organisation received over 2022–2025; the y-axis shows their median grant size. Organisations in the top-left corner are high-value, low-frequency recipients; those in the bottom-right are high-frequency implementers of smaller grants.

```{r org-dotchart}
#| fig-width: 13
#| fig-height: 7

org_stats <- grants_cl %>%
  filter(AllocationYear >= 2022, AllocationYear <= 2025,
         !is.na(Budget), Budget > 0,
         !is.na(OrganizationName)) %>%
  group_by(Context, OrganizationName, OrganizationType) %>%
  summarise(
    n_grants    = n(),
    median_size = median(Budget),
    total_usd   = sum(Budget),
    pct_health  = mean(Cluster == "Health"),
    .groups     = "drop"
  ) %>%
  mutate(
    dot_group = case_when(
      pct_health >= 0.5               ~ "Health (≥50% health grants)",
      Context   == "Ukraine UHF"      ~ "Ukraine UHF",
      TRUE                            ~ "Global CBPF"
    )
  )

# Dynamic ceilings: y = max Ukraine median + 4M, x = max Ukraine n_grants + 3
ukr_max   <- org_stats %>% filter(Context == "Ukraine UHF") %>% pull(median_size) %>% max()
ukr_xmax  <- org_stats %>% filter(Context == "Ukraine UHF") %>% pull(n_grants)    %>% max()
y_ceiling <- ukr_max  + 4e6
x_ceiling <- ukr_xmax + 3

dot_colors <- c(
  "Ukraine UHF"                  = "#f4a261",
  "Global CBPF"                  = "#94d2bd",
  "Health (≥50% health grants)"  = "#e63946"
)

# Parametric ellipse for annotation (log-y space)
theta      <- seq(0, 2 * pi, length.out = 200)
ellipse_df <- data.frame(
  x = 9 + 4.2 * cos(theta),
  y = 10^(6.45 + 0.55 * sin(theta))
)

ggplot(org_stats,
       aes(x = n_grants, y = median_size,
           color = dot_group, size = total_usd)) +
  geom_path(data = ellipse_df, aes(x = x, y = y),
            inherit.aes = FALSE,
            color = "black", linewidth = 0.9, linetype = "solid") +
  annotate("text", x = 9, y = 10^6.9,
           label = "The UHF Exception",
           hjust = 0.5, vjust = 1, size = 4.5, fontface = "bold.italic", color = "black") +
  geom_point(alpha = 0.65, stroke = 0) +
  scale_y_log10(
    labels = scales::label_dollar(scale_cut = scales::cut_short_scale()),
    breaks = c(1e4, 5e4, 1e5, 5e5, 1e6, 5e6, 1e7)
  ) +
  scale_x_continuous(labels = scales::label_comma(),
                     breaks = c(1, 2, 3, 5, 7, 10, 15, 20)) +
  scale_color_manual(values = dot_colors, name = NULL,
                     guide  = guide_legend(override.aes = list(size = 4))) +
  scale_size_continuous(
    name   = "Total disbursed (2022–2025)",
    range  = c(2, 16),
    labels = scales::label_dollar(scale_cut = scales::cut_short_scale()),
    breaks = c(1e6, 5e6, 1e7, 3e7, 5e7),
    trans  = "sqrt"
  ) +
  coord_cartesian(xlim = c(NA, x_ceiling), ylim = c(NA, y_ceiling)) +
  labs(
    title    = "Grant Size vs Number of Grants per Organisation (2022–2025)",
    subtitle = "Each dot = one organisation · Y-axis log-scaled · Dot size = total disbursed · Red = majority health grants",
    x = "Number of grants received",
    y = "Median grant size (USD, log scale)",
    caption = "Source: CBPF ProjectSummary dataset, Feb 2026 extract. Axes clipped at max Ukraine values + buffer."
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold", size = 14, color = "#005f73"),
    plot.subtitle = element_text(size = 11, color = "#555555"),
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )
```

```{r org-dotchart-table}
# Who are the high-value Ukraine outliers?
org_stats %>%
  filter(Context == "Ukraine UHF") %>%
  arrange(desc(total_usd)) %>%
  mutate(
    median_size = scales::dollar(median_size, scale_cut = scales::cut_short_scale(), accuracy = 0.01),
    total_usd   = scales::dollar(total_usd,   scale_cut = scales::cut_short_scale(), accuracy = 0.01),
    is_pure_health = pct_health == 1,
    across(
      c(OrganizationName, OrganizationType),
      ~ ifelse(is_pure_health,
               paste0("<span style='color:#e63946;font-weight:bold'>", .x, "</span>"),
               .x)
    ),
    n_grants    = ifelse(is_pure_health,
                         paste0("<span style='color:#e63946;font-weight:bold'>", n_grants, "</span>"),
                         as.character(n_grants)),
    median_size = ifelse(is_pure_health,
                         paste0("<span style='color:#e63946;font-weight:bold'>", median_size, "</span>"),
                         median_size),
    total_usd   = ifelse(is_pure_health,
                         paste0("<span style='color:#e63946;font-weight:bold'>", total_usd, "</span>"),
                         total_usd),
    pct_health  = ifelse(is_pure_health,
                         "<span style='color:#e63946;font-weight:bold'>100%</span>",
                         paste0(round(pct_health * 100), "%"))
  ) %>%
  select(
    Organisation      = OrganizationName,
    Type              = OrganizationType,
    `Grants received` = n_grants,
    `Median grant`    = median_size,
    `Total disbursed` = total_usd,
    `% Health grants` = pct_health
  ) %>%
  knitr::kable(caption = "Ukraine UHF organisations, 2022–2025 — ranked by total disbursed",
               escape  = FALSE)
```

------------------------------------------------------------------------

# Partners per Allocation

Each CBPF allocation is a discrete funding round (e.g. "1st Standard Allocation 2023 — Ukraine"). The number of implementing partners selected per round is a direct signal of concentration: a fund that channels money through 10 partners per allocation is far more concentrated than one that routes it through 40. Ukraine's profile as a high-intensity conflict fund, combined with its relatively small NNGO ecosystem, raises the question of whether it selects fewer partners than comparable funds globally.

## Distribution: Ukraine vs Global {#sec-partners-dist}

```{r partners-per-alloc-data}
#| message: false

# Number of distinct partners per allocation round
partners_per_alloc <- project_summary %>%
  filter(AllocationYear >= 2014,
         !is.na(AllocationSourceID),
         !is.na(OrganizationName)) %>%
  group_by(PooledFundName, AllocationSourceID, AllocationSourceName, AllocationYear) %>%
  summarise(
    n_partners   = n_distinct(OrganizationName),
    total_budget = sum(Budget, na.rm = TRUE),
    n_projects   = n(),
    .groups = "drop"
  ) %>%
  mutate(Context = if_else(PooledFundName == "Ukraine", "Ukraine UHF", "Global CBPF"))

# Summary stats for callout
ppa_summary <- partners_per_alloc %>%
  group_by(Context) %>%
  summarise(
    n_allocs  = n(),
    med       = median(n_partners),
    q1        = quantile(n_partners, 0.25),
    q3        = quantile(n_partners, 0.75),
    mean_n    = round(mean(n_partners), 1),
    .groups   = "drop"
  )
```

```{r partners-per-alloc-plot}
#| fig-width: 11
#| fig-height: 5

ctx_colors_ppa <- c("Ukraine UHF" = "#005f73", "Global CBPF" = "#94d2bd")

med_labels <- partners_per_alloc %>%
  group_by(Context) %>%
  summarise(med = median(n_partners), .groups = "drop")

ggplot(partners_per_alloc,
       aes(x = n_partners, fill = Context, color = Context)) +
  geom_density(alpha = 0.35, linewidth = 0.9, adjust = 0.9) +
  geom_vline(data = med_labels,
             aes(xintercept = med, color = Context),
             linetype = "dashed", linewidth = 1.1) +
  geom_label(data = med_labels,
             aes(x = med, y = Inf,
                 label = paste0("Median: ", round(med), " partners"),
                 color = Context),
             vjust = 1.3, hjust = -0.07, size = 3.5, fontface = "bold",
             fill = "white", label.size = 0.3, show.legend = FALSE) +
  scale_fill_manual(values  = ctx_colors_ppa, name = NULL) +
  scale_color_manual(values = ctx_colors_ppa, name = NULL) +
  scale_x_continuous(breaks = c(5, 10, 20, 30, 40, 50, 60, 80, 100)) +
  labs(
    title    = "Partners per Allocation Round: Global CBPF vs Ukraine UHF (2014–2025)",
    subtitle = "Each observation is one allocation round; dashed lines mark medians",
    x = "Number of distinct implementing partners",
    y = "Density",
    caption = "Source: CBPF ProjectSummary dataset, Feb 2026 extract"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold", size = 14, color = "#005f73"),
    plot.subtitle = element_text(size = 11, color = "#555555"),
    legend.position = "top",
    legend.text   = element_text(size = 11, face = "bold"),
    panel.grid.minor = element_blank()
  )
```

```{r partners-callout-vals, include=FALSE}
ukr_med  <- ppa_summary$med[ppa_summary$Context == "Ukraine UHF"]
glob_med <- ppa_summary$med[ppa_summary$Context == "Global CBPF"]
ratio    <- round(glob_med / ukr_med, 1)
```

::: {.callout-note}
**Ukraine allocations reach `r ukr_med` partners at the median** vs `r glob_med` globally — global rounds involve roughly `r ratio`× more partners. This means Ukraine concentrates its funding in far fewer organisations per round, reflecting both the smaller partner ecosystem and a deliberate strategy of larger, consolidated grants.
:::

## Partners per Allocation: Trend Over Time (Ukraine) {#sec-partners-trend}

Ukraine's partner count per allocation has evolved markedly since the full-scale invasion in 2022.

```{r partners-trend-ukraine}
#| fig-width: 11
#| fig-height: 5

# Yearly median and IQR for Ukraine
ukr_trend <- partners_per_alloc %>%
  filter(Context == "Ukraine UHF") %>%
  group_by(AllocationYear) %>%
  summarise(
    med    = median(n_partners),
    q1     = quantile(n_partners, 0.25),
    q3     = quantile(n_partners, 0.75),
    n_obs  = n(),
    .groups = "drop"
  )

# Global yearly median for comparison
glob_trend <- partners_per_alloc %>%
  filter(Context == "Global CBPF") %>%
  group_by(AllocationYear) %>%
  summarise(
    med   = median(n_partners),
    q1    = quantile(n_partners, 0.25),
    q3    = quantile(n_partners, 0.75),
    .groups = "drop"
  )

ggplot() +
  # Global ribbon + line
  geom_ribbon(data = glob_trend,
              aes(x = AllocationYear, ymin = q1, ymax = q3),
              fill = "#94d2bd", alpha = 0.25) +
  geom_line(data = glob_trend,
            aes(x = AllocationYear, y = med, color = "Global CBPF median"),
            linewidth = 0.9, linetype = "dotted") +
  # Ukraine ribbon + line
  geom_ribbon(data = ukr_trend,
              aes(x = AllocationYear, ymin = q1, ymax = q3),
              fill = "#005f73", alpha = 0.20) +
  geom_line(data = ukr_trend,
            aes(x = AllocationYear, y = med, color = "Ukraine UHF median"),
            linewidth = 1.2) +
  geom_point(data = ukr_trend,
             aes(x = AllocationYear, y = med, color = "Ukraine UHF median"),
             size = 3) +
  geom_text(data = ukr_trend,
            aes(x = AllocationYear, y = med,
                label = round(med)),
            vjust = -0.9, size = 3.2, fontface = "bold", color = "#005f73") +
  annotate("rect", xmin = 2021.5, xmax = Inf, ymin = -Inf, ymax = Inf,
           fill = "#e63946", alpha = 0.04) +
  annotate("text", x = 2022, y = Inf, label = "Full-scale\ninvasion",
           vjust = 1.4, hjust = 0, size = 3, color = "#e63946", fontface = "italic") +
  scale_color_manual(
    values = c("Ukraine UHF median" = "#005f73", "Global CBPF median" = "#2ca58d"),
    name   = NULL
  ) +
  scale_x_continuous(breaks = 2014:2025) +
  labs(
    title    = "Median Partners per Allocation Round by Year",
    subtitle = "Shaded bands = IQR; dotted line = global median; solid = Ukraine",
    x = "Allocation year", y = "Partners per allocation (median)",
    caption = "Source: CBPF ProjectSummary dataset, Feb 2026 extract"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold", size = 14, color = "#005f73"),
    plot.subtitle = element_text(size = 11, color = "#555555"),
    legend.position = "top",
    legend.text   = element_text(size = 11),
    panel.grid.minor = element_blank(),
    axis.text.x   = element_text(angle = 45, hjust = 1)
  )
```

## Where Does Ukraine Rank? Cross-Fund Comparison {#sec-partners-crossfund}

The dot plot below ranks every CBPF fund by its median number of partners per allocation (2014–2025), positioning Ukraine in the global distribution.

```{r partners-crossfund-dotplot}
#| fig-width: 11
#| fig-height: 9

fund_medians <- partners_per_alloc %>%
  filter(AllocationYear >= 2014) %>%
  group_by(PooledFundName) %>%
  summarise(
    med      = median(n_partners),
    n_allocs = n(),
    .groups  = "drop"
  ) %>%
  mutate(is_ukraine = PooledFundName == "Ukraine") %>%
  filter(n_allocs >= 3) %>%   # at least 3 allocation rounds
  arrange(med) %>%
  mutate(PooledFundName = factor(PooledFundName, levels = PooledFundName))

ggplot(fund_medians,
       aes(x = med, y = PooledFundName,
           color = is_ukraine, size = is_ukraine)) +
  geom_vline(xintercept = median(fund_medians$med),
             linetype = "dashed", color = "grey60", linewidth = 0.8) +
  geom_point() +
  geom_text(aes(label = round(med)),
            hjust = -0.5, size = 3,
            fontface = ifelse(fund_medians$PooledFundName == "Ukraine", "bold", "plain"),
            color = ifelse(fund_medians$PooledFundName == "Ukraine", "#005f73", "grey40")) +
  scale_color_manual(values = c(`TRUE` = "#005f73", `FALSE` = "#94d2bd"),
                     guide  = "none") +
  scale_size_manual(values = c(`TRUE` = 5, `FALSE` = 3), guide = "none") +
  annotate("text",
           x = median(fund_medians$med), y = Inf,
           label = paste0("Overall median: ", round(median(fund_medians$med))),
           hjust = -0.05, vjust = 1.4, size = 3.2, color = "grey50", fontface = "italic") +
  labs(
    title    = "Median Partners per Allocation Round by CBPF Fund (2014–2025)",
    subtitle = "Ukraine (teal) vs all other funds (green); dashed = cross-fund median",
    x = "Median number of partners per allocation", y = NULL,
    caption = "Source: CBPF ProjectSummary dataset, Feb 2026 extract. Funds with < 3 allocations excluded."
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title    = element_text(face = "bold", size = 14, color = "#005f73"),
    plot.subtitle = element_text(size = 11, color = "#555555"),
    panel.grid.minor = element_blank(),
    axis.text.y    = element_text(
      size  = ifelse(levels(fund_medians$PooledFundName) == "Ukraine", 12, 10),
      face  = ifelse(levels(fund_medians$PooledFundName) == "Ukraine", "bold", "plain"),
      color = ifelse(levels(fund_medians$PooledFundName) == "Ukraine", "#005f73", "grey30")
    )
  )
```

------------------------------------------------------------------------

# Localization in Context: Ukraine vs All CBPF Funds

The INGO/NNGO grant-size story gains its full meaning when compared across all 26 CBPF pooled funds. How unusual is Ukraine's localization trajectory? Where did it stand in 2022, and where does it stand now?

```{r crossfund-prep}
#| message: false
#| warning: false

# --- Fund-level NNGO budget share, 2022-2025 aggregate ---
loc_agg <- grants %>%
  filter(AllocationYear >= 2022,
         OrganizationType %in% c("International NGO", "National NGO")) %>%
  group_by(PooledFundName) %>%
  summarise(
    total_M       = sum(Budget) / 1e6,
    nngo_share    = sum(Budget[OrganizationType == "National NGO"]) / sum(Budget) * 100,
    ingo_orgs     = n_distinct(OrganizationName[OrganizationType == "International NGO"]),
    nngo_orgs     = n_distinct(OrganizationName[OrganizationType == "National NGO"]),
    .groups = "drop"
  ) %>%
  mutate(
    is_ukraine  = PooledFundName == "Ukraine",
    fund_label  = PooledFundName,
    fund_label  = factor(fund_label, levels = fund_label[order(nngo_share)])
  )

# --- Year-by-year NNGO share for the 8 largest funds ---
top_funds <- loc_agg %>%
  slice_max(total_M, n = 8) %>%
  pull(PooledFundName)

loc_trend <- grants %>%
  filter(AllocationYear >= 2019, AllocationYear <= 2025,
         PooledFundName %in% top_funds,
         OrganizationType %in% c("International NGO", "National NGO")) %>%
  group_by(PooledFundName, AllocationYear) %>%
  summarise(
    nngo_share = sum(Budget[OrganizationType == "National NGO"]) / sum(Budget) * 100,
    .groups = "drop"
  ) %>%
  mutate(is_ukraine = PooledFundName == "Ukraine")
```

## Where Ukraine Ranks Among All Funds {#sec-loc-rank}

```{r crossfund-rank}
#| fig-width: 9
#| fig-height: 7
#| fig-cap: "NNGO share of total budget (2022–2025 combined). Dot size reflects total fund volume. Ukraine sits 12th overall — but the aggregate masks a sharp recent acceleration."

ggplot(loc_agg,
       aes(x = fund_label, y = nngo_share)) +
  geom_segment(
    aes(xend = fund_label, y = 0, yend = nngo_share,
        colour = is_ukraine, linewidth = is_ukraine),
    show.legend = FALSE
  ) +
  geom_point(
    aes(size = total_M, fill = is_ukraine),
    shape = 21, colour = "white", stroke = 0.6
  ) +
  geom_text(
    data = filter(loc_agg, is_ukraine),
    aes(label = paste0(round(nngo_share, 0), "%")),
    hjust = -0.5, fontface = "bold", size = 3.8, colour = "#005f73"
  ) +
  geom_text(
    data = filter(loc_agg, !is_ukraine & (nngo_share > 60 | nngo_share < 6)),
    aes(label = paste0(round(nngo_share, 0), "%")),
    hjust = -0.45, size = 3, colour = "grey40"
  ) +
  scale_colour_manual(values = c("FALSE" = "grey70", "TRUE" = "#005f73")) +
  scale_linewidth_manual(values = c("FALSE" = 0.5, "TRUE" = 1.1)) +
  scale_fill_manual(values = c("FALSE" = "#94d2bd", "TRUE" = "#005f73"),
                    guide = "none") +
  scale_size_continuous(
    name   = "Total fund size (2022–2025, $M)",
    range  = c(3, 11),
    breaks = c(10, 100, 300, 500),
    labels = scales::dollar_format(suffix = "M", accuracy = 1)
  ) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1, scale = 1),
    limits = c(0, 115),
    expand = expansion(mult = c(0, 0.05))
  ) +
  coord_flip() +
  labs(
    title   = "NNGO Share of CBPF Funding, All 26 Funds (2022–2025)",
    subtitle = "Ukraine highlighted in <b style='color:#005f73'>teal</b> — 12th overall, but aggregate masks a sharp recent rise",
    x = NULL, y = "NNGO share of total budget",
    caption = "Source: CBPF ProjectSummary dataset, Feb 2026 extract. INGO/NNGO grants only."
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title       = element_text(face = "bold", size = 14, colour = "#005f73"),
    plot.subtitle    = element_markdown(size = 11, colour = "#555555"),
    panel.grid.major.y = element_blank(),
    panel.grid.minor   = element_blank(),
    legend.position  = "bottom",
    legend.title     = element_text(size = 10)
  )
```

::: {.callout-note}
Ukraine's **37% aggregate NNGO share** places it 12th among all 26 CBPF funds — in the middle of the field, and well below high-localization contexts like Somalia (70%), Venezuela (69%), or Syria Cross-border (54%). However, this aggregate combines years when Ukraine's share was as low as 21% (2023) with 2025, when it reached 58%. The aggregate rank understates where Ukraine currently stands.
:::

## The Trajectory: Ukraine vs the Eight Largest Funds {#sec-loc-trend}

```{r crossfund-trend}
#| fig-width: 10
#| fig-height: 6
#| fig-cap: "NNGO budget share over time for the 8 largest CBPF funds (by total 2022–2025 volume). Ukraine's trajectory from 2023 to 2025 is the steepest acceleration of any large fund."

# End-of-line labels
trend_labels <- loc_trend %>%
  group_by(PooledFundName) %>%
  filter(AllocationYear == max(AllocationYear)) %>%
  ungroup()

ggplot(loc_trend,
       aes(x = AllocationYear, y = nngo_share,
           group = PooledFundName, colour = is_ukraine)) +
  geom_hline(yintercept = 50, linetype = "dashed", colour = "grey60", linewidth = 0.5) +
  annotate("text", x = 2019.1, y = 51.5, label = "50% parity line",
           colour = "grey50", size = 3, hjust = 0) +
  geom_line(
    data = filter(loc_trend, !is_ukraine),
    aes(linewidth = is_ukraine),
    colour = "#94d2bd", alpha = 0.7
  ) +
  geom_line(
    data = filter(loc_trend, is_ukraine),
    linewidth = 1.6, colour = "#005f73"
  ) +
  geom_point(
    data = filter(loc_trend, is_ukraine),
    size = 3.5, colour = "#005f73"
  ) +
  geom_text_repel(
    data = trend_labels,
    aes(label = PooledFundName, colour = is_ukraine),
    hjust = 0, nudge_x = 0.2, size = 3.2,
    segment.size = 0.3, segment.colour = "grey60",
    direction = "y", show.legend = FALSE
  ) +
  scale_colour_manual(values = c("FALSE" = "#94d2bd", "TRUE" = "#005f73"),
                      guide = "none") +
  scale_linewidth_manual(values = c("FALSE" = 0.7, "TRUE" = 1.6),
                         guide = "none") +
  scale_x_continuous(
    breaks = 2019:2025,
    limits = c(2019, 2026.8)
  ) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1, scale = 1),
    limits = c(0, 100)
  ) +
  labs(
    title    = "NNGO Budget Share Over Time — Ukraine vs 7 Largest CBPF Funds",
    subtitle = "<b style='color:#005f73'>Ukraine</b> shown in teal  |  Other large funds in light green",
    x = NULL, y = "NNGO share of total budget",
    caption = "Source: CBPF ProjectSummary dataset, Feb 2026 extract. Top 8 funds by 2022–2025 total volume."
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title       = element_text(face = "bold", size = 14, colour = "#005f73"),
    plot.subtitle    = element_markdown(size = 11, colour = "#555555"),
    panel.grid.minor = element_blank(),
    legend.position  = "none"
  )
```

::: {.callout-note}
Among the eight largest CBPF funds, **Ukraine's acceleration from 2023 to 2025 is the steepest of any fund** — a +38 percentage point swing in two years, moving from 21% to 58%. Sudan moves in the opposite direction, dropping to near zero. Syria Cross-border and Ethiopia show similarly strong localization trajectories but started from a higher baseline. Afghanistan, the largest single fund at \$546M, remains consistently below 40%.
:::

------------------------------------------------------------------------

*Report generated on `r Sys.Date()` using Quarto and the assistance of Gemini 3. Pro., Claude 4,6 and Ollama models. *